---
title: "📦 ACTED Data Helper Functions"
output:
  github_document:
    toc: true
    toc_depth: 3
always_allow_html: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

```

# acheck

The goal of `acheck` is to streamline data verification, cleaning, sampling, and reporting workflows.

## Installation

You can install the development version of `acheck` like so:

```{r installation, eval=FALSE}
# install.packages("devtools")
pak::pkg_install("bokola/acheck")

```

## Basic Examples

```{r example-load, message=FALSE, warning=FALSE}
library(acheck)

# clean string inputs safely
clean_string("  Nachola124! ")

# clean phone numbers by standardizing county codes
df_phones <- data.frame(telephone_number = c("712345678", "0712345678"))
clean_phone_number(df_phones, col = "telephone_number")

# categorize numeric inputs
cut_numeric(c(1, 5, 12, 19, 25), breaks = c(-Inf, 18, Inf), labels = c("<18", ">=18"))

```

## Core Functions Summary

| Function | Description |
| --- | --- |
| `format_num()` | Formats numeric vectors using `gtsummary` styles. |
| `add_gt_to_word()` | Appends `gtsummary` objects directly into Word documents. |
| `clean_username()` | Standardizes string formats to clean title case. |
| `export_list_to_word()` | Exports a list of data frames or tables to a Word file. |
| `create_summary_table()` | Builds customizable `gtsummary` tables. |
| `create_survey_summary_table()` | Generates summary tables for weighted survey design objects. |
| `create_listing_table()` | Creates formatted record listings with `gtreg`. |
| `clean_string()` | Converts strings to lowercase and strips non-alphanumeric characters. |
| `clean_phone_number()` | Normalizes Kenyan phone numbers by prepending `254`. |
| `process_payroll()` | Runs comprehensive cleaning pipelines on payroll data frames. |
| `clean_camp()` | Recodes beneficiary location labels into standardized refugee camps. |
| `get_group_cols()` | Extracts variable names from data dictionaries by positional index. |
| `check_esr_registrations()` | Checks for duplicate household entries across dataset pairs. |
| `get_standard_villages()` | Splits and standardizes multi-village entries. |
| `find_extra_agency_records()` | Identifies unmatched audit records between datasets. |
| `grep_columns_in_list()` | Searches for matching column names across lists of data frames. |
| `pull_kobo_data()` | Fetches survey submissions directly from KoboToolbox API. |
| `pull_kobo_data_mpca()` | Downloads and aligns complex multi-version Kobo dataset structures. |
| `clean_kobo_excel()` | Normalizes manual Excel exports to match API formatting standards. |
| `cut_numeric()` | Bins numeric vectors into labeled factor categories. |
| `export_to_excel_formatted()` | Writes summary objects to Excel with custom formatting rules. |
| `export_to_excel_appended()` | Appends new worksheets safely to existing Excel workbooks. |
| `get_duplicates()` | Returns duplicate record instances based on key variables. |
| `get_distinct()` | Deduplicates records based on phone and name variations. |
| `plot_bar()` | Generates standard summary bar charts using `ggplot2`. |
| `plot_bar_top_n()` | Plots frequency distributions filtered to top $N$ categories. |
| `plot_age_gender()` | Builds age-gender population pyramids using `patchwork`. |
| `patchwork_grid()` | Arranges multiple plots into structured grid layouts. |
| `add_custom_total()` | Calculates custom summary aggregate rows for data frames. |
| `collapse_villages()` | Merges shared target rows while concatenating village labels. |
| `get_partial_matches()` | Detects approximate string inclusion between two variables. |
| `get_partial_matches_extend()` | Maps the best string match between data frames using Jaro-Winkler distances. |
| `get_pii()` | Isolates personally identifiable information from datasets. |
| `write_list_to_excel()` | Writes a list of data frames into a single Excel workbook. |
| `flatten_dm_by_names()` | Extracts and flattens relational data models (`dm`) into data frame lists. |
| `preprocess_reg()` | Runs complete pre-processing logic for registration data. |
| `preprocess()` | Standardizes demographic variables, labels, and categories. |
| `slice_reg()` | Filters registration records based on target date boundaries. |
| `get_edit_data()` | Imports and prepares partner data modification files. |
| `draw_pps_sample()` | Draws a Probability Proportional to Size (PPS) household sample. |
| `draw_pps_hamilton()` | Performs PPS sampling using the Hamilton largest-remainder method. |
| `draw_strs_sample()` | Draws a Stratified Simple Random Sample across strata. |
| `draw_cluster_sample()` | Performs two-stage cluster random sampling across villages. |
| `draw_systematic_across_villages()` | Executes systematic random sampling across ordered clusters. |

```

```

## 👨‍💻 Author Notes

Designed for **humanitarian data workflows**, **Kobo pipelines**, **beneficiary validation**, and **report automation**.

<!-- badges: start -->
  [![R-CMD-check](https://github.com/Bokola/acheck/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/Bokola/acheck/actions/workflows/R-CMD-check.yaml)
  [![codecov](https://codecov.io/gh/Bokola/acheck/branch/main/graph/badge.svg)](https://codecov.io/gh/Bokola/acheck)
  <!-- badges: end -->
