---
title: "Data checks"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- Run this 'development' chunk -->
<!-- Store every call to library() that you need to explore your functions -->

```{r development, include=FALSE}
library(testthat)
library(roxygen2)
```

<!--
 You need to run the 'description' chunk in the '0-dev_history.Rmd' file before continuing your code there.

If it is the first time you use {fusen}, after 'description', you can directly run the last chunk of the present file with inflate() inside.
--> 

```{r development-load}
# Load already included functions if relevant
pkgload::load_all(export_all = FALSE)
```

# Include some data examples in your package

<!-- 
 Store your dataset in a directory named "inst/" at the root of your project.
 Use it for your tests in this Rmd thanks to `pkgload::load_all()` to make it available
and `system.file()` to read it in your examples.

- There already is a dataset in the "inst/" directory to be used in the examples below
-->

```{r development-dataset}
# Run all this chunk in the console directly
# There already is a dataset in the "inst/" directory
# Make the dataset file available to the current Rmd during development
pkgload::load_all(path = here::here(), export_all = FALSE)

# You will be able to read your example data file in each of your function examples and tests as follows - see chunks below
# datafile <- system.file("nyc_squirrels_sample.csv", package = "acheck")
# nyc_squirrels <- read.csv(datafile, encoding = "UTF-8")
# 
# nyc_squirrels
```


## Add gtsummary object to word

<!--
Create a chunk for the core of the function

- The chunk needs to be named `function` at least
- It contains the code of a documented function
- The chunk can also be named `function-my_median` to make it easily
findable in your Rmd
- Let the `@examples` part empty, and use the next `examples` chunk instead to present reproducible examples

After inflating the template

-  This function code will automatically be added in a new file in the "R/" directory
-->

```{r development-function}
# Prepare the code of your function here
```



```{r function-format_num}
#' format numeric vectors using gtsummary style number labels safely
#'
#' @param x numeric vector to be formatted
#' @param ... optional arguments passed along to label_style_number
#'
#' @returns a character vector of formatted numbers
#' @importFrom gtsummary label_style_number
#' @export
format_num <- function(x, ...) {
  # build the formatting constructor inside an isolated execution context to bypass compiler warnings
  formatter <- gtsummary::label_style_number(...)
  
  # execute the function directly on the incoming numeric elements
  formatter(x)
}
```


```{r function-add_gt_to_word}

#' Add a gtsummary objet to word doc
#'
#' @param doc word doc created as `officer::read_docx()`
#' @param gt_obj gtsummary object
#' @param title character header
#' @param landscape logical. page orientation
#' @importFrom flextable  body_add_flextable
#' @importFrom officer body_add_par body_add_break body_end_section_landscape
#'
#' @returns a word doc
#' @export
#'
#' @examples
#' \dontrun{
#' add_gt_to_word(doc, gt_obj, title, landscape = FALSE)
#' }
add_gt_to_word <- function(doc, gt_obj, title, landscape = FALSE) {
  # Convert gtsummary to flextable
  ft <- gtsummary::as_flex_table(gt_obj) %>%
    flextable::autofit()
  
  # Add title and table
  doc <- doc %>%
    body_add_par(title, style = "heading 2") %>%
    body_add_flextable(ft) %>%
    body_add_par("") # spacing
  
  # Handle Orientation and Page Breaks
  if (landscape) {
    # Close the current landscape section
    doc <- doc %>% body_end_section_landscape() 
  } else {
    # If not landscape, just add a standard page break
    doc <- doc %>% body_add_break()
  }
  
  return(doc)
}

```



<!--
Create a chunk with an example of use for your function

- The chunk needs to be named `examples` at least
- It contains working examples of your function
- The chunk is better be named `examples-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This example will automatically be added in the '@examples' part of our function above in the "R/" directory
- This example will automatically be added in the vignette created from this Rmd template
-->


<!--
Create a chunk with a test of use for your function

- The chunk needs to be named `tests` at least
- It contains working tests of your function
- The chunk is better be named `tests-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This test code will automatically be added in the "tests/testthat/" directory
-->

```{r tests-add_gt_to_word}
library(testthat)
library(officer)
library(gtsummary)
library(dplyr)
test_that("add_gt_to_word works correctly under different conditions", {
  
  # setup: create dummy data and a basic gtsummary object
  mock_df <- data.frame(
    status = c("active", "inactive", "active"),
    age = c(25, 34, 45)
  )
  mock_gt <- gtsummary::tbl_summary(mock_df, by = status)
  
  # test 1: verification of the standard portrait flow
  doc_portrait <- officer::read_docx()
  res_portrait <- add_gt_to_word(
    doc = doc_portrait, 
    gt_obj = mock_gt, 
    title = "Test Portrait Title", 
    landscape = FALSE
  )
  
  # check that an officer document object is returned
  expect_s3_class(res_portrait, "rdocx")
  
  # convert doc to content dataframe to check text insertion
  content_p <- officer::docx_summary(res_portrait)
  expect_true(any(content_p$text == "Test Portrait Title"))
  
  # test 2: verification of the landscape flow
  doc_landscape <- read_docx()
  res_landscape <- add_gt_to_word(
    doc = doc_landscape, 
    gt_obj = mock_gt, 
    title = "Test Landscape Title", 
    landscape = TRUE
  )
  
  expect_s3_class(res_landscape, "rdocx")
  
  content_l <- docx_summary(res_landscape)
  expect_true(any(content_l$text == "Test Landscape Title"))
  
  # test 3: safe verification of landscape structural layout differences
  # landscape structural blocks alter row structures or total summary elements
  expect_false(identical(nrow(content_p), nrow(content_l)))
})
```

## clean names

```{r function-clean_username}
#' clean names by removing special characters, deleting extra spaces, and converting to title case
#'
#' @param x character vector of names to clean
#'
#' @returns a vector
#' @export
#'
#' @examples
#' \dontrun{clean_username()}
clean_username <- function(x) {
  x %>%
    # remove all special characters like brackets, braces, and punctuation
    stringr::str_remove_all("[^[:alnum:][:space:]]") %>%
    # collapse duplicate whitespace down to single spaces
    stringr::str_squish() %>%
    # convert the cleaned text string to title case format
    stringr::str_to_title()
}
```



## export list of gt objects/dataframes to word

```{r  function-export_list_to_word}
#' Export a list of dataframes to word
#'
#' @param table_list list of dataframes
#' @param doc_path character vector of output directory
#' @importFrom flextable  body_add_flextable
#' @importFrom officer body_add_par body_add_break 
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_list_to_word()}
export_list_to_word <- function(table_list, doc_path) {
  
  # initialize a new word document
  doc <- officer::read_docx()
  
  # iterate through the named list
  # .x is the table/df, .y is the name of the list element
  purrr::iwalk(table_list, function(.x, .y) {
    
    # logic to handle different table types
    if (inherits(.x, "gtsummary")) {
      # convert gtsummary to flextable
      ft <- gtsummary::as_flex_table(.x)
    } else {
      # treat as standard dataframe and format
      ft <- flextable::flextable(.x) %>%
        flextable::theme_booktabs() %>%
        flextable::autofit() %>%
        flextable::bold(part = "header")
    }
    
    # add the header and the table to the document
    doc <<- doc %>%
      # use the list name as a header 2
      body_add_par(.y, style = "heading 2") %>%
      body_add_flextable(ft) %>%
      # add a small spacer after each table
      body_add_par("") 
  })
  
  # save the completed document
  print(doc, target = doc_path)
}
```


## Create gtsummary table

```{r function-create_summary_table}
#' create a `gtsummary` table
#'
#' @param data dataframe
#' @param cols columns to include
#' @param labels character. labels
#' @param group_by character. grouping variable
#' @param strata character. stratifying variable
#' @param percent_type character. percents by row or column
#' @param dichotomous_as_continuous logical. treat binary 0/1  and likert scale columns as continuous
#' @param report_median logical. whether to include the median alongside the mean for continuous variables (ignored if numeric_summary_type is "sum")
#' @param label_characteristic character. custom header text for the characteristic column
#' @param round_continuous logical. whether to round continuous summaries to whole numbers instead of decimals
#' @param numeric_summary_type character. summary format for numeric columns, either "mean" or "sum"
#'
#' @returns a gtsummary object that renders across all formats natively
#' @export
create_summary_table <- function(data, 
                                 cols = dplyr::everything(), 
                                 labels = list(), 
                                 group_by = NULL,
                                 strata = NULL,
                                 percent_type = "column",
                                 dichotomous_as_continuous = FALSE,
                                 report_median = FALSE,
                                 label_characteristic = "Characteristic",
                                 round_continuous = FALSE,
                                 numeric_summary_type = c("mean", "sum")) {
  
  # validate the requested numeric summary type layout choice
  numeric_summary_type <- match.arg(numeric_summary_type)
  
  # internal helper to build the base table structure
  build_summary <- function(df) {
    
    # establish structural types for the columns
    summary_types <- list(gtsummary::all_categorical() ~ "categorical")
    
    if (dichotomous_as_continuous) {
      # find columns selected by the user that are numeric but classified as categorical or dichotomous
      selected_df <- df %>% dplyr::select({{cols}})
      numeric_likert_cols <- names(selected_df)[sapply(selected_df, is.numeric)]
      
      if (length(numeric_likert_cols) > 0) {
        # explicitly convert these names into a formula assigning them to continuous format
        likert_type_list <- stats::setNames(rep(list("continuous"), length(numeric_likert_cols)), numeric_likert_cols)
        summary_types <- c(summary_types, likert_type_list)
      }
      
      # add standard dichotomous continuous conversion selector rule
      summary_types <- c(summary_types, list(gtsummary::all_dichotomous() ~ "continuous"))
    }
    
    # assign the aggregate display token macro string based on chosen structural type
    continuous_stat <- if (numeric_summary_type == "sum") {
      "{sum}"
    } else if (report_median) {
      "{mean} ({median})"
    } else {
      "{mean}"
    }
    
    # set continuous digit precision based on rounding argument choice
    continuous_digits <- if (round_continuous || numeric_summary_type == "sum") 0 else 1
    
    # build out the core table skeleton structures
    base_summary <- df %>%
      dplyr::select({{cols}}, dplyr::any_of(group_by)) %>%
      gtsummary::tbl_summary(
        by = dplyr::any_of(group_by),
        label = labels,
        percent = percent_type,
        type = summary_types,
        statistic = list(
          gtsummary::all_continuous() ~ continuous_stat,
          gtsummary::all_categorical() ~ "{n} ({p}%)"
        ),
        digits = list(
          # lower case comments without dots or dashes
          # dynamically switch style format if handling raw numbers vs percentage indexes
          gtsummary::all_continuous() ~ function(x) {
            val <- na.omit(x)
            # bypass index scaling logic entirely if calculating aggregate total sums
            if (numeric_summary_type == "sum") {
              return(gtsummary::style_number(x, digits = continuous_digits))
            }
            if (length(val) > 0 && max(val) > 1) {
              return(gtsummary::style_number(x, digits = continuous_digits))
            } else {
              return(paste0(gtsummary::style_number(x * 100, digits = continuous_digits), "%"))
            }
          },
          gtsummary::all_categorical() ~ c(0, 1)
        ),
        missing = "no"
      )
    
    # append an overall summary tracking column explicitly if group_by is actively passed
    # modify label to read total instead of overall when running a sum aggregation table
    if (!is.null(group_by)) {
      overall_label <- if (numeric_summary_type == "sum") "**Total**" else "**Overall**"
      base_summary <- base_summary %>%
        gtsummary::add_overall(last = FALSE, col_label = overall_label)
    }
    
    # append global metadata and formatting styles down the remaining pipeline
    # omit adding the sample size n column if calculating running metric sums
    if (numeric_summary_type != "sum") {
      base_summary <- base_summary %>% gtsummary::add_n()
    }
    
    base_summary %>%
      gtsummary::bold_labels() %>%
      # selectively target non overall group headers using all_stat_cols helper wrapper
      # lower case comments without dots
      gtsummary::modify_header(
        label ~ paste0("**", label_characteristic, "**"),
        gtsummary::all_stat_cols(stat_0 = FALSE) ~ "**{level}**"
      ) %>%
      gtsummary::modify_footnote(gtsummary::all_stat_cols() ~ NA) %>%
      gtsummary::modify_table_body(function(x) {
        attr(x, "table_styling")$columns_width <- NULL
        return(x)
      })
  }
  
  # logic to handle stratification cleanly
  raw_gtsummary <- if (!is.null(strata)) {
    data %>%
      gtsummary::tbl_strata(
        strata = dplyr::all_of(strata),
        .tbl_fun = ~ build_summary(.x)
      )
  } else {
    build_summary(data)
  }
  
  return(raw_gtsummary)
}
```

```{r function-create_survey_summary_table}
#' create a `gtsummary` table from a survey design object
#'
#' @param data a survey design object created via the survey package
#' @param cols columns to include
#' @param labels character. labels
#' @param group_by character. grouping variable
#' @param strata character. stratifying variable
#' @param percent_type character. percents by row or column
#' @param dichotomous_as_continuous logical. treat binary 0/1 and likert scale columns as continuous
#' @param report_median logical. whether to include the median alongside the mean for continuous variables (ignored if numeric_summary_type is "sum")
#' @param label_characteristic character. custom header text for the characteristic column
#' @param round_continuous logical. whether to round continuous summaries to whole numbers instead of decimals
#' @param numeric_summary_type character. summary format for numeric columns, either "mean" or "sum"
#'
#' @returns a gtsummary object that renders across all formats natively
#' @export
create_survey_summary_table <- function(data, 
                                        cols = dplyr::everything(), 
                                        labels = list(), 
                                        group_by = NULL,
                                        strata = NULL,
                                        percent_type = "column",
                                        dichotomous_as_continuous = FALSE,
                                        report_median = FALSE,
                                        label_characteristic = "Characteristic",
                                        round_continuous = FALSE,
                                        numeric_summary_type = c("mean", "sum")) {
  
  # validate the requested numeric summary type layout choice
  numeric_summary_type <- match.arg(numeric_summary_type)
  
  # internal helper to build the base survey table structure
  build_survey_summary <- function(srv_obj) {
    
    # establish structural types for the columns
    summary_types <- list(gtsummary::all_categorical() ~ "categorical")
    
    if (dichotomous_as_continuous) {
      # extract the raw dataframe variables from inside the survey design object matrix boundary
      extracted_df <- srv_obj$variables %>% dplyr::select({{cols}})
      numeric_likert_cols <- names(extracted_df)[sapply(extracted_df, is.numeric)]
      
      if (length(numeric_likert_cols) > 0) {
        # explicitly convert these names into a formula assigning them to continuous format
        likert_type_list <- stats::setNames(rep(list("continuous"), length(numeric_likert_cols)), numeric_likert_cols)
        summary_types <- c(summary_types, likert_type_list)
      }
      
      # add standard dichotomous continuous conversion selector rule
      summary_types <- c(summary_types, list(gtsummary::all_dichotomous() ~ "continuous"))
    }
    
    # assign the aggregate display token macro string based on chosen structural type
    continuous_stat <- if (numeric_summary_type == "sum") {
      "{sum}"
    } else if (report_median) {
      "{mean} ({median})"
    } else {
      "{mean}"
    }
    
    # set continuous digit precision based on rounding argument choice
    continuous_digits <- if (round_continuous || numeric_summary_type == "sum") 0 else 1
    
    # build out the core survey table skeleton structures using tbl_svysummary
    base_summary <- srv_obj %>%
      gtsummary::tbl_svysummary(
        by = dplyr::any_of(group_by),
        include = c({{cols}}, dplyr::any_of(group_by)),
        label = labels,
        percent = percent_type,
        type = summary_types,
        statistic = list(
          gtsummary::all_continuous() ~ continuous_stat,
          gtsummary::all_categorical() ~ "{n_unweighted} ({p}%)"
        ),
        digits = list(
          # lower case comments without dots
          # dynamically switch style format if handling raw numbers vs percentage indexes
          gtsummary::all_continuous() ~ function(x) {
            val <- na.omit(x)
            # bypass index scaling logic entirely if calculating aggregate total sums
            if (numeric_summary_type == "sum") {
              return(gtsummary::style_number(x, digits = continuous_digits))
            }
            if (length(val) > 0 && max(val) > 1) {
              return(gtsummary::style_number(x, digits = continuous_digits))
            } else {
              return(paste0(gtsummary::style_number(x * 100, digits = continuous_digits), "%"))
            }
          },
          gtsummary::all_categorical() ~ c(0, 1)
        ),
        missing = "no"
      )
    
    # append an overall summary tracking column explicitly if group_by is actively passed
    if (!is.null(group_by)) {
      base_summary <- base_summary %>%
        gtsummary::add_overall(last = FALSE, col_label = "**Overall**")
    }
    
    # append global metadata and formatting styles down the remaining pipeline
    base_summary %>%
      gtsummary::bold_labels() %>%
      # selectively target non overall group headers using all_stat_cols helper wrapper
      gtsummary::modify_header(
        label ~ paste0("**", label_characteristic, "**"),
        gtsummary::all_stat_cols(stat_0 = FALSE) ~ "**{level}**"
      ) %>%
      gtsummary::modify_footnote(gtsummary::all_stat_cols() ~ NA) %>%
      gtsummary::modify_table_body(function(x) {
        attr(x, "table_styling")$columns_width <- NULL
        return(x)
      })
  }
  
  # logic to handle stratification cleanly across survey configurations
  raw_gtsummary <- if (!is.null(strata)) {
    data %>%
      gtsummary::tbl_strata(
        strata = dplyr::all_of(strata),
        .tbl_fun = ~ build_survey_summary(.x)
      )
  } else {
    build_survey_summary(data)
  }
  
  return(raw_gtsummary)
}
```


## create a listing table using `gtreg{}`

```{r function-create_listing_table}

#' create `gtreg` table listing of data
#'
#' @param data dataframe
#' @param cols columns to include
#' @param labels character. labels
#' @param group_by character. grouping variable 
#'
#' @returns a list
#' @export
#'
#' @examples
#'\dontrun{create_listing_table()}
create_listing_table <- function(data,
                                 cols = dplyr::everything(),
                                 labels = list(),
                                 group_by = NULL) {
  data %>%
    {
      if (!is.null(group_by))
        dplyr::arrange(., .data[[group_by]])
      else
        .
    } %>%
    dplyr::select({{ cols }}, dplyr::any_of(group_by)) %>%
    gtreg::tbl_listing() %>%
    gtsummary::bold_labels()
}

```

## clean a vector of strings

```{r function-clean_string}
# clean string

#' clean string of spaces and non alphanumeric characters and make lower case
#'
#' @param x a vector of characters
#'
#' @returns a vector
#' @export
#'
#' @examples
#' \dontrun{clean_String('Nachola124')}


clean_string <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_to_lower() %>%
    stringr::str_replace_all("[^a-z0-9]", "") %>%
    stringr::str_trim()
}
```

## clean phone number

```{r function-clean_phone_number}
#' clean phone number by adding 254 prexif
#'
#' @param x dataframe
#' @param col character vector of phone number columns
#'
#' @returns dataframe
#' @export
#'
#' @examples
clean_phone_number <- function(x, col = c("reg_num", "telephone_number", "proxy_num")) {
  x %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::any_of(col),
        ~ {
          # convert elements to character format to guarantee stable evaluation
          char_vec <- base::as.character(.x)
          
          dplyr::case_when(
            # match format missing a leading country code or standard zero index
            base::nchar(char_vec) == 9  ~ base::paste0("254", char_vec),
            # match format structured with a leading single local digit zero index
            base::nchar(char_vec) == 10 ~ base::gsub("^0", "254", char_vec),
            # fall back and preserve raw entries unchanged if lengths do not match
            TRUE ~ char_vec
          )
        }
      )
    )
}
```

## process payroll

```{r function-process_payroll}
#' clean payroll data
#'
#' @param x dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
process_payroll <- function(x) {
  # find columns matching pattern to pass safely to standard selectors
  drop_cols <- base::grep("^x", base::names(x), ignore.case = TRUE, value = TRUE)
  
  # target accounting columns requiring explicit type casting
  numeric_targets <- c(
    "cash_transfer_kes", 
    "mpesa_sending_charges", 
    "mpesa_withdrawal_charges", 
    "total_amount_kes"
  )
  
  x %>%
    # drops matching columns safely if they exist, ignores if empty
    dplyr::select(-dplyr::any_of(drop_cols)) %>%
    
    # defensive filter check: if "no" exists, filters out NAs; if missing, lets all rows pass
    dplyr::filter(dplyr::if_any(dplyr::any_of("no"), ~ !base::is.na(.x)) | !("no" %in% base::names(.))) %>%
    
    # defensive mutate check: targets telephone_number safely using any_of
    dplyr::mutate(
      # dplyr::across(
      #   dplyr::any_of("telephone_number"),
      #   ~ base::gsub("^254", "0", base::as.character(.x))
      # ),
      # defensive numeric conversion check: loops through specified financial column keys if present
      dplyr::across(
        dplyr::any_of(numeric_targets),
        ~ base::as.numeric(.x)
      )
    ) %>%
    mutate(
      across(
        any_of(c("name_of_beneficiary", "name_of_proxy", "telephone_number", "id_number")),
        clean_username
      )
    ) %>% clean_phone_number()
}
```



## recode beneficiary camps to known refugee camps

```{r function-clean_camp}
#' clean recode beneficiary camp 
#'
#' @param x dataframe
#' @param col_name character column with camp names, defaults to 'ben_camp'
#' @param partner_col character column with partner organization names, defaults to 'reforgp'
#'
#' @returns dataframe
#' @importFrom dplyr case_when all_of across
#' @export
#'
#' @examples
#' \dontrun{clean_camp(x, col_name = "ben_camp", partner_col = "reforgp")}
clean_camp <- function(x, col_name = "ben_camp", partner_col = "reforgp") {
  x %>% 
    dplyr::mutate(
      dplyr::across(dplyr::all_of(col_name), ~ dplyr::case_when(
        # use the .data pronoun to dynamically evaluate the partner column name input
        grepl("\\bna\\b|N/A|kawangware|kayole|kasarani|kangemi|kikuyu|home", .x, ignore.case = TRUE) & grepl("HIAS", .data[[partner_col]], ignore.case = TRUE) ~ "Nairobi",
        grepl("^k|^rece", .x, ignore.case = TRUE) ~ "Kakuma",
        grepl("^nai|east|urban|umoja|rongai|settl", .x, ignore.case = TRUE) ~ "Nairobi",
        grepl("^daga|ley$|^ifo|f0|fo|^haga|^v|dad|daad", .x, ignore.case = TRUE) ~ "Daadab",
        is.na(.x) ~ "Nairobi",
        TRUE ~ .x
      ))
    )
}
```


## get group columns

```{r function-get_group_cols}
#' Extract column names from a data dictionary based on structural boundaries
#'
#' @param dict dataframe representing the data dictionary
#' @param name_col string. The column containing the variable names to extract
#' @param vulnerability_df a dataframe with vulnerability categories
#' @param vulnerability_class character vector of vulnerability class
#'
#' @returns a character vector of column names
#' @export
get_group_cols <- function(dict, vulnerability_df, vulnerability_class = "physio", name_col = "name") {
  
  df <- vulnerability_df %>%
    dplyr::filter(
      grepl(vulnerability_class, label, ignore.case = TRUE)
    )
  dict %>%
    dplyr::filter(n >= df$n & n < df$n_end) %>%
    # remove true NAs and any literal "NA" character strings
    dplyr::filter(!is.na(!!rlang::sym(name_col)) & !!rlang::sym(name_col) != "NA") %>%
    dplyr::pull(!!rlang::sym(name_col))
}
```


## find households in ESR data

```{r function-check_esr_registrations}
# registrations appearing in ESR
#' Title
#'
#' @param data1 dataframe. registration data
#' @param data2 dataframe. ESR data
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{check_esr_registrations(data1 = df_clean, data2 = df_esr)}
check_esr_registrations <- function(data1 = df_clean, data2 = df_esr) {

  
  # 1. prepare lookup keys from data2
  # combine names and clean id and phone for matching
  data2_prepared <- data2 %>%
    dplyr::mutate(
      full_name_clean = clean_string(paste(first_name, middle_name, last_name)),
      first_surname = clean_string(paste(first_name, last_name)),
      first_secname = clean_string(paste(first_name, middle_name)),
      id_clean        = clean_string(id_number1),
      phone_clean     = clean_string(phone_number)
    )
  
  lookup_ids    <- unique(data2_prepared$id_clean)
  lookup_names  <- unique(data2_prepared$full_name_clean)
  lookup_names1  <- unique(data2_prepared$first_surname)
  lookup_names2  <- unique(data2_prepared$first_secname)
  lookup_phones <- unique(data2_prepared$phone_clean)
  
  # 2. apply matching logic to data1
  data1 %>%
    dplyr::mutate(
      is_in_second_dataset = case_when(
        # check by id number
        clean_string(ben_id_number) %in% lookup_ids ~ "yes",
        
        # check by phone number using reg_num_phone
        clean_string(reg_num_phone) %in% lookup_phones ~ "yes",
        
        # check by combined name
        clean_string(ben_name) %in% lookup_names ~ "yes",
        clean_string(ben_first_surname) %in% lookup_names1 ~ "yes",
        clean_string(ben_first_secname) %in% lookup_names2 ~ "yes",
        
        # default to no if no matches are found
        TRUE ~ "no"
      )
    )
}
```

## get standard villages

```{r function-get_standard_villages}
#' separate villages collapsed by forward slash ("|")
#'
#' @param data1 dataframe. ESR/partner data
#' @param data2 dataframe. registration data
#'
#' @returns a dataframe
#' @importFrom stringr str_extract str_replace_all
#' @export
#'
#' @examples
#' \dontrun{get_standard_villages(data1 = df_partner, data2 = df_clean)}
get_standard_villages <- function(data1 = df_partner, data2 = df_clean) {
  
  # get unique reference villages and sort by length descending
  # this ensures long names are matched before substrings
  villages <- unique(data2$village2)
  villages_sorted <- villages[order(nchar(villages), decreasing = TRUE)]
  
  # escape special characters and collapse for regex
  village_pattern <- paste(stringr::str_escape(villages_sorted), collapse = "|")
  
  data1 %>%
    # separate rows based on the forward slash
    # this turns "village1/village2" into two distinct rows
    tidyr::separate_rows(village, sep = "/") %>%
    dplyr::mutate(
      # extract the standard name using the pattern
      village2 = str_extract(village, regex(village_pattern, ignore_case = TRUE)),
      # capitalize first letters of village names as per your requirement
      # this handles slashes if they appear in the result
      village2 = str_replace_all(village2, "(^|/)([a-z])", function(m) toupper(m))
    )
}
```

## find extra agency records

```{r function-find_extra_agency_records}
#' find unique extra records in a matched dataset missing from the reference payroll
#'
#' @param matched_df dataframe containing the processed matched records (e.g., df_matched)
#' @param payroll_df dataframe containing the raw reference payroll records (e.g., d_pay_out)
#' @param target_agency character string identifying the specific agency group to audit
#' @param matched_agency_col character name of the agency column in matched_df
#' @param payroll_agency_col character name of the agency column in payroll_df
#'
#' @returns a filtered dataframe containing the rogue records present only in matched_df
#' @export
find_extra_agency_records <- function(matched_df = df_matched, 
                                      payroll_df = d_pay_out, 
                                      target_agency = "ALDEF", 
                                      matched_agency_col = "enum_egency", 
                                      payroll_agency_col = "partner") {
  
  # slice down both data matrices to focus exclusively on the target agency group
  subset_matched <- matched_df %>%
    dplyr::filter(.data[[matched_agency_col]] == target_agency)
  
  subset_payroll <- payroll_df %>%
    dplyr::filter(.data[[payroll_agency_col]] == target_agency)
  
  # capture tracking records that exist in the match pool but have no match on payroll
  extra_records <- subset_matched %>%
    dplyr::anti_join(
      subset_payroll,
      by = c(
        "id_number" = "id_number",
        "telephone_number" = "telephone_number"
      )
    )
  
  return(extra_records)
}
```


## find column names in a list

```{r function-grep_columns_in_list}
#' grep column names matching a pattern across a list of dataframes
#'
#' @param df_list named list of dataframes to scan
#' @param pattern character string or regex pattern to search for
#' @param ignore_case logical. whether the search should be case-insensitive
#'
#' @returns a named list showing only the matched column names for each dataframe
#' @export
grep_columns_in_list <- function(df_list, pattern, ignore_case = TRUE) {
  
  # verify the incoming object is structurally a list format
  if (!base::is.list(df_list)) {
    base::stop("the input df_list must be a valid list object")
  }
  
  # ensure all list elements have tracking names assigned
  if (base::is.null(base::names(df_list))) {
    base::names(df_list) <- base::paste0("df_", base::seq_along(df_list))
  }
  
  # map across the list to extract matching column labels
  matched_results <- purrr::map(df_list, function(df) {
    
    # pull all column elements present in the data frame boundary matrix
    all_cols <- base::names(df)
    
    # run regular expression filter against the column metadata layer
    matching_cols <- stringr::str_subset(
      all_cols, 
      pattern = stringr::regex(pattern, ignore_case = ignore_case)
    )
    
    # return the matches if found, otherwise return null
    if (base::length(matching_cols) > 0) {
      return(matching_cols)
    } else {
      return(NULL)
    }
  })
  
  # compact down the list results to drop elements with zero matches
  purrr::compact(matched_results)
}
```


## pull kobo data


```{r function-pull_kobo_data}
#' pull data from kobo API
#'
#' @param url character. Kobotoolbox url
#' @param project_id character. project id
#' @param token character. project token
#'
#' @returns a list
#' @export
#'
#' @examples
#' \dontrun{pull_kobo_data()}
pull_kobo_data <- function(
    url = Sys.getenv("kobo_url"),
    project_id = Sys.getenv("unocha_id"),
    token = Sys.getenv("unocha_token")
    
){
  
  robotoolbox::kobo_setup(url = url, token = token)
  
  df <- robotoolbox::kobo_data(project_id)
  return(df)
}
```


```{r function-pull_kobo_data_mpca}
#' pull data from any kobo API project and align structures dynamically
#'
#' @param url character. Kobotoolbox url
#' @param project_id character. project id
#' @param token character. project token
#'
#' @returns a clean dataframe matching manual excel column conventions universally
#' @export
pull_kobo_data_mpca <- function(
      url = Sys.getenv("kobo_url2"),
  project_id = Sys.getenv("mpca_id"),
  token = Sys.getenv("kobo_token")
){
  
  robotoolbox::kobo_setup(url = url, token = token)
  
  # pull all data rows and historical columns
  df <- robotoolbox::kobo_data(project_id, all_versions = TRUE)
  
  # force convert metadata levels into plain text strings safely
  df <- df %>%
    dplyr::mutate(dplyr::across(dplyr::everything(), function(col) {
      lbls <- attr(col, "labels")
      if (!is.null(lbls)) {
        lookup <- names(lbls)
        names(lookup) <- as.character(lbls)
        col_str <- as.character(col)
        
        return(dplyr::if_else(col_str %in% names(lookup), lookup[col_str], col_str))
      }
      return(col)
    }))
  
  # 1. drop the plain 'uuid' column first to prevent naming collisions
  df <- df %>%
    dplyr::select(-dplyr::any_of("uuid"))
  
  # 2. rename standard system fields to match manual clean excel structures
  df <- df %>%
    dplyr::rename(
      id = dplyr::any_of("_id"),
      uuid = dplyr::any_of("_uuid"),
      status = dplyr::any_of("_status"),
      submission_time = dplyr::any_of("_submission_time"),
      validation_status = dplyr::any_of("_validation_status"),
      submitted_by = dplyr::any_of("_submitted_by"),
      version = dplyr::any_of("__version__"),
      meta_root_uuid = dplyr::any_of("rootUuid")
    ) %>%
    # project agnostic: converts ALL group delimiter dots to underscores
    # handles patterns like 'a.sec_cluster' -> 'a_sec_cluster' or 'b.age' -> 'b_age'
    dplyr::rename_with(~ stringr::str_replace_all(.x, "\\.", "_"))
  
  # 3. drop general platform-specific headers not in standard clean excel exports
  df <- df %>%
    dplyr::select(
      -dplyr::any_of(c(
        "_attachments",
        "_xform_id_string",
        "instanceID",
        "instanceName",
        "cordinates_latitude",
        "cordinates_longitude",
        "cordinates_altitude",
        "cordinates_precision",
        "index",
        "tags",
        "notes"
      ))
    )
  
  return(df)
}
```

## clean manual kobo excel data

```{r function-clean_kobo_excel}
#' clean manual kobo excel data to align with api structure
#'
#' @param data data.frame. the raw excel data frame (ddd)
#'
#' @returns a cleaned data frame matching the transformed api structure
#' @export
#'
#' @examples
#' \dontrun{clean_kobo_excel(ddd)}
clean_kobo_excel <- function(data) {
  
  if (!is.data.frame(data)) {
    stop("input must be a data frame")
  }
  
  # 1. remove trailing or leading white spaces from character columns
  cleaned_df <- data %>%
    dplyr::mutate(dplyr::across(dplyr::where(is.character), stringr::str_trim))
  
  # 2. handle potential duplicate columns before renaming
  # if both "middle_name" and "ben_reff" exist, drop the plain "ben_reff" to avoid a collision
  if ("middle_name" %in% names(cleaned_df) && "ben_reff" %in% names(cleaned_df)) {
    cleaned_df <- cleaned_df %>% dplyr::select(-dplyr::any_of("ben_reff"))
  }
  
  # 3. handle structural header renames to match code/value naming layout
  cleaned_df <- cleaned_df %>%
    dplyr::rename(
      # map core name metadata variations
      ben_first = dplyr::any_of("first_name"),
      ben_reff = dplyr::any_of("middle_name"),
      benf_last = dplyr::any_of("last_name"),
      data_entry_name = dplyr::any_of("name_of_person_collecting_information"),
      cordinates = dplyr::any_of("please_collect_the_beneficiary_gps_coordinates"),
      
      # convert exploded select_multiple text fields to match api value strings
      livelihood_source_dnk = dplyr::any_of("livelihood_source_dont_know"),
      livelihood_source_f_employment = dplyr::any_of("livelihood_source_formal_employment"),
      livelihood_source_frm_lvstk = dplyr::any_of("livelihood_source_farming_livestock"),
      livelihood_source_other_benflivelihood = dplyr::any_of("livelihood_source_other"),
      livelihood_source_ss_business = dplyr::any_of("livelihood_source_small_scale_business")
    )
  
  # 4. safely drop internal coordinate splits and extra tracker tags 
  cleaned_df <- cleaned_df %>%
    dplyr::select(
      -dplyr::any_of(c(
        "cordinates_latitude",
        "cordinates_longitude",
        "cordinates_altitude",
        "cordinates_precision",
        "please_collect_the_beneficiary_gps_coordinates_latitude",
        "please_collect_the_beneficiary_gps_coordinates_longitude",
        "please_collect_the_beneficiary_gps_coordinates_altitude",
        "please_collect_the_beneficiary_gps_coordinates_precision",
        "index",
        "tags",
        "notes"
      ))
    )
  
  # 5. ensure placeholder columns exist if they are present in df_lab structure
  if ("ben_reff" %in% names(cleaned_df) && !("benf_sec" %in% names(cleaned_df))) {
    cleaned_df <- cleaned_df %>% dplyr::rename(benf_sec = ben_reff)
  }
  
  if (!"ref_priority" %in% names(cleaned_df)) {
    cleaned_df$ref_priority <- NA_character_
  }
  
  return(cleaned_df)
}
```


## categorize numerics

```{r function-cut_numeric}
#' categorize numeric cols
#'
#' @param x a vector of numeric column
#' @param breaks numeric vector of breaks
#' @param labels character vector of labels
#'
#' @returns vector
#' @export
#'
#' @examples
#' cut_numeric(c(1:10), c(-Inf,5, Inf), c("<=5", ">=6"))
cut_numeric <- function(
    x, 
    breaks = c(-Inf, 18, Inf), 
    labels = c("<18", ">=18")
){
  # find the true minimum value of the input vector
  min_val <- min(x, na.rm = TRUE)
  
  # if the minimum value is less than the first break point adjust the breaks and labels
  if (min_val < breaks[1]) {
    breaks <- c(min_val, breaks)
    labels <- c(paste(min_val, "-", breaks[2] - 1), labels)
  }
  
  out <- cut(
    x,
    # breaks now safely includes the true minimum value
    breaks = breaks,
    # updated labels matching the intervals
    labels = labels,
    # include the lowest value and make intervals right open
    include.lowest = TRUE,
    right = FALSE
  )
  return(out)
}
```

## export gtsummary object to excel workbook

```{r function-export_to_excel_formatted}
#' Export/write gtsummary output to excel workbook
#'
#' @param list_of_objects list of `gtsummary` object
#' @param file_name character. output filename and associated path
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_to_excel_formatted()} 
export_to_excel_formatted <- function(
    list_of_objects,
    file_name = paste0("UNOCHA_Summary_Report_", Sys.Date(), ".xlsx")
) {
  
  # create the workbook object
  wb <- openxlsx::createWorkbook()
  
  # define styles
  neg_style <- createStyle(fontColour = "#FF0000", textDecoration = "bold")
  # style for bolding the row headers (the first column)
  bold_header_style <- createStyle(textDecoration = "bold")
  
  # iterate through the named list
  purrr::iwalk(list_of_objects, function(obj, sheet_name) {
    
    openxlsx::addWorksheet(wb, sheet_name)
    
    # conversion logic
    df_to_write <- if (inherits(obj, "gtsummary")) {
      as_tibble(obj)
    } else {
      as.data.frame(obj)
    }
    
    openxlsx::writeData(wb, sheet_name, df_to_write)
    
    # 1. bold the row headers (first column)
    # rows starts at 2 to skip the excel header row
    openxlsx::addStyle(
      wb, 
      sheet_name, 
      style = bold_header_style, 
      cols = 1, 
      rows = 2:(nrow(df_to_write) + 1), 
      gridExpand = TRUE
    )
    
    # 2. apply red color coding for negative values
    openxlsx::conditionalFormatting(
      wb, 
      sheet_name, 
      cols = 1:ncol(df_to_write), 
      rows = 2:(nrow(df_to_write) + 1), 
      rule = "< 0", 
      style = neg_style
    )
    
    openxlsx::setColWidths(wb, sheet_name, cols = 1:ncol(df_to_write), widths = "auto")
  })
  
  openxlsx::saveWorkbook(wb, file = file.path("output", file_name), overwrite = TRUE)
}
```

## add a new sheet to existing workbook

```{r function-export_to_excel_appended}
#' Export a sheet to existing workbook
#'
#' @param list_of_objects list of `gtsummary` object
#' @param sheetname character. worksheet name
#' @param file_name character. output file name
#' @param output_dir character. output directory
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_to_excel_appended()}
export_to_excel_appended <- function(
    list_of_objects = df_target_p,
    sheetname = "Target and Registered ESR",
    file_name = paste0("UNOCHA_Summary_Report_", Sys.Date(), ".xlsx"),
    output_dir = "output"
) {
  
  # create output dir
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # build full path safely
  file_path <- file.path(output_dir, basename(file_name))
  
  message("Using file path: ", normalizePath(file_path, mustWork = FALSE))
  
  # load existing workbook OR create new
  if (file.exists(file_path)) {
    
    message("Loading existing workbook...")
    
    wb <- tryCatch(
      openxlsx::loadWorkbook(file_path),
      error = function(e) {
        message("Failed to load workbook: ", e$message)
        openxlsx::createWorkbook()
      }
    )
    
  } else {
    
    message("Workbook does not exist. Creating new workbook...")
    wb <- openxlsx::createWorkbook()
    
  }
  
  # excel sheet name limit
  sheetname <- substr(sheetname, 1, 31)
  
  # remove existing sheet if present
  if (sheetname %in% names(wb)) {
    openxlsx::removeWorksheet(wb, sheetname)
  }
  
  addWorksheet(wb, sheetname)
  
  # styles
  header_style <- createStyle(
    fontSize = 12,
    textDecoration = "bold",
    fontColour = "#0066A1"
  )
  
  bold_row_style <- openxlsx::createStyle(textDecoration = "bold")
  
  neg_style <- openxlsx::createStyle(
    fontColour = "#FF0000",
    textDecoration = "bold"
  )
  
  current_row <- 1
  
  purrr::iwalk(list_of_objects, function(obj, list_name) {
    
    df_to_write <- if (inherits(obj, "gtsummary")) {
      as_tibble(obj)
    } else {
      as.data.frame(obj)
    }
    
    # section title
    openxlsx::writeData(
      wb,
      sheet = sheetname,
      x = list_name,
      startRow = current_row
    )
    
    openxlsx::addStyle(
      wb,
      sheet = sheetname,
      style = header_style,
      cols = 1,
      rows = current_row
    )
    
    current_row <<- current_row + 1
    
    # write table
    openxlsx::writeData(
      wb,
      sheet = sheetname,
      x = df_to_write,
      startRow = current_row,
      headerStyle = bold_row_style
    )
    
    # bold first column
    if (nrow(df_to_write) > 0) {
      
      openxlsx::addStyle(
        wb,
        sheet = sheetname,
        style = bold_row_style,
        cols = 1,
        rows = (current_row + 1):(current_row + nrow(df_to_write)),
        gridExpand = TRUE,
        stack = TRUE
      )
      
      openxlsx::conditionalFormatting(
        wb,
        sheet = sheetname,
        cols = 1:ncol(df_to_write),
        rows = (current_row + 1):(current_row + nrow(df_to_write)),
        rule = "<0",
        style = neg_style
      )
    }
    
    current_row <<- current_row + nrow(df_to_write) + 3
  })
  
  # autofit
  openxlsx::setColWidths(
    wb,
    sheet = sheetname,
    cols = 1:50,
    widths = "auto"
  )
  
  # save
  openxlsx::saveWorkbook(
    wb,
    file = file_path,
    overwrite = TRUE
  )
  
  message("Workbook saved successfully.")
}
```

## get duplicates

```{r function-get_duplicates}
#' get duplicate beneficiaries on select variables
#'
#' @param x dataframe
#' @param vars_to_check character vector of variables to check
#' @param phone_var character vector. phone number variable for hard check
#' @param name_var character vector. name of respondent
#' @importFrom dplyr group_by across n n_distinct filter any_of ungroup distinct
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_duplicates()}
get_duplicates <- function(
    x,
    vars_to_check = c("benef_name", "ben_number"),
    phone_var = "ben_number",
    name_var = "benef_name"
){
  
  # 1. find phone sharing: same phone, different names
  phone_sharing <- x %>%
    group_by(across(any_of(phone_var))) %>%
    filter(n() > 1 & n_distinct(across(any_of(name_var))) > 1)
  
  # 2. find name sharing: same name, different phones
  name_sharing <- x %>%
    group_by(across(any_of(name_var))) %>%
    filter(n() > 1 & n_distinct(across(any_of(phone_var))) > 1)
  
  # 3. find exact duplicates across the vars_to_check
  exact_dups <- x %>%
    group_by(across(any_of(vars_to_check))) %>%
    filter(n() > 1)
  
  # combine all instances and remove overlapping rows
  res <- bind_rows(phone_sharing, name_sharing, exact_dups) %>%
    distinct() %>%
    ungroup()
  
  # sort by the primary variables for easy review
  res %>%
    arrange(across(any_of(vars_to_check)))
}
```

## get distinct

```{r function-get_distinct}
#' get distinct beneficiaries on select variables
#'
#' @param x dataframe
#' @param vars_to_check character vector of columns to check
#' @param phone_var character vector of phone number column
#' @param name_var character vector of respondent name
#' @importFrom dplyr group_by across n n_distinct filter any_of ungroup distinct select pull row_number bind_rows slice
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_distinct()}
get_distinct <- function(
    x,
    vars_to_check = c("benef_name", "ben_number"),
    phone_var = "ben_number",
    name_var = "benef_name"
){
  # capture exact duplicates and instances of shared name with different phones
  anomalies <- x %>%
    mutate(temp_id = row_number()) %>%
    {
      bind_rows(
        # scenario 2 shared name different phones
        group_by(., across(any_of(name_var))) %>% 
          filter(n() > 1 & n_distinct(across(any_of(phone_var))) > 1),
        
        # scenario 3 exact duplicates
        group_by(., across(any_of(vars_to_check))) %>% 
          filter(n() > 1)
      )
    } %>%
    distinct(temp_id, .keep_all = TRUE)
  
  # isolate clean rows that dont participate in those loops
  clean_data <- x %>%
    mutate(temp_id = row_number()) %>%
    filter(!(temp_id %in% anomalies$temp_id))
  
  # resolve the anomalies by keeping just the first instance per unique variable combo
  resolved_anomalies <- anomalies %>%
    group_by(across(any_of(vars_to_check))) %>%
    slice(1) %>%
    ungroup()
  
  # bind clean entries and resolved entries back together
  bind_rows(clean_data, resolved_anomalies) %>%
    arrange(temp_id) %>%
    select(-temp_id)
}
```

## plot bar

```{r function-plot_bar}
#' plot bar graphs with automatic summary calculations
#'
#' @param data dataframe
#' @param x_var character vector x-axis variable
#' @param y_var character vector variable to calculate mean or proportion on
#' @param type character type of calculation: "count", "proportion", or "mean"
#' @param fill_col character vector of hex color code for fill
#' @param title character vector of title
#' @param subtitle character vector of subtitle
#' @param x_lab character vector of label for x-axis
#' @param y_lab character vector of label for y-axis
#' @importFrom rlang ensym as_label
#' @importFrom dplyr filter group_by summarize mutate ungroup
#' @importFrom ggplot2 ggplot aes geom_col geom_text labs scale_y_continuous expansion theme_minimal theme element_blank element_text
#' @importFrom scales percent comma
#'
#' @returns a ggplot object
#' @export
#'
#' @examples
#' \dontrun{plot_bar(data, x_var = "job_type", type = "count")}
plot_bar <- function(data, 
                     x_var, 
                     y_var = NULL, 
                     type = c("count", "proportion", "mean"),
                     fill_col = "#0066A1",
                     title = "variable distribution", 
                     subtitle = NULL, 
                     x_lab = NULL, 
                     y_lab = NULL) {
  
  type <- match.arg(type)
  x_sym <- rlang::ensym(x_var)
  x_char <- rlang::as_label(x_sym)
  
  # drop missing values from the grouping variable
  data <- data %>% dplyr::filter(!is.na(!!x_sym))
  
  # standardizing labels based on the calculation metric
  if (is.null(y_lab)) {
    y_lab <- switch(type,
                    "count" = "count",
                    "proportion" = "proportion",
                    "mean" = "mean")
  }
  
  # internal data processing engine
  if (!is.null(y_var)) {
    y_sym <- rlang::ensym(y_var)
    
    if (type == "mean") {
      summary_df <- data %>%
        dplyr::group_by(!!x_sym) %>%
        dplyr::summarize(metric = round(mean(!!y_sym, na.rm = TRUE)), .groups = "drop")
      
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = !!x_sym, y = metric)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = scales::comma(metric, accuracy = 0.01)),
          vjust = -0.5, size = 3, fontface = "bold"
        )
      
    } else if (type == "proportion") {
      summary_df <- data %>%
        dplyr::group_by(!!x_sym) %>%
        dplyr::summarize(
          metric = if (is.numeric(!!y_sym) || is.logical(!!y_sym)) {
            mean(!!y_sym == 1, na.rm = TRUE)
          } else {
            mean(!is.na(!!y_sym), na.rm = TRUE)
          }, 
          .groups = "drop"
        )
      
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = !!x_sym, y = metric)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = scales::percent(metric, accuracy = 0.1)),
          vjust = -0.5, size = 3, fontface = "bold"
        )
      
    } else if (type == "count") {
      summary_df <- data %>%
        dplyr::group_by(!!x_sym) %>%
        dplyr::summarize(metric = sum(!!y_sym, na.rm = TRUE), .groups = "drop") %>%
        dplyr::mutate(pct = metric / sum(metric, na.rm = TRUE))
      
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = !!x_sym, y = metric)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = paste0(scales::comma(metric), " (", scales::percent(pct, accuracy = 0.1), ")")),
          vjust = -0.5, size = 3, fontface = "bold"
        )
    }
    
  } else {
    # case when no y variable is supplied (calculating straight counts/proportions of rows)
    summary_df <- data %>%
      dplyr::count(!!x_sym) %>%
      dplyr::mutate(pct = n / sum(n))
    
    if (type == "proportion") {
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = !!x_sym, y = pct)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = scales::percent(pct, accuracy = 0.1)),
          vjust = -0.5, size = 3, fontface = "bold"
        )
    } else {
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = !!x_sym, y = n)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = paste0(scales::comma(n), " (", scales::percent(pct, accuracy = 0.1), ")")),
          vjust = -0.5, size = 3, fontface = "bold"
        )
    }
  }
  
  # chart polishing and design options
  p <- p +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = if(is.null(x_lab)) x_char else x_lab,
      y = y_lab
    ) +
    ggplot2::scale_y_continuous(labels = scales::comma, expand = ggplot2::expansion(mult = c(0, 0.2))) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold", size = 14, color = "#231F20"),
      plot.subtitle = ggplot2::element_text(size = 11, color = "#58595B"),
      axis.title = ggplot2::element_text(face = "bold", color = "#231F20"),
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, color = "#231F20"),
      axis.text.y = ggplot2::element_text(color = "#231F20")
    )
  
  return(p)
}
```

## plot bar top n

```{r function-plot_bar_top_n}
#' plot bar graphs for top n frequencies
#'
#' @param data dataframe
#' @param x_var character vector x-axis variable
#' @param y_var character vector variable to calculate mean or proportion on
#' @param type character type of calculation: "count", "proportion", or "mean"
#' @param n_top integer maximum number of highest value categories to display
#' @param fill_col character vector of hex color code for fill
#' @param title character vector of title
#' @param subtitle character vector of subtitle
#' @param x_lab character vector of label for x-axis
#' @param y_lab character vector of label for y-axis
#' @importFrom rlang ensym as_label
#' @importFrom tidyr separate_rows
#' @importFrom dplyr filter group_by summarize mutate slice_max ungroup desc
#' @importFrom ggplot2 ggplot aes geom_col geom_text labs scale_y_continuous scale_x_discrete expansion theme_minimal theme element_blank element_text
#' @importFrom scales percent comma
#' @importFrom forcats fct_reorder
#'
#' @returns a ggplot object
#' @export
#'
#' @examples
#' \dontrun{plot_bar_top_n(data, x_var = "job_type", type = "count", n_top = 10)}
plot_bar_top_n <- function(data, 
                           x_var, 
                           y_var = NULL, 
                           type = c("count", "proportion", "mean"),
                           n_top = 10,
                           fill_col = "#0066A1",
                           title = "top variable distribution", 
                           subtitle = NULL, 
                           x_lab = NULL, 
                           y_lab = NULL) {
  
  type <- match.arg(type)
  x_sym <- rlang::ensym(x_var)
  x_char <- rlang::as_label(x_sym)
  
  # auto expand multiple response space separated text strings if present
  # if (is.character(data[[x_char]])) {
  #   data <- data %>% tidyr::separate_rows(!!x_sym, sep = " ")
  # }
  # 
  # drop missing values from the grouping variable
  data <- data %>% dplyr::filter(!is.na(!!x_sym))
  
  # standardizing labels based on the calculation metric
  if (is.null(y_lab)) {
    y_lab <- switch(type,
                    "count" = "count",
                    "proportion" = "proportion",
                    "mean" = "mean")
  }
  
  # internal data processing engine
  if (!is.null(y_var)) {
    y_sym <- rlang::ensym(y_var)
    
    if (type == "mean") {
      summary_df <- data %>%
        dplyr::group_by(!!x_sym) %>%
        dplyr::summarize(metric = round(mean(!!y_sym, na.rm = TRUE)), .groups = "drop") %>%
        dplyr::slice_max(order_by = metric, n = n_top, with_ties = FALSE)
      
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = forcats::fct_reorder(!!x_sym, dplyr::desc(metric)), y = metric)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = scales::comma(metric, accuracy = 0.01)),
          vjust = -0.5, size = 3, fontface = "bold"
        )
      
    } else if (type == "proportion") {
      summary_df <- data %>%
        dplyr::group_by(!!x_sym) %>%
        dplyr::summarize(metric = mean(!!y_sym == 1, na.rm = TRUE), .groups = "drop") %>%
        dplyr::slice_max(order_by = metric, n = n_top, with_ties = FALSE)
      
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = forcats::fct_reorder(!!x_sym, dplyr::desc(metric)), y = metric)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = scales::percent(metric, accuracy = 0.1)),
          vjust = -0.5, size = 3, fontface = "bold"
        )
      
    } else if (type == "count") {
      summary_df <- data %>%
        dplyr::group_by(!!x_sym) %>%
        dplyr::summarize(metric = sum(!!y_sym, na.rm = TRUE), .groups = "drop") %>%
        dplyr::mutate(pct = metric / sum(metric, na.rm = TRUE)) %>%
        dplyr::slice_max(order_by = metric, n = n_top, with_ties = FALSE)
      
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = forcats::fct_reorder(!!x_sym, dplyr::desc(metric)), y = metric)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = paste0(scales::comma(metric), " (", scales::percent(pct, accuracy = 0.1), ")")),
          vjust = -0.5, size = 3, fontface = "bold"
        )
    }
    
  } else {
    # case when no y variable is supplied (calculating straight counts/proportions of rows)
    summary_df <- data %>%
      dplyr::count(!!x_sym) %>%
      dplyr::mutate(pct = n / sum(n)) %>%
      dplyr::slice_max(order_by = n, n = n_top, with_ties = FALSE)
    
    if (type == "proportion") {
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = forcats::fct_reorder(!!x_sym, dplyr::desc(pct)), y = pct)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = scales::percent(pct, accuracy = 0.1)),
          vjust = -0.5, size = 3, fontface = "bold"
        )
    } else {
      p <- ggplot2::ggplot(summary_df, ggplot2::aes(x = forcats::fct_reorder(!!x_sym, dplyr::desc(n)), y = n)) +
        ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
        ggplot2::geom_text(
          ggplot2::aes(label = paste0(scales::comma(n), " (", scales::percent(pct, accuracy = 0.1), ")")),
          vjust = -0.5, size = 3, fontface = "bold"
        )
    }
  }
  
  # chart polishing and design options
  p <- p +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = if(is.null(x_lab)) x_char else x_lab,
      y = y_lab
    ) +
    ggplot2::scale_y_continuous(labels = scales::comma, expand = ggplot2::expansion(mult = c(0, 0.2))) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.major.x = ggplot2::element_blank(),
      panel.grid.minor = ggplot2::element_blank(),
      plot.title = ggplot2::element_text(face = "bold", size = 14, color = "#231F20"),
      plot.subtitle = ggplot2::element_text(size = 11, color = "#58595B"),
      axis.title = ggplot2::element_text(face = "bold", color = "#231F20"),
      axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, color = "#231F20"),
      axis.text.y = ggplot2::element_text(color = "#231F20")
    )
  
  return(p)
}
```


## patchwork grid plots

```{r function-patchwork_grid}
#' patchwork a list of plots with headers for sections arranged left to right with unified axes labels
#'
#' @param plot_list list of ggplot objects
#' @param section_headers character vector of headers to place above sections
#' @param plots_per_section integer number of plots that belong under each header
#' @param nrow integer number of rows for the internal plot grids, defaults to 1
#'
#' @returns a patchwork layout object
#' @importFrom patchwork wrap_elements wrap_plots plot_annotation
#' @importFrom ggplot2 labs theme element_text margin element_blank
#' @export
patchwork_grid <- function(plot_list, section_headers, plots_per_section = 1, nrow = 1) {
  
  # validate that we have enough headers for the plots provided
  num_sections <- ceiling(length(plot_list) / plots_per_section)
  if (length(section_headers) < num_sections) {
    stop("not enough section_headers provided for the number of plot groups")
  }
  
  # extract global axis labels from the first plot before stripping them
  global_x_lab <- plot_list[[1]]$labels$x
  global_y_lab <- plot_list[[1]]$labels$y
  
  # strip inner labels to prevent repeating clutter across the panels
  clean_plots <- lapply(plot_list, function(p) {
    p + ggplot2::labs(x = NULL, y = NULL)
  })
  
  wrapped_sections <- list()
  
  # process plots group by group to wrap them under a styled section header
  for (i in seq_len(num_sections)) {
    
    # isolate the subset of plots for the current section
    start_idx <- ((i - 1) * plots_per_section) + 1
    end_idx <- min(i * plots_per_section, length(clean_plots))
    sub_plots <- clean_plots[start_idx:end_idx]
    
    # assemble the sub-grid for this specific section arranged by row
    section_grid <- patchwork::wrap_plots(sub_plots, nrow = nrow)
    
    # wrap the sub-grid as an element and attach a clean overarching title
    wrapped_sections[[i]] <- patchwork::wrap_elements(panel = section_grid) +
      ggplot2::labs(title = section_headers[i]) +
      ggplot2::theme(
        plot.title = ggplot2::element_text(
          face = "bold", 
          size = 14, 
          color = "#231F20", 
          margin = ggplot2::margin(b = 10, t = 5)
        )
      )
  }
  
  # stitch all header-managed sections horizontally side-by-side
  final_layout <- patchwork::wrap_plots(wrapped_sections, nrow = 1) +
    patchwork::plot_annotation(
      caption = global_x_lab, # uses caption style at bottom for clean centered x-axis label alignment
      theme = ggplot2::theme(
        plot.caption = ggplot2::element_text(face = "bold", size = 12, color = "#231F20", hjust = 0.5, margin = ggplot2::margin(t = 10))
      )
    )
  
  return(final_layout)
}
```


## add custom total

```{r function-add_custom_total}
#' calculate column-wise totals
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{add_custom_total()}
add_custom_total <- function(df) {
  # calculate totals
  totals <- df %>%
    dplyr::summarise(
      Village = "Total",
      # only sum unique targeted values per village to avoid double counting
      `Targeted BNF` = base::sum(unique(`Targeted BNF`), na.rm = TRUE),
      `Registered BNF` = sum(`Registered BNF`, na.rm = TRUE),
      Variance = sum(Variance, na.rm = TRUE),
      `ESR Households` = sum(`ESR Households`, na.rm = TRUE),
      partner = NA
    )
  
  # combine with original data
  out <- bind_rows(df, totals)
  
  return(out)
}
```

## collapse villages

```{r function-collapse_villages}
#' collapse villages that share targets, separating by /
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{collapse_villages()}
collapse_villages <- function(df) {
  df_collapsed <- df %>%
    # group by the columns where you want to identify duplicates
    dplyr::group_by(`Targeted BNF`, `ESR Households`, partner) %>%
    dplyr::summarise(
      # collapse village names with a slash
      village = paste(unique(village), collapse = " / "),
      # sum the registered beneficiaries
      `Registered BNF` = sum(`Registered BNF`, na.rm = TRUE),
      # keep the group order clean
      .groups = "drop"
    ) %>%
    # relocate columns to match your original structure
    dplyr::select(village, `Targeted BNF`, `ESR Households`, partner, `Registered BNF`)
  
  return(df_collapsed)
}
```

## find partial matches

```{r function-get_partial_matches}
#' find partial matches between focal and beneficiary
#'
#' @param df dataframe
#' @param focal character vector of focal variable
#' @param beneficiary character vector of beneficiary variable
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{get_partial_matches()}
get_partial_matches <- function(df, focal = "focal_name", beneficiary = "benef_name") {
  df %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      is_partial = {
        f_raw <- .data[[focal]]
        b_raw <- .data[[beneficiary]]
        
        # 1 early exit check for missing values or empty strings
        if (is.na(f_raw) || is.na(b_raw) || f_raw == "" || b_raw == "" || f_raw == "NA" || b_raw == "NA") {
          FALSE
        } else {
          f_str <- as.character(f_raw)
          b_str <- as.character(b_raw)
          
          # 2 basic substring check first using base r grepl to avoid vector length bugs
          matched <- grepl(b_str, f_str, fixed = TRUE) | grepl(f_str, b_str, fixed = TRUE)
          
          # 3 if basic check fails look for out of order word inclusion
          if (isTRUE(matched)) {
            matched
          } else {
            # clean and split both strings into words
            f_words <- unlist(stringr::str_split(f_str, "\\s+"))
            b_words <- unlist(stringr::str_split(b_str, "\\s+"))
            
            # identify which name is shorter by token count
            if (length(f_words) <= length(b_words)) {
              short_words <- f_words
              long_str <- b_str
            } else {
              short_words <- b_words
              long_str <- f_str
            }
            
            # verify if every word of the shorter name exists in the longer string
            all(purrr::map_lgl(short_words, ~ grepl(.x, long_str, fixed = TRUE)))
          }
        }
      }
    ) %>% 
    dplyr::ungroup()
}
```

```{r function-get_partial_matches_extend}
#' find partial matches and map the single best-matching focal name onto dataframe 2
#'
#' @param df1 dataframe containing the focal variable (source of matched names)
#' @param df2 dataframe containing the beneficiary variable (target destination)
#' @param focal character name of the focal variable column in df1
#' @param beneficiary character name of the beneficiary variable column in df2
#' @param min_shared_words integer minimum number of unique words that must match to trigger a connection
#'
#' @returns dataframe 2 with an appended 'name_possible' column containing the highest probability match from df1
#' @export
get_partial_matches_extend <- function(df1, df2, focal = "focal_name", beneficiary = "benef_name", min_shared_words = 2) {
  
  # create unique row tracking keys to link matching vectors back safely
  df1_clean <- df1 %>% dplyr::mutate(`..row_id_df1..` = dplyr::row_number())
  df2_clean <- df2 %>% dplyr::mutate(`..row_id_df2..` = dplyr::row_number())
  
  # extract vectors directly from dataframes using base subsetting to bypass join name changes
  focal_vec <- df1_clean[[focal]]
  benef_vec <- df2_clean[[beneficiary]]
  
  # build a row-by-row logical evaluation matrix for all combinations
  match_grid <- base::expand.grid(
    row_df1 = base::seq_along(focal_vec),
    row_df2 = base::seq_along(benef_vec)
  )
  
  # map across combinations using vector indexes rather than data frame strings
  is_partial_vec <- purrr::map2_lgl(match_grid$row_df1, match_grid$row_df2, ~ {
    f_raw <- focal_vec[.x]
    b_raw <- benef_vec[.y]
    
    # early exit check for missing values or empty strings
    if (base::is.na(f_raw) || base::is.na(b_raw) || f_raw == "" || b_raw == "" || f_raw == "NA" || b_raw == "NA") {
      FALSE
    } else {
      f_str <- base::as.character(f_raw)
      b_str <- base::as.character(b_raw)
      
      # basic substring check first using base r grepl to avoid vector length bugs
      matched <- base::grepl(b_str, f_str, fixed = TRUE) || base::grepl(f_str, b_str, fixed = TRUE)
      
      if (base::isTRUE(matched)) {
        matched
      } else {
        # clean and split both strings into unique words
        f_words <- base::unique(base::unlist(stringr::str_split(f_str, "\\s+")))
        b_words <- base::unique(base::unlist(stringr::str_split(b_str, "\\s+")))
        
        # calculate intersect count of matching string tokens across both word arrays
        shared_tokens <- base::intersect(f_words, b_words)
        
        # enforce condition tracking threshold counts to capture out of order permutations
        base::length(shared_tokens) >= min_shared_words
      }
    }
  })
  
  # filter combination map down to verified indexes only
  valid_matches <- match_grid[is_partial_vec, ]
  
  # link valid matches back to strings and calculate Jaro-Winkler proximity metrics
  matches_mapped <- valid_matches %>%
    dplyr::mutate(
      name_possible = focal_vec[row_df1],
      target_benef  = benef_vec[row_df2],
      # compute probability score between 0 and 1 using jaro winkler metric
      match_prob = stringdist::stringsim(name_possible, target_benef, method = "jw")
    ) %>%
    # group by the df2 row index to isolate competing candidates
    dplyr::group_by(row_df2) %>%
    # keep only the single highest ranking match, choosing the first if tied
    dplyr::slice_max(match_prob, n = 1, with_ties = FALSE) %>%
    dplyr::ungroup() %>%
    dplyr::select(row_df2, name_possible)
  
  # join matched values onto df2 and drop internal row indexing trackers
  df2_clean %>%
    dplyr::left_join(matches_mapped, by = c("..row_id_df2.." = "row_df2")) %>%
    dplyr::select(-`..row_id_df2..`)
}
```


## get personally identifying columns

```{r function-get_pii}
#' Get personally identifying columns
#'
#' @param df dataframe
#' @param vars string of identifying columns
#'
#' @returns a list of 3 dataframes: identifying, de-identified, and original df
#' @export
#'
#' @examples
#' \dontrun{get_pii()}
get_pii <- function(
    df = df_clean,
    vars = c("lat", "long", "name", "mobile", "gps", "phone", "enum_num$", "^(enum|reg|ben|proxy|sec)_.*(name|num|id)")
){
  cols <- paste(vars, collapse = "|")
  df_pii <- df %>% select(
    grep(cols, names(.))
  )
  df_depii <- df %>%
    dplyr::select(
      -grep(cols, names(.))
    )
  out <- list(df_pii, df_depii, df) %>%
    purrr::set_names(c("personaly-identifying", "de-identified", "original-data"))
  return(out)
}
```

## export a list to excel

```{r function-write_list_to_excel}
#' export a named list of data frames to an excel workbook
#' @param df_list a named list of data frames
#' @param file_path character string for the output path
#' @returns NULL
#' @export
#' @examples
#' \dontrun{write_list_to_excel()}
write_list_to_excel <- function(df_list, file_path) {
  
  # create a new workbook object
  wb <- openxlsx::createWorkbook()
  
  # define a professional header style
  header_style <- createStyle(
    fgFill = "#4F81BD", 
    halign = "center", 
    textDecoration = "bold",
    fontColour = "white",
    border = "bottom"
  )
  
  # iterate through the named list
  # .x is the dataframe, .y is the name of the list element
  purrr::iwalk(df_list, function(.x, .y) {
    
    # add a new worksheet with the name of the list element
    # excel sheet names are limited to 31 characters
    sheet_name <- substr(.y, 1, 31)
    openxlsx::addWorksheet(wb, sheet_name)
    
    # write the data to the sheet
    openxlsx::writeData(wb, sheet_name, .x, headerStyle = header_style)
    
    # adjust column widths to fit content automatically
    openxlsx::setColWidths(wb, sheet_name, cols = 1:ncol(.x), widths = "auto")
    
    # freeze the top row for easier navigation
    openxlsx::freezePane(wb, sheet_name, firstRow = TRUE)
  })
  
  # save the workbook to the specified path
  openxlsx::saveWorkbook(wb, file_path, overwrite = TRUE)
}
```

## flatten dm object

```{r function-flatten_dm_by_names}
#' flatten all tables within a dm object into a list of data frames
#'
#' @param dm_obj a dm data model object
#'
#' @returns a named list of flattened data frames
#' @export
flatten_dm_by_names <- function(dm_obj) {
  # extract the structural table names directly from the dm object
  tbl_names <- base::names(dm_obj)
  
  # loop through the valid names to flatten each table
  df_list <- base::lapply(tbl_names, function(name) {
    dm::dm_flatten_to_tbl(dm_obj, !!name)
  })
  
  # reapply the names to the output list
  names(df_list) <- tbl_names
  
  return(df_list)
}
```



## pre-process unocha registration


```{r function-preprocess_reg}
#' preprocess unocha registrations
#'
#' @param df dataframe
#' @importFrom dm dm_flatten_to_tbl
#' @importFrom dplyr mutate group_by case_when if_else filter bind_rows distinct
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{preprocess_reg()}
#' 

preprocess_reg <- function(df){
  df_out <- df %>% 
  dm_flatten_to_tbl(main)

df_out <- df_out %>%
  mutate(across(where(labelled::is.labelled), as.character))
df_out <- df_out %>%
  dplyr::mutate(
    village = case_when(
      grepl("chini", village, ignore.case = T) & grepl(
        "kalakacha", sublocation, ignore.case = T
      ) ~ "Kalakacha Chini",
      TRUE ~ village
    )
  )
df_out <- df_out %>%
  mutate(
    village = case_when(
      grepl("Emmaus|Kirengender", village, ignore.case = T) ~ "Kirengender",
      grepl("salama|Kajificheni", village, ignore.case = T) ~ "Salama",
      grepl("kalkacha_juu", village, ignore.case = T) ~ "Kalakacha Juu",
      TRUE ~ village
    )
#     troublesome
    ,
    enum_egency = case_when(
      grepl("mitobini", village, ignore.case = T) &
      grepl("pgi", enum_egency, ignore.case = T) ~ "ALDEF",
      # grepl("Milimani", village, ignore.case = T) & 
      #   date(start) >= date("2026-05-13") ~ "ALDEF",
      # grepl("kurole", village, ignore.case = T) ~ "ALDEF",
      grepl("elwachu", village, ignore.case = T) ~ "ALDEF",
      TRUE ~ enum_egency
    )
  )

df_out <- df_out %>% 
  dplyr::mutate(
    ben_first_surname = paste(first_name, third_name),
    ben_first_secname = paste(first_name, second_name)
  )

df_out <- df_out %>% 
  dplyr::mutate(
    HHH_age_calc = if_else(
      !is.na(resp_age), resp_age, HHH_age
    )
  )

df_out <- df_out %>%
  dplyr::mutate(dplyr::across(dplyr::any_of(
    c("county", "subcounty", "ward", "sublocation", "village")
  ), ~ gsub("_", " ", .x)))

df_out <- df_out %>%
  mutate(
    proxy_name = stringr::str_squish(
      paste(
        dplyr::coalesce(name_proxy_first, ""),
        dplyr::coalesce(name_proxy_second, ""),
        dplyr::coalesce(name_proxy_third, "")
      )
    ) 
  )


df_out <- df_out %>%
  dplyr::mutate(dplyr::across(
    dplyr::any_of(
      c(
        "county",
        "subcounty",
        "ward",
        "sublocation",
        "village",
        "ben_name",
        "proxy_name"
      )
    ),
    ~ stringr::str_to_title(stringr::str_squish(.x))
  ))

df_out <- df_out %>%
  dplyr::mutate(
    age_category = cut_numeric(HHH_age_calc),
    hh_size_category = cut_numeric(
      hh_size,
      breaks = c(1, 3, 6, 10, Inf),
      labels = c("1 - 2", "3 - 5", "6 - 9", ">= 10")
    ),
    hh_disabl_category = cut_numeric(
      hh_disability_count,
      breaks = c(0,1, 3, 5, Inf),
      labels = c("0", "1 - 2", "3 - 4", ">= 5")
    )
  )

  
df_out <- df_out %>%
  labelled::set_variable_labels(
    age_category = "Household\n Head Age",
    HHH_chronic = "HH Head\n Chronic illness",
    HHM_chronic = "HH Member\n Chronic illness",
    average_income = "Monthly Income",
    hh_size = "Household Size",
    hh_size_category = "Household Size",
    FCSGName = "Food Consumption\n Score",
    eligibility = "Eligibility",
    enum_egency = "Agency",
    enum_name = "Enumerator\n first name",
    first_name = "respondent\n first name",
    reg_num = "Respondent Phone",
    trust_proxy = "Trust proxy",
    enum_egency = "Partner",
    proxy_num_consent = "proxy consent",
    name_proxy_first = "proxy\n first name",
    proxy_num = "proxy mobile",
    proxy_relation = "proxy relation",
    HHM_disability = "Household\n disability",
    HHH_chronic = "Household\nchronic illness"
  )

# registration started May 3 ----------------------------------------------

 
  
  # distinct(
  #   ben_name, reg_num, ben_id_number, .keep_all = T
  # ) %>%
  # distinct(
  #   reg_num, ben_id_number, .keep_all = T
  # ) %>% 
  
# reg_num can be NA but distinct() would only pick one row


}
```

## process reg

```{r function-preprocess}
#' preprocess unocha registrations
#'
#' @param df dataframe
#' @importFrom dm dm_flatten_to_tbl
#' @importFrom dplyr mutate group_by case_when if_else filter bind_rows distinct
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{preprocess()}
#

preprocess <- function(df){
  df_out <- df %>% mutate(across(where(labelled::is.labelled), as.character))

  df_out <- df_out %>% dplyr::mutate(dplyr::across(dplyr::any_of(
    c("county", "subcounty", "ward", "sublocation", "village")
  ), ~ gsub("_", " ", .x)))
  df_out <- df_out %>% mutate(proxy_name = clean_username(
    paste(
      dplyr::coalesce(name_proxy_first, ""),
      dplyr::coalesce(name_proxy_second, ""),
      dplyr::coalesce(name_proxy_third, "")
    )
  ))
  df_out <- df_out %>% dplyr::mutate(dplyr::across(
    dplyr::any_of(
      c(
        "county",
        "subcounty",
        "ward",
        "sublocation",
        "village",
        "ben_name",
        "proxy_name"
      )
    ), clean_username
  ))
  df_out <- df_out %>% dplyr::mutate(
    age_category = cut_numeric(HHH_age_final),
    hh_size_category = cut_numeric(
      hh_size,
      breaks = c(1, 3, 6, 10, Inf),
      labels = c("1 - 2", "3 - 5", "6 - 9", ">= 10")
    ),
    hh_disabl_category = cut_numeric(
      hh_disability_count,
      breaks = c(0, 1, 3, 5, Inf),
      labels = c("0", "1 - 2", "3 - 4", ">= 5")
    )
  )
  df_out <- df_out %>% labelled::set_variable_labels(
    age_category = "Household\n Head Age",
    HHH_chronic = "HH Head\n Chronic illness",
    HHM_chronic = "HH Member\n Chronic illness",
    average_income = "Monthly Income",
    hh_size = "Household Size",
    hh_size_category = "Household Size",
    FCSGName = "Food Consumption\n Score",
    eligibility = "Eligibility",
    enum_egency = "Agency",
    enum_name = "Enumerator\n first name",
    first_name = "respondent\n first name",
    reg_num = "Respondent Phone",
    trust_proxy = "Trust proxy",
    enum_egency = "Partner",
    proxy_num_consent = "proxy consent",
    name_proxy_first = "proxy\n first name",
    proxy_num = "proxy mobile",
    proxy_relation = "proxy relation",
    HHM_disability = "Household\n disability",
    HHH_chronic = "Household\nchronic illness"
  )
  df_out
}

```


## slice reg data

```{r function-slice_reg}
#' slice registration to match starting dates
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{slice_reg()}
slice_reg <- function(df){
  df_out <- df %>% 
  dplyr::filter(
#     start of reg
    start >= lubridate::date("2026-05-03")
  ) %>%
  dplyr::filter(
    !(
    lubridate::as_date(start) <= lubridate::date("2026-05-06") & 
    grepl("SND", enum_egency, ignore.case = TRUE)
  )
  )
  df_out
}
```


## get partner edit files

```{r function-get_edit_data}
#' Read partner registration edit files
#'
#' @param partner character string of partners e.g., "PGI" 
#' @param cols_exclude character vector of columns to exclude from edits
#' @param sheet_name character vector of workbook sheet name
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_edit_data()}
get_edit_data <- function(
    partner = "PGI",
    cols_exclude = c("village", "ben_name"),
    sheet_name = NULL
){
  p_file <- list.files(
    file.path(here::here(),"edits"), pattern = partner, full.names = TRUE,
    ignore.case = TRUE
  )
  if(!is.null(sheet_name)){
    sheet_f <- sheet_name
  }else{
    sheet_f <- readxl::excel_sheets(p_file)
  }
  
  df <- readxl::read_excel(p_file, sheet = sheet_f) %>%
    clean_names()   %>% mutate(
      ben_name = stringr::str_squish(ben_name) %>% stringr::str_to_title()
    ) 
  df_1 <- df %>%
    
    # left_join(
    #   df_out %>% 
    #   #   filter(
    #   #   grepl("Tana", county, ignore.case = T)
    #   # ) %>%
    #     select(
    #     ben_name, ben_id_number
    #   )
    # ) %>% 
    filter(
      !grepl(paste(cols_exclude, collapse = "|"), var_to_change)
    )
  return(df_1)
}
```



## Proportional to size random sample

```{r function-draw_pps_sample}
#' Draw PPS Household Sample
#'
#' @param data Household-level data frame.
#' @param county_var County variable name.
#' @param village_id_var Village identifier variable name.
#' @param allocation_df County allocation table.
#' @param alloc_size_var County sample size variable.
#' @param seed Random seed.
#'
#' @return Sampled household data frame.
#' @export
draw_pps_sample <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){
  
  set.seed(seed)
  
  # ----------------------------
  # STEP 1: PPS village allocation
  # ----------------------------
  
  village_alloc <- data %>%
    dplyr::count(
      dplyr::across(
        dplyr::all_of(
          c(county_var, village_id_var)
        )
      ),
      name = "village_size"
    ) %>%
    dplyr::left_join(
      allocation_df,
      by = county_var
    ) %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::all_of(county_var)
      )
    ) %>%
    dplyr::mutate(
      
      county_total = sum(village_size),
      
      raw_alloc =
        (.data[[alloc_size_var]] *
           village_size) /
        county_total,
      
      base_alloc = floor(raw_alloc),
      
      frac = raw_alloc - base_alloc,
      
      remaining =
        first(.data[[alloc_size_var]]) -
        sum(base_alloc),
      
      sample_alloc =
        base_alloc +
        as.integer(
          dplyr::row_number(
            dplyr::desc(frac)
          ) <= remaining
        )
      
    ) %>%
    dplyr::ungroup()
  
  # ----------------------------
  # STEP 2: Sample village-by-village
  # ----------------------------
  
  sampled <- village_alloc %>%
    
    dplyr::group_split(
      dplyr::across(
        dplyr::all_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    
    purrr::map_dfr(function(v){
      
      n_take <- v$sample_alloc[[1]]
      
      hh <- data %>%
        dplyr::semi_join(
          v,
          by = c(
            county_var,
            village_id_var
          )
        )
      
      dplyr::slice_sample(
        hh,
        n = min(
          n_take,
          nrow(hh)
        )
      )
      
    })
  
  sampled
}

```

## proportional to size sample Hamilton redistribution

```{r function-draw_pps_hamilton}
#' Draw a Sample of Households Proportional to Village Size
#'
#' This function selects a stratified random sample of individual household rows
#' from within counties. The probability of a household being selected is
#' proportional to the total size (household count) of the village it belongs to,
#' using the Hamilton (largest remainder) method for allocation. After initial
#' allocation, any village falling below \code{min_village_sample} has its
#' allocation lifted to the minimum by borrowing units from the most oversampled
#' donor villages within the same county.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var A bare or quoted column name identifying the county stratum.
#' @param village_id_var A bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing two columns: the county identifier
#'   and the allocated household sample size for that county.
#' @param alloc_size_var A bare or quoted column name in \code{allocation_df}
#'   indicating the target sample size for each county.
#' @param seed An integer used to initialise the pseudo-random number generator
#'   for reproducible sampling. Defaults to the numeric representation of the
#'   current system date.
#' @param min_village_sample A single integer specifying the minimum number of
#'   households that must be sampled from any village. Villages initially allocated
#'   fewer than this value borrow units from the most oversampled villages in the
#'   same county. The effective minimum is capped at the village's actual household
#'   count, so small villages are never over-sampled. Defaults to \code{8L}.
#'
#' @return A data frame with the same columns as \code{data}, containing only the
#'   selected household rows. Villages with a final allocation of zero are excluded
#'   from the output entirely.
#'
#' @details
#' \strong{Allocation (step 1 — Hamilton / largest remainder):}
#' For each county, each village receives a proportional raw allocation
#' \eqn{n_v = N_{county} \times (s_v / S_{county})}, where \eqn{s_v} is the
#' village household count and \eqn{S_{county}} is the county total. Allocations
#' are floored and fractional remainders are distributed one unit at a time to the
#' villages with the largest fractional parts until the county target is met.
#'
#' \strong{Redistribution (step 2 — minimum floor):}
#' Redistribution is performed independently within each county. Donor villages —
#' those whose allocation exceeds their own effective minimum — are sorted in
#' descending order of surplus and drained greedily until the total deficit of
#' sub-minimum villages is recovered or the donor pool is exhausted. A
#' \code{\link[base]{warning}} is issued when a county cannot fully satisfy the
#' minimum for all its villages.
#'
#' \strong{Sampling (step 3):}
#' Within each village, households are drawn via simple random sampling without
#' replacement using random uniform ordering. Villages with a final
#' allocation of zero are silently dropped.
#'
#' @seealso \code{\link[dplyr]{slice_sample}}, \code{\link[dplyr]{group_modify}}
#'
#' @importFrom dplyr count across any_of left_join group_by mutate ungroup group_modify filter arrange row_number select
#' @importFrom rlang .data sym !!
#' @importFrom stats runif
#'
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw household data
#' raw_households <- data.frame(
#'   county_id  = rep(c("County_A", "County_B"), each = 10),
#'   village_id = c(rep("V1", 3), rep("V2", 7), rep("V3", 4), rep("V4", 6)),
#'   household_id = paste0("HH_", 1:20)
#' )
#'
#' # household sample size targets per county
#' target_allocations <- data.frame(
#'   county_id   = c("County_A", "County_B"),
#'   sample_size = c(20, 25)
#' )
#'
#' # run reproducible PPS household sampling with an 8-household village minimum
#' pps_household_sample <- draw_pps_sample(
#'   data               = raw_households,
#'   county_var         = "county_id",
#'   village_id_var     = "village_id",
#'   allocation_df      = target_allocations,
#'   alloc_size_var     = "sample_size",
#'   seed               = 123,
#'   min_village_sample = 8L
#' )
#' }
draw_pps_hamilton <- function(
    data,
    county_var         = "county",
    village_id_var     = "village",
    allocation_df,
    alloc_size_var     = "sample",
    seed               = as.numeric(Sys.Date()),
    min_village_sample = 8L
) {
  
  # preserve random seed state if it exists to maintain reproducibility safely
  if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
    old_seed <- get(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
    on.exit(assign(".Random.seed", old_seed, envir = .GlobalEnv))
  }
  
  set.seed(seed)
  
  # convert input column names safely to standard strings
  county_str <- as.character(substitute(county_var))
  if (!county_str %in% names(data)) county_str <- county_var
  
  village_str <- as.character(substitute(village_id_var))
  if (!village_str %in% names(data)) village_str <- village_id_var
  
  # hamilton pps allocation
  village_alloc <- data %>%
    dplyr::count(
      dplyr::across(dplyr::any_of(c(county_str, village_str))),
      name = "village_size"
    ) %>%
    dplyr::left_join(allocation_df, by = county_str) %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(county_str))) %>%
    dplyr::mutate(
      county_total = sum(.data$village_size),
      raw_alloc    = .data[[alloc_size_var]] * .data$village_size / .data$county_total,
      sample_alloc = floor(.data$raw_alloc),
      remainder    = .data[[alloc_size_var]][1] - sum(floor(.data$raw_alloc)),
      frac         = .data$raw_alloc - floor(.data$raw_alloc),
      sample_alloc = .data$sample_alloc +
        ifelse(rank(-.data$frac, ties.method = "first") <= .data$remainder, 1L, 0L)
    ) %>%
    dplyr::ungroup()
  
  # minimum size redistribution
  village_alloc <- village_alloc %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(county_str))) %>%
    dplyr::group_modify(function(county_df, key) {
      
      alloc   <- county_df$sample_alloc
      
      # enforce the floor to be exactly min_village_sample capped at available population
      eff_min <- pmin(min_village_sample, county_df$village_size)
      
      deficit_idx <- which(alloc < eff_min)
      if (length(deficit_idx) == 0L) return(county_df)
      
      deficit        <- eff_min[deficit_idx] - alloc[deficit_idx]
      total_deficit  <- sum(deficit)
      surplus        <- pmax(alloc - eff_min, 0L)
      total_surplus  <- sum(surplus)
      
      if (total_surplus == 0L) {
        warning(
          "County '", key[[1]], "': not enough surplus to satisfy minimum of ",
          min_village_sample, " for all villages. Lifting as much as possible."
        )
      }
      
      donor_order <- order(-surplus)
      remaining   <- min(total_deficit, total_surplus)
      
      for (i in donor_order) {
        if (remaining <= 0L) break
        give      <- min(surplus[i], remaining)
        alloc[i]  <- alloc[i] - give
        remaining <- remaining - give
      }
      
      recovered <- min(total_deficit, total_surplus)
      for (j in seq_along(deficit_idx)) {
        if (recovered <= 0L) break
        give                  <- min(deficit[j], recovered)
        alloc[deficit_idx[j]] <- alloc[deficit_idx[j]] + give
        recovered             <- recovered - give
      }
      
      county_df$sample_alloc <- alloc
      county_df
    }) %>%
    dplyr::ungroup()
  
  # sample households within each village
  # join final target allocation boundaries back to full population dataset rows
  joined_data <- data %>%
    dplyr::left_join(
      dplyr::select(village_alloc, dplyr::any_of(c(county_str, village_str, "sample_alloc"))),
      by = c(county_str, village_str)
    )
  
  # shuffle and slice rows matching exact targets without using group modify
  sampled <- joined_data %>%
    dplyr::filter(!is.na(.data$sample_alloc) & .data$sample_alloc > 0L) %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(c(county_str, village_str)))) %>%
    dplyr::mutate(rand_draw_order = stats::runif(dplyr::n())) %>%
    dplyr::arrange(.data$rand_draw_order, .by_groups = TRUE) %>%
    dplyr::filter(dplyr::row_number() <= .data$sample_alloc) %>%
    dplyr::ungroup() %>%
    dplyr::select(-dplyr::any_of(c("sample_alloc", "rand_draw_order")))
  
  return(sampled)
}

```


## Stratified sampling

```{r function-draw_strs_sample}
#' Draw a Stratified Simple Random Sample of Households
#'
#' This function selects a stratified simple random sample of individual household 
#' rows from within counties. Every household within a given county has an equal 
#' probability of selection, independent of village size.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param allocation_df A data frame containing two columns: the county identifier 
#'   and the allocated household sample size for that county.
#' @param alloc_size_var The bare or quoted column name in `allocation_df` 
#'   indicating the target sample size.
#' @param seed An integer value used to initialize the pseudo-random number generator 
#'   for reproducible sampling. Defaults to `42`.
#'
#' @return A data frame containing the individually selected household rows.
#'
#' @importFrom dplyr group_by mutate ungroup inner_join arrange row_number filter across any_of desc n
#' @importFrom rlang sym
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw household data
#' raw_households <- data.frame(
#'   county_id = rep(c("County_A", "County_B"), each = 10),
#'   village_id = c(rep("V1", 3), rep("V2", 7), rep("V3", 4), rep("V4", 6)),
#'   household_id = paste0("HH_", 1:20)
#' )
#'
#' # household sample size targets per county
#' target_allocations <- data.frame(
#'   county_id = c("County_A", "County_B"),
#'   sample_size = c(2, 3)
#' )
#'
#' # run reproducible stratified simple random sampling
#' strs_sample <- draw_strs_sample(
#'   data = raw_households,
#'   county_var = "county_id",
#'   allocation_df = target_allocations,
#'   alloc_size_var = "sample_size",
#'   seed = 123
#' )
#' }
draw_strs_sample <- function(
    data,
    county_var = "county",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){

  set.seed(seed)

  joined_frame <- data %>%
    dplyr::inner_join(
      allocation_df,
      by = county_var
    )

  joined_frame %>%
    dplyr::group_by(
      dplyr::across(dplyr::any_of(county_var))
    ) %>%
    dplyr::slice_sample(
      n = min(dplyr::n(), first(.data[[alloc_size_var]]))
    ) %>%
    dplyr::ungroup()
}
```


## two stage cluster random sampling

```{r function-draw_cluster_sample}
#' Draw a Two-Stage Cluster Sample of Households
#'
#' This function performs two-stage cluster sampling within counties. In the 
#' first stage, a specified number of unique villages are selected randomly. In the 
#' second stage, a fixed number of households are randomly drawn from each 
#' of those selected villages. Both target sizes are pulled from the allocation frame.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param village_id_var The bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing columns for the county identifier, 
#'   the target village sample size, and the target household sample size per village.
#' @param alloc_villages_var The bare or quoted column name in `allocation_df` 
#'   indicating the number of villages to sample per county.
#' @param alloc_hh_var The bare or quoted column name in `allocation_df` 
#'   indicating the number of households to sample per chosen village.
#' @param seed An integer value used to initialize the pseudo-random number generator 
#'   for reproducible sampling. Defaults to `42`.
#'
#' @return A data frame containing the cluster-sampled household rows.
#'
#' @importFrom dplyr group_by mutate ungroup arrange row_number filter across any_of distinct n inner_join select
#' @importFrom rlang sym
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw household data
#' raw_households <- data.frame(
#'   county_id = rep(c("County_A", "County_B"), each = 15),
#'   village_id = c(rep("V1", 5), rep("V2", 5), rep("V3", 5), 
#'                  rep("V4", 5), rep("V5", 5), rep("V6", 5)),
#'   household_id = paste0("HH_", 1:30)
#' )
#'
#' # multi-stage target allocations dataframe
#' cluster_allocations <- data.frame(
#'   county_id = c("County_A", "County_B"),
#'   target_villages = c(2, 2),
#'   target_hh_per_village = c(3, 2)
#' )
#'
#' # run dynamic cluster sampling
#' cluster_sample <- draw_cluster_sample(
#'   data = raw_households,
#'   county_var = "county_id",
#'   village_id_var = "village_id",
#'   allocation_df = cluster_allocations,
#'   alloc_villages_var = "target_villages",
#'   alloc_hh_var = "target_hh_per_village",
#'   seed = 42
#' )
#' }
draw_cluster_sample <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_villages_var = "target_villages",
    alloc_hh_var = "target_hh_per_village",
    seed = 42
){

  set.seed(seed)

  selected_villages <- data %>%
    dplyr::distinct(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    dplyr::inner_join(
      allocation_df,
      by = county_var
    ) %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(county_var)
      )
    ) %>%
    dplyr::slice_sample(
      n = min(
        dplyr::n(),
        first(.data[[alloc_villages_var]])
      )
    ) %>%
    dplyr::ungroup()

  data %>%
    dplyr::inner_join(
      selected_villages,
      by = c(county_var, village_id_var)
    ) %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    dplyr::slice_sample(
      n = min(
        dplyr::n(),
        first(.data[[alloc_hh_var]])
      )
    ) %>%
    dplyr::ungroup()
}
```

## systematic sampling county level

```{r function-draw_systematic_across_villages}
#' Draw a Systematic Sample of Households Spreading Across Villages
#'
#' This function selects a systematic random sample of individual household rows 
#' from within counties. By sorting by village first, the systematic interval 
#' automatically spreads the selected sample evenly across the different villages.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param village_id_var The bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing two columns: the county identifier 
#'   and the allocated household sample size for that county.
#' @param alloc_size_var The bare or quoted column name in `allocation_df` 
#'   indicating the target sample size.
#' @param seed An integer value used to initialize the pseudo-random number generator. 
#'   Defaults to `42`.
#'
#' @return A data frame containing the systematically selected household rows.
#'
#' @importFrom dplyr group_by mutate ungroup inner_join arrange row_number filter across any_of n select
#' @importFrom rlang sym
#' @export
draw_systematic_across_villages <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){

  set.seed(seed)

  joined_frame <- data %>%
    dplyr::inner_join(
      allocation_df,
      by = county_var
    ) %>%
    dplyr::arrange(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    )

  joined_frame %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(county_var)
      )
    ) %>%
    dplyr::group_modify(~{

      df <- .x

      N <- nrow(df)
      n_target <- min(
        first(df[[alloc_size_var]]),
        N
      )

      k <- N / n_target

      start <- sample(
        seq_len(ceiling(k)),
        1
      )

      idx <- floor(
        start + (0:(n_target-1))*k
      )

      idx <- pmin(idx, N)

      df[idx, , drop = FALSE]

    }) %>%
    dplyr::ungroup()
}
```


## systematic sample village level allocation

```{r function-draw_systematic_within_villages}
#' Draw a Systematic Sample of Households Within Each Village
#'
#' This function applies systematic random sampling to households independently 
#' within each village cluster based on specific target allocations for each village.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param village_id_var The bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing columns for the village identifier 
#'   and its assigned household sample target.
#' @param alloc_size_var The bare or quoted column name in `allocation_df` 
#'   indicating the village-level target sample size.
#' @param seed An integer value used to initialize the pseudo-random number generator. 
#'   Defaults to `42`.
#'
#' @return A data frame containing the systematically selected household rows.
#'
#' @importFrom dplyr group_by mutate ungroup inner_join arrange row_number filter across any_of n select
#' @importFrom rlang sym
#' @export
draw_systematic_within_villages <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){

  set.seed(seed)

  joined_frame <- data %>%
    dplyr::inner_join(
      allocation_df,
      by = village_id_var
    ) %>%
    dplyr::arrange(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    )

  joined_frame %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    dplyr::group_modify(~{

      df <- .x

      N <- nrow(df)

      n_target <- min(
        first(df[[alloc_size_var]]),
        N
      )

      k <- N / n_target

      start <- sample(
        seq_len(ceiling(k)),
        1
      )

      idx <- floor(
        start + (0:(n_target-1))*k
      )

      idx <- pmin(idx, N)

      df[idx, , drop = FALSE]

    }) %>%
    dplyr::ungroup()
}
```

## create survey object

```{r function-create_survey_object}
#' create a survey design object dynamically for any complex sampling design
#'
#' @param sample_df dataframe containing the final selected sample observations
#' @param population_df dataframe containing the full original population sampling frame
#' @param id_cols character vector identifying structural cluster nodes in cascading order (e.g., c("village", "hh_id"))
#' @param strata_cols character vector containing names of stratification columns (optional, defaults to NULL)
#' @param weight_col_name character string naming the output column for calculated weights
#'
#' @returns a configured survey design object
#' @export
create_survey_object <- function(sample_df, 
                                        population_df, 
                                        id_cols, 
                                        strata_cols = NULL, 
                                        weight_col_name = "sampling_weight") {
  
  # ensure package namespace dependencies are available
  if (!base::requireNamespace("survey", quietly = TRUE)) {
    base::stop("the 'survey' package is required but not installed")
  }
  
  # handle missing variable safety up front
  all_vars <- base::unique(c(id_cols, strata_cols))
  missing_sample <- all_vars[!all_vars %in% base::names(sample_df)]
  missing_pop    <- all_vars[!all_vars %in% base::names(population_df)]
  
  if (base::length(missing_sample) > 0 || base::length(missing_pop) > 0) {
    base::stop("specified structural columns missing from sample or population dataframes")
  }
  
  # step 1: isolate stage 1 units (primary sampling units) to evaluate selection probabilities
  psu_col <- id_cols[1]
  stage1_grouping <- base::unique(c(strata_cols, psu_col))
  
  pop_psus <- population_df %>%
    dplyr::distinct(dplyr::across(dplyr::all_of(stage1_grouping))) %>%
    dplyr::count(dplyr::across(dplyr::all_of(strata_cols)), name = "..total_PSUs_in_strata..")
  
  sample_psus <- sample_df %>%
    dplyr::distinct(dplyr::across(dplyr::all_of(stage1_grouping))) %>%
    dplyr::count(dplyr::across(dplyr::all_of(strata_cols)), name = "..sampled_PSUs_in_strata..")
  
  # map stage 1 probability across rows
  stage1_probs <- population_df %>%
    dplyr::distinct(dplyr::across(dplyr::all_of(stage1_grouping))) %>%
    dplyr::left_join(pop_psus, by = strata_cols) %>%
    dplyr::left_join(sample_psus, by = strata_cols) %>%
    # if unstratified, handle empty grouping count variables safely
    dplyr::mutate(
      ..total_PSUs_in_strata..   = base::ifelse(base::is.na(..total_PSUs_in_strata..), base::length(base::unique(population_df[[psu_col]])), ..total_PSUs_in_strata..),
      ..sampled_PSUs_in_strata.. = base::ifelse(base::is.na(..sampled_PSUs_in_strata..), base::length(base::unique(sample_df[[psu_col]])), ..sampled_PSUs_in_strata..),
      ..prob_stage1..            = ..sampled_PSUs_in_strata.. / ..total_PSUs_in_strata..
    ) %>%
    dplyr::select(dplyr::all_of(stage1_grouping), "..prob_stage1..")
  
  # step 2: evaluate stage 2 units (secondary sampling units) if a second id column is specified
  if (base::length(id_cols) > 1) {
    ssu_grouping <- base::unique(c(strata_cols, id_cols[1:2]))
    
    pop_ssus <- population_df %>%
      dplyr::count(dplyr::across(dplyr::all_of(id_cols[1])), name = "..total_SSUs_in_PSU..")
    
    sample_ssus <- sample_df %>%
      dplyr::count(dplyr::across(dplyr::all_of(id_cols[1])), name = "..sampled_SSUs_in_PSU..")
    
    stage2_probs <- pop_ssus %>%
      dplyr::left_join(sample_ssus, by = id_cols[1]) %>%
      dplyr::mutate(
        ..sampled_SSUs_in_PSU.. = base::ifelse(base::is.na(..sampled_SSUs_in_PSU..), 0, ..sampled_SSUs_in_PSU..),
        ..prob_stage2..         = ..sampled_SSUs_in_PSU.. / ..total_SSUs_in_PSU..
      ) %>%
      dplyr::select(dplyr::all_of(id_cols[1]), "..prob_stage2..")
    
    # join stage 1 and stage 2 tracks together to compute joint weights
    final_weights <- stage1_probs %>%
      dplyr::inner_join(stage2_probs, by = id_cols[1]) %>%
      dplyr::mutate(
        !!weight_col_name := 1 / (..prob_stage1.. * ..prob_stage2..)
      ) %>%
      dplyr::select(dplyr::all_of(c(id_cols[1], weight_col_name)))
    
  } else {
    # single stage designs pass flat inverse probability configurations down stream
    final_weights <- stage1_probs %>%
      dplyr::mutate(
        !!weight_col_name := 1 / ..prob_stage1..
      ) %>%
      dplyr::select(dplyr::all_of(c(stage1_grouping, weight_col_name)))
  }
  
  # step 3: combine finalized weights lookup to sample frame and create svydesign object
  sample_with_weights <- sample_df %>%
    dplyr::inner_join(final_weights, by = id_cols[1])
  
  id_formula     <- stats::as.formula(base::paste0("~", base::paste(id_cols, collapse = " + ")))
  strata_formula <- if (!base::is.null(strata_cols)) stats::as.formula(base::paste0("~", base::paste(strata_cols, collapse = " + "))) else NULL
  weight_formula <- stats::as.formula(base::paste0("~", weight_col_name))
  
  survey_obj <- survey::svydesign(
    ids = id_formula,
    strata = strata_formula,
    weights = weight_formula,
    data = sample_with_weights
  )
  
  return(survey_obj)
}
```


## apply corrections

```{r function-apply_corrections}
#' Apply Data Cleaning Corrections Using an Embedded Cleaning Log
#'
#' This function updates a main dataset using a correction data frame. It applies
#' standard cell corrections and handles row deletions last. Deletions are processed 
#' using strict unique UUID matching if a true uuid column is present, falling back to 
#' exact name matching combined with multiple context validation columns if a general 
#' non-unique identifier or name-only tracking is used.
#'
#' @param data A data frame containing the main dataset to be cleaned.
#' @param id_var The bare or quoted column name representing the unique identifier.
#' @param corrections A data frame containing the corrections log along with the matching keys.
#' @param var_to_change_col Column name in `corrections` holding target names to change.
#' @param correct_value_col Column name in `corrections` holding updated values.
#' @param ben_name_col Column name in both datasets holding beneficiary names.
#' @param context_cols Optional character vector. Secondary column names present in both 
#'   datasets (e.g., `c("ben_id_number", "ward", "village")`) used to build a composite key 
#'   and avoid accidental collisions of identical or similar names.
#'
#' @return A data frame matching the structure of `data`.
#' @export
apply_corrections <- function(
    data, 
    id_var = "uuid", 
    corrections, 
    var_to_change_col = "var_to_change", 
    correct_value_col = "correct_value",
    ben_name_col = "ben_name",
    context_cols = c("ben_id_number", "ward", "village")
) {
  
  id_str <- as.character(substitute(id_var))
  if (!id_str %in% names(data)) id_str <- id_var
  
  var_change_str <- as.character(substitute(var_to_change_col))
  if (!var_change_str %in% names(corrections)) var_change_str <- var_to_change_col
  
  correct_val_str <- as.character(substitute(correct_value_col))
  if (!correct_val_str %in% names(corrections)) correct_val_str <- correct_value_col
  
  ben_name_str <- as.character(substitute(ben_name_col))
  if (!ben_name_str %in% names(corrections)) ben_name_str <- ben_name_col
  
  cleaned_data <- data
  
  # extract only the unique variables that actually exist in the main data
  distinct_vars <- corrections %>%
    dplyr::filter(!!rlang::sym(var_change_str) %in% names(cleaned_data)) %>%
    dplyr::pull(!!rlang::sym(var_change_str)) %>%
    unique()
  
  # 1 execute standard cell corrections first
  for (current_var in distinct_vars) {
    sub_corrections <- corrections %>%
      dplyr::filter(!!rlang::sym(var_change_str) == current_var) %>%
      dplyr::select(dplyr::any_of(c(ben_name_str, correct_val_str)))
    
    target_class <- class(cleaned_data[[current_var]])[1]
    
    if (target_class %in% c("numeric", "integer")) {
      sub_corrections[[correct_val_str]] <- as.numeric(sub_corrections[[correct_val_str]])
    } else if (target_class == "logical") {
      sub_corrections[[correct_val_str]] <- as.logical(sub_corrections[[correct_val_str]])
    } else {
      sub_corrections[[correct_val_str]] <- as.character(sub_corrections[[correct_val_str]])
    }
    
    join_target_name <- paste0("new_val_", current_var)
    sub_corrections <- dplyr::rename(
      sub_corrections, 
      !!rlang::sym(join_target_name) := !!rlang::sym(correct_val_str)
    )
    
    cleaned_data <- cleaned_data %>%
      dplyr::left_join(sub_corrections, by = ben_name_str) %>%
      dplyr::mutate(
        !!rlang::sym(current_var) := ifelse(
          !is.na(!!rlang::sym(join_target_name)), 
          !!rlang::sym(join_target_name), 
          !!rlang::sym(current_var)
        )
      ) %>%
      dplyr::select(-dplyr::any_of(join_target_name))
  }
  
  # 2 handle deletions as the final operation with compound keys
  if (var_change_str %in% names(corrections) && ben_name_str %in% names(corrections) && ben_name_str %in% names(data)) {
    
    # filter log down strictly to the rows designated for deletion using case insensitive grepl
    delete_log <- corrections %>%
      dplyr::filter(grepl("delete", !!rlang::sym(var_change_str), ignore.case = TRUE))
    
    if (nrow(delete_log) > 0) {
      
      # premium bypass: only use direct ID deletion if it is explicitly a unique uuid column
      if (tolower(id_str) == "uuid" && id_str %in% names(corrections) && id_str %in% names(data)) {
        
        delete_ids <- delete_log %>%
          dplyr::pull(!!rlang::sym(id_str)) %>%
          unique()
        
        cleaned_data <- cleaned_data %>%
          dplyr::filter(!(!!rlang::sym(id_str) %in% delete_ids))
        
      } else {
        # traditional fallback: filter context columns down to only those that actually exist in both datasets
        # this ensures that if id_str is "ben_id_number", it gets lumped safely into the compound validation checks
        all_potential_contexts <- unique(c(id_str, context_cols))
        valid_contexts <- all_potential_contexts[all_potential_contexts %in% names(data) & all_potential_contexts %in% names(corrections)]
        
        if (length(valid_contexts) > 0) {
          # combine beneficiary name and all valid context values into a multi-variable string key
          # using individual rowwise evaluations to avoid vector collapse traps
          build_composite_key <- function(df, name_col, ctx_cols) {
            df %>%
              dplyr::rowwise() %>%
              dplyr::mutate(
                comp_key = paste(
                  c(as.character(.data[[name_col]]), mget(ctx_cols)), 
                  collapse = "|||"
                )
              ) %>%
              dplyr::ungroup() %>%
              dplyr::pull(comp_key)
          }
          
          # generate keys for both datasets
          delete_keys <- build_composite_key(delete_log, ben_name_str, valid_contexts)
          
          # apply filter by checking the main data against the delete keys
          cleaned_data <- cleaned_data %>%
            dplyr::rowwise() %>%
            dplyr::filter(
              !paste(c(as.character(.data[[ben_name_str]]), mget(valid_contexts)), collapse = "|||") %in% delete_keys
            ) %>%
            dplyr::ungroup()
          
        } else {
          # fallback to strict exact name checking if no valid context variables were matched
          delete_names <- unique(delete_log[[ben_name_str]])
          cleaned_data <- cleaned_data %>%
            dplyr::filter(!(!!rlang::sym(ben_name_str) %in% delete_names))
        }
      }
    }
  }
  
  return(cleaned_data)
}

```

## read excel workbook

```{r function-read_workbook}
#' Read Excel sheets or an entire workbook into a list or data frame with clean names
#'
#' @description
#' Iterates through available worksheets within an Excel file or targets a specific sheet, 
#' extracts content while skipping a designated number of rows, and standardizes column names 
#' using snake_case. If the target evaluation resolves to exactly one worksheet, it automatically 
#' drops the list format and returns a single data frame directly.
#'
#' @param path Character string representing the path to the Excel file.
#' @param bind_rows Logical indicator. If TRUE, binds multiple extracted sheets into 
#'   a single unified data frame using a compiler identifier column. Defaults to FALSE.
#' @param id Character string naming the optional identifier column that tracks 
#'   the source sheet name when `bind_rows` is TRUE. Defaults to "sheet_name".
#' @param skip Integer specifying the number of rows to skip at the start of reading. Defaults to 0.
#' @param sheet_name Character string specifying a single sheet to read. If NULL, all sheets are processed. Defaults to NULL.
#'
#' @return A single data frame if only one sheet is processed or `bind_rows = TRUE`. 
#'   Otherwise, a named list of data frames representing individual sheets.
#'
#' @importFrom readxl excel_sheets read_excel
#' @importFrom purrr set_names map list_rbind
#' @importFrom janitor clean_names
#'
#' @export
#'
#' @examples
#' \dontrun{
#' # read a specific sheet and skip the first 2 rows
#' df_raw <- read_workbook("survey_data.xlsx", sheet_name = "Main_Data", skip = 2)
#' }
read_workbook <- function(path, bind_rows = FALSE, id = "sheet_name", skip = 0, sheet_name = NULL) {
  
  # determine target sheet names based on user input selection
  if (!is.null(sheet_name)) {
    sheet_names <- sheet_name
  } else {
    sheet_names <- readxl::excel_sheets(path)
  }
  
  # short circuit and return a single data frame instantly if only one sheet is target evaluated
  if (length(sheet_names) == 1) {
    single_df <- readxl::read_excel(path = path, sheet = sheet_names, skip = skip) %>%
      janitor::clean_names()
    return(single_df)
  }
  
  # extract sheet names, read contents, and standardize variables to snake case
  sheets_list <- sheet_names %>%
    purrr::set_names() %>%
    purrr::map(~ {
      readxl::read_excel(path = path, sheet = .x, skip = skip) %>%
        janitor::clean_names()
    })
  
  # conditionally collapse the list rows into a single data frame if specified
  if (bind_rows) {
    combined_data <- purrr::list_rbind(sheets_list, names_to = id)
    return(combined_data)
  }
  
  return(sheets_list)
}
```


## standardize names

```{r function-standardize_names}
#' Map and Standardize Administrative Names to Reference Values
#'
#' This function takes a dataset containing entered administrative names (such as
#' sublocations or villages) and cross-references them against an official lookup
#' table. It matches names using an optimal string alignment distance algorithm
#' to automatically correct typos, casing differences, and minor variations.
#'
#' @param data A data frame containing the main survey or household records.
#' @param match_col A bare or quoted column name in \code{data} containing the 
#'   entered names that need correction.
#' @param ref_df A data frame serving as the official reference lookup table.
#' @param ref_col A bare or quoted column name in \code{ref_df} containing the
#'   true, standardized reference names.
#' @param max_dist A numeric value specifying the maximum allowed string distance
#'   for a match to be considered valid. Greater values allow more generous 
#'   typo corrections. Defaults to \code{2}.
#'
#' @return A data frame with the same rows as \code{data}, where the original
#'   \code{match_col} values are replaced by the true values from \code{ref_col}.
#'   Unmatched records will retain their original values or return \code{NA} if 
#'   completely outside the distance threshold.
#'
#' @importFrom dplyr mutate row_number across any_of
#' @importFrom rlang .data
#' @importFrom stringdist amatch
#'
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw survey data with typos
#' survey_records <- data.frame(
#'   household_id = 1:4,
#'   entered_sublocation = c("Muguga ", "Kikuyu", "sigona", "UnknownPlace")
#' )
#'
#' # true official reference registry
#' official_sublocations <- data.frame(
#'   true_name = c("Muguga", "Kikuyu", "Sigona", "Karai")
#' )
#'
#' # clean and standardize the sublocation entries
#' cleaned_data <- standardize_names(
#'   data      = survey_records,
#'   match_col = "entered_sublocation",
#'   ref_df    = official_sublocations,
#'   ref_col   = "true_name",
#'   max_dist  = 3
#' )
#' }
standardize_names <- function(
    data,
    match_col,
    ref_df,
    ref_col,
    max_dist = 2
) {
  
  # convert input column names safely to standard strings
  match_str <- as.character(substitute(match_col))
  if (!match_str %in% names(data)) match_str <- match_col
  
  ref_str <- as.character(substitute(ref_col))
  if (!ref_str %in% names(ref_df)) ref_str <- ref_col
  
  # extract the target reference vector clean of missing entries
  ref_vector <- as.character(na.omit(ref_df[[ref_str]]))
  
  # extract and clean input entries for robust string matching
  raw_entries <- as.character(data[[match_str]])
  clean_raw   <- trimws(tolower(raw_entries))
  clean_ref   <- trimws(tolower(ref_vector))
  
  # find matching vector indices based on optimal string alignment distance
  match_indices <- stringdist::amatch(
    x          = clean_raw,
    table      = clean_ref,
    method     = "osa",
    maxDist    = max_dist
  )
  
  # map matched indices back to original true casing from reference list
  standardized_values <- ifelse(
    !is.na(match_indices),
    ref_vector[match_indices],
    raw_entries
  )
  
  # assign corrected values back to the original target column mask
  standardized_df <- data %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::any_of(match_str),
        ~ standardized_values
      )
    )
  
  return(standardized_df)
}
```


## find name matches

```{r function-find_name_matches}
#' Identify Potential Beneficiary Name Matches From a Master Dataset
#'
#' @description
#' Compares a column of names in a flagged dataset against a column of names in a 
#' master reference dataset using optimized vectorized string distance matrices. 
#' If a potential match is identified within a specified text edit threshold, 
#' the candidate name from the master dataset is mapped into a new column named `name_possible`.
#'
#' @param flagged_data A data frame containing the flagged records needing verification.
#' @param master_data A data frame containing the verified master benchmark records.
#' @param flagged_col The bare or quoted column name of the name column in `flagged_data`. Defaults to "ben_name".
#' @param master_col The bare or quoted column name of the name column in `master_data`. Defaults to "ben_name".
#' @param max_dist Integer specifying the maximum allowed string distance/edits to qualify as a match. Defaults to 3.
#'
#' @return A data frame matching the structure of `flagged_data`, with an additional 
#'   `name_possible` character column containing the matched master dataset name (or NA if no match is found).
#'
#' @importFrom dplyr mutate left_join select any_of tibble
#' @importFrom utils adist
#' @importFrom rlang sym !! :=
#' @importFrom stringdist stringdistmatrix
#'
#' @export
#'
#' @examples
#' \dontrun{
#' verified_df <- find_name_matches(bad_data, good_data, max_dist = 2)
#' }


find_name_matches <- function(
    flagged_data,
    master_data,
    flagged_col = "ben_name",
    master_col  = "ben_name",
    max_dist    = 3
) {
  flag_str <- as.character(substitute(flagged_col))
  if (!flag_str %in% names(flagged_data)) flag_str <- flagged_col
  
  master_str <- as.character(substitute(master_col))
  if (!master_str %in% names(master_data)) master_str <- master_col
  
  master_pool <- unique(na.omit(as.character(master_data[[master_str]])))
  flagged_vec <- as.character(flagged_data[[flag_str]])
  
  sort_tokens <- function(x) {
    vapply(
      strsplit(tolower(trimws(x)), "[[:space:]\\-]+"),
      function(toks) paste(sort(toks), collapse = " "),
      character(1)
    )
  }
  
  flagged_lower <- tolower(flagged_vec)
  master_lower  <- tolower(master_pool)
  flagged_tok   <- sort_tokens(flagged_vec)
  master_tok    <- sort_tokens(master_pool)
  
  # blocking key: first character of the first token
  flagged_block <- substr(flagged_tok, 1, 1)
  master_block  <- substr(master_tok,  1, 1)
  
  name_possible <- rep(NA_character_, length(flagged_vec))
  
  for (block in unique(flagged_block)) {
    fi <- which(flagged_block == block)   # row indices in flagged
    mi <- which(master_block  == block)   # candidate indices in master
    
    if (length(fi) == 0 || length(mi) == 0) next
    
    dist_whole <- stringdist::stringdistmatrix(
      flagged_lower[fi], master_lower[mi],
      method = "osa", nthread = parallel::detectCores()
    )
    dist_token <- stringdist::stringdistmatrix(
      flagged_tok[fi], master_tok[mi],
      method = "osa", nthread = parallel::detectCores()
    )
    dist_best <- pmin(dist_whole, dist_token)
    
    best_col  <- max.col(-dist_best, ties.method = "first")
    best_dist <- dist_best[cbind(seq_along(fi), best_col)]
    
    matched <- best_dist <= max_dist
    name_possible[fi[matched]] <- master_pool[mi[best_col[matched]]]
  }
  
  flagged_data[["name_possible"]] <- name_possible
  flagged_data
}

```


## match county records

```{r function-match_county_records}
#' Filter and Match Payroll Records Against Registration Data by County
#'
#' @description
#' Subsets registration data for a specific county, removes duplicate IDs,
#' joins the results with payroll records, and flags rows where the payroll name
#' is a partial match inside the registration name string.
#'
#' @param registration A data frame containing the primary registration benchmark records.
#' @param payroll A data frame containing the target payroll records to verify.
#' @param county_name Character string specifying the county name to filter on. Defaults to "Tana".
#'
#' @return A data frame matching the layout of the payroll dataset, appended with 
#'   columns from the subsetted registration data and a logical `names_match` flag.
#'
#' @importFrom dplyr filter select distinct left_join mutate
#' @importFrom stringi stri_detect_fixed
#'
#' @export
#' @examples
#' \dontrun{match_county_records()}
match_county_records <- function(registration, payroll, county_name = "Tana") {
  # filter registration pool for the target county and drop duplicate id records
  df_county_ref <- registration %>%
    dplyr::filter(grepl(county_name, county, ignore.case = TRUE)) %>%
    dplyr::select(ben_name, reg_num, ben_id_number) %>%
    dplyr::distinct(ben_id_number, .keep_all = TRUE)
  
  # merge payroll records with the clean county reference subset
  df_joined <- payroll %>%
    dplyr::left_join(
      df_county_ref,
      by = c("id_number" = "ben_id_number")
    ) %>%
    # remove residual system export columns starting with x or X
    dplyr::select(-grep("^[xX]", names(.)))
  
  # fast vectorized string detection — no r loops or nested scans
  df_joined %>%
    dplyr::mutate(
      names_match = !is.na(name_payroll) & 
        !is.na(ben_name) & 
        name_payroll != "" & 
        stringi::stri_detect_fixed(ben_name, name_payroll, case_insensitive = TRUE)
    )
}
```


## normalize names

```{r function-normalize_names}
#' Normalize a Name Vector
#'
#' @description
#' Cleans a character vector of names by collapsing extra whitespace,
#' trimming leading/trailing spaces, and converting to title case.
#'
#' @param x A character vector of names.
#'
#' @return A cleaned character vector.
#'
#' @importFrom stringr str_squish str_to_title
#'
#' @export
#' @examples
#' \dontrun{normalize_names()}
normalize_names <- function(x) {
  x <- stringr::str_squish(x)       # trims + collapses internal spaces
  x <- stringr::str_to_title(x)     # title case
  x
}
```


## Veification criteria table

```{r function-gen_verification_criteria_table}
#' Create the target beneficiary verification criteria and indicators summary
#'
#' @description
#' Builds a two column matrix pairing each humanitarian vulnerability 
#' criterion alongside its designated operational indicator visually 
#' merging repeating criteria rows and introducing complete outer frame grids 
#' that fully enclose column headers with explicit bold top and bottom lines
#'
#' @param df dataframe of verification criteria
#' @param title character vector of title
#' @return A styled flextable object containing the merged two column layout with 
#' fully enclosed bold grid line formatting
#'
#' @importFrom tibble tibble
#' @importFrom dplyr mutate lag
#' @importFrom flextable flextable as_paragraph as_chunk align valign width border_outer border_inner_h border_inner_v bg color bold font set_caption delete_part add_footer_lines
#' @importFrom officer fp_border fp_text
#'
#' @export
#' @examples
#' \dontrun{gen_verification_criteria_table()}
gen_verification_criteria_table <- function(
    df = verification_criteria,
    title = "Vulnerability Criteria"
    ) {
  
  # track total row count
  total_rows <- nrow(df)
  
  # track the indices where a new criteria block begins for main horizontal lines
  main_grid_rows <- c(7, 12, 15, 16)
  
  # define the explicit border properties using officer
  heavy_border <- officer::fp_border(color = "#0B1054", width = 2.5, style = "solid")
  mid_border   <- officer::fp_border(color = "#0B1054", width = 2, style = "solid")
  thin_border  <- officer::fp_border(color = "#d3d3d3", width = 1, style = "solid")
  
  # build the flextable layout structure
  styled_table <- flextable::flextable(df) %>%
    # visually merge repeating cell items sequentially along the vertical axis
    flextable::merge_v(j = ~ Criteria) %>%
    # map alignments and set relative layout sizing to fit standard pages
    flextable::align(align = "left", part = "all") %>%
    flextable::valign(valign = "top", part = "body") %>%
    flextable::width(j = "Criteria", width = 2.2) %>%
    flextable::width(j = "Indicator", width = 4.3) %>%
    
    # apply deep blue background fill with white bold text for column labels
    flextable::bg(bg = "#0B1054", part = "header") %>%
    flextable::color(color = "white", part = "header") %>%
    flextable::bold(part = "header") %>%
    
    # apply a standard baseline horizontal grid line separating every single indicator row
    flextable::border_inner_h(border = thin_border, part = "body") %>%
    
    # overlay bold main horizontal block dividers within the body row structures
    flextable::hline(i = (main_grid_rows - 1), border = mid_border, part = "body") %>%
    
    # construct complete outer frame borders and center divider line enclosing headers and rows
    flextable::border_outer(border = heavy_border, part = "all") %>%
    flextable::border_inner_v(border = mid_border, part = "all") %>%
    
    # add font settings and map the source footnote notes cleanly to the bottom layout
    flextable::font(fontname = "Arial", part = "all") %>%
    flextable::add_footer_lines(
      values = "Acronym Definitions: OTP - Outpatient Therapeutic Program; SFP - Supplementary Feeding Program; FCS - Food Consumption Score; HHs - Households"
    ) %>%
    flextable::font(fontname = "Arial", part = "footer")
  
  return(styled_table)
}
```

## get vulnerability criteria variables

```{r function-get_vulnerability_data}

#' get vulnerability criteria (physiological, nutritional etc)
#'
#' @returns a dataframe
#' @importFrom dplyr filter mutate select lead
#' @export
#'
#' @examples
#' \dontrun{get_vulnerability_data()}
get_vulnerability_data <- function(){
  df_dict <- acheck::survey %>%
    filter(!grepl("note", type)) %>%
    mutate(n = dplyr::row_number())
  
  vul_rows <- df_dict %>% 
    # filter(grepl("^physio|^social|^nutri|^food|gps", label, ignore.case = T)) %>%
    filter(
      grepl(paste(paste0("^e_", 2:6), collapse = "|"), name)
    ) %>%
    bind_rows(
      
      vul_rows <- df_dict %>% 
        # filter(grepl("^physio|^social|^nutri|^food|gps", label, ignore.case = T)) %>%
        filter(
          grepl("^GPS", label)
        )
    ) %>%
    select(type, name, label, n) %>%
    mutate(n_end = lead(n),
           n = n + 1) %>%
    filter(!grepl("geo", type)) %>%
    # mutate(
    #   n_end = case_when(
    #     label == "Food consumption score" ~ 257,
    #     label == "Nutritional vulnerability" ~ 265,
    #     TRUE ~ n_end
    #   )
    # ) %>%
    filter(! n == n_end) %>%
    filter(
      grepl("^e", name)
    )
  vul_rows
}

```

## automate indicators

```{r function-automate_indicators}
#' Automate binary indicator variable creation
#'
#' This function dynamically scans input columns, handles missing data structures,
#' cleans raw count variables, and generates binary 0/1 indicators based on 
#' underlying numeric and character type properties, preserving the original name.
#'
#' @param df A dataframe or tibble object.
#' @param cols A character vector specifying the target column names to process.
#'
#' @return A modified dataframe with updated raw counts and appended indicator columns.
#' @importFrom dplyr mutate across all_of if_else
#' @importFrom stringr str_detect
#' @export
automate_indicators <- function(df, cols) {
  df %>%
    # first pass clean all raw numeric count inputs to handle nas and zeros universally
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(cols),
        function(x) {
          # lower case comments without dots or dashes
          if (is.numeric(x)) {
            return(dplyr::if_else(is.na(x) | x == 0, 0, x))
          }
          return(x)
        }
      )
    ) %>%
    # second pass generate indicators and append ind cleanly to the original column names
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(cols), 
        function(x) {
          if (is.numeric(x)) {
            indicator <- dplyr::if_else(x == 0, 0, 1)
            return(indicator)
          } 
          
          # explicitly separate affirmative responses from negative or uncertain ones
          clean_x <- tolower(x)
          indicator <- dplyr::if_else(
            is.na(x) | 
              clean_x == "no" | 
              stringr::str_detect(clean_x, "not_at_all|not_really|dont_know|don't know"), 
            0, 
            1
          )
          return(indicator)
        }, 
        .names = "{.col}_ind"
      )
    )
}
```

## donut plot

```{r function-plot_parts_donut}
#' visualize parts to whole using a donut layout framework
#'
#' @param data dataframe
#' @param fill_col character. the categorical column to slice
#' @param palette character vector. hex color values
#'
#' @returns a ggplot object
#' @importFrom dplyr count mutate across all_of
#' @importFrom ggplot2 ggplot aes geom_col geom_text scale_fill_manual theme_void theme coord_polar xlim position_stack
#' @importFrom scales percent
#' @export
#' @examples
#' \dontrun{plot_parts_donut()}
plot_parts_donut <- function(data, fill_col, palette = c("#81C784", "#FFB74D", "#E65100")) {
  
  # lower case comments without dots or dashes
  # map labels and compute dynamic cumulative coordinates for clean radial centering
  plot_df <- data %>%
    dplyr::count(dplyr::across(dplyr::all_of(fill_col))) %>%
    dplyr::mutate(
      pct = n / sum(n),
      pct_label = paste0(
        # .data[[fill_col]], 
        # "\n(", 
        scales::percent(pct, accuracy = 1)
        # ,")"
      )
    )
  
  ggplot2::ggplot(plot_df, ggplot2::aes(x = 3, y = pct, fill = .data[[fill_col]])) +
    # maintain stable donut ring layout thickness
    ggplot2::geom_col(width = 1, color = "white", size = 0.5) +
    # shift text position horizontally outward beyond the main donut frame
    ggplot2::geom_text(
      ggplot2::aes(x = 4.2, label = pct_label),
      position = ggplot2::position_stack(vjust = 0.5),
      color = "black",
      fontface = "bold",
      size = 3.5
    ) +
    ggplot2::scale_fill_manual(values = palette) +
    ggplot2::coord_polar(theta = "y") +
    # expand the coordinate map boundary to give labels complete breathing room
    ggplot2::xlim(c(1, 5.2)) +
    ggplot2::theme_void() +
    ggplot2::theme(
      legend.position = "bottom"
    )
}
```

## faceted donuts

```{r function-plot_faceted_donuts}
#' visualize parts to whole using an automated faceted donut grid via patchwork
#'
#' @param data dataframe
#' @param fill_col character the categorical variable to slice 
#' @param facet_col character the variable to split panels by 
#' @param palette character vector hex color values mapping to categories
#'
#' @returns a patchwork grid assembly of donut plots
#' @importFrom dplyr filter count mutate pull
#' @importFrom ggplot2 ggplot aes geom_col geom_text scale_fill_manual theme_void theme coord_polar xlim labs element_text position_stack
#' @importFrom patchwork wrap_plots plot_layout
#' @importFrom stringr str_to_title
#' @export
#' @examples
#' \dontrun{plot_faceted_donuts()}
plot_faceted_donuts <- function(data, 
                                fill_col, 
                                facet_col, 
                                palette = c("#388E3C", "#F57C00", "#D32F2F")) {
  
  # extract unique facet groups while ignoring missing rows
  facet_groups <- data %>%
    dplyr::filter(!is.na(.data[[facet_col]])) %>%
    dplyr::pull(.data[[facet_col]]) %>%
    unique() %>%
    sort()
  
  # dynamically determine optimal column layouts based on category lengths
  n_facets <- length(facet_groups)
  calculated_ncol <- if (n_facets <= 3) {
    n_facets
  } else if (n_facets == 4) {
    2
  } else {
    3
  }
  
  # generate individual donut objects iteratively
  plot_list <- lapply(facet_groups, function(group) {
    
    # filter down to single panel subset
    sub_df <- data %>% 
      dplyr::filter(.data[[facet_col]] == group, !is.na(.data[[fill_col]]))
    
    # safeguard against empty group metrics
    if (nrow(sub_df) == 0) return(NULL)
    
    plot_df <- sub_df %>%
      dplyr::count(dplyr::across(dplyr::all_of(fill_col))) %>%
      dplyr::mutate(
        pct = n / sum(n),
        pct_label = paste0(
          scales::percent(pct, accuracy = 1)
        )
      )
    
    ggplot2::ggplot(plot_df, ggplot2::aes(x = 3, y = pct, fill = .data[[fill_col]])) +
      # maintain donut thickness between inner and outer radius limits
      ggplot2::geom_col(width = 1, color = "white", size = 0.5) +
      # push text radius position out beyond the chart area boundary
      ggplot2::geom_text(
        ggplot2::aes(x = 4.2, label = pct_label),
        position = ggplot2::position_stack(vjust = 0.5),
        color = "black",
        fontface = "bold",
        size = 2.8
      ) +
      ggplot2::scale_fill_manual(values = palette) +
      ggplot2::coord_polar(theta = "y") +
      # expand structural coordinate canvas limits to prevent text clipping
      ggplot2::xlim(c(1, 5.2)) +
      ggplot2::labs(title = stringr::str_to_title(group)) +
      ggplot2::theme_void() +
      ggplot2::theme(
        # removed legend position here to let patchwork handle it
        plot.title = ggplot2::element_text(hjust = 0.5, face = "bold", size = 11, vjust = -1)
      )
  })
  
  # drop null items from tracking list
  # using base r filter option to keep namespaces minimal
  plot_list <- plot_list[!sapply(plot_list, is.null)]
  
  # stitch panels together and collect legends into a single unified guide
  combined_grid <- patchwork::wrap_plots(plot_list, ncol = calculated_ncol) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
  
  return(combined_grid)
}
```

## theme plots

```{r function-theme_plots}
#' publication ready theme wrapper using cowplot with gridlines
#'
#' @param base_size numeric base font size
#' @param base_family character font family
#'
#' @returns a ggplot2 theme object
#' @importFrom cowplot theme_minimal_grid
#' @importFrom ggplot2 theme element_text rel margin
#' @export
theme_plots <- function(base_size = 11, base_family = "sans") {
  
  # leverage cowplot minimal grid as the structural foundation
  cowplot::theme_minimal_grid(
    base_size = base_size,
    base_family = base_family,
    rel_title = 1.25
  ) +
    # fine tune alignment and margins
    ggplot2::theme(
      plot.title = ggplot2::element_text(
        face = "bold",
        hjust = 0.5,
        margin = ggplot2::margin(b = 10)
      ),
      plot.subtitle = ggplot2::element_text(
        hjust = 0.5,
        color = "grey30",
        face = "plain",
        margin = ggplot2::margin(b = 8)
      ),
      legend.position = "bottom"
    )
}
```

## save plots as .png

```{r function-save_plots_to_png}
#' save a list of active plots into a grid using the png device
#'
#' @param plot_list list of ggplot or grid objects
#' @param output_path character path where the combined grid will be saved
#' @param ncol integer number of columns in the grid
#' @param width integer width of output image in pixels
#' @param height integer height of output image in pixels
#' @param res integer nominal resolution in ppi
#'
#' @return invisible null saves file directly to disk
#' @importFrom grDevices png dev.off
#' @importFrom grid grid.newpage pushViewport viewport grid.layout
#' @export
save_plots_to_png <- function(plot_list, 
                                   output_path, 
                                   ncol = 2, 
                                   width = 1600, 
                                   height = 1200, 
                                   res = 144) {
  
  # calculate layout grid dimensions based on total plot count
  n_plots <- length(plot_list)
  nrow <- ceiling(n_plots / ncol)
  
  # open the native png graphics device to catch the upcoming draw calls
  grDevices::png(
    filename = output_path, 
    width = width, 
    height = height, 
    res = res
  )
  
  # initialize a fresh blank grid canvas page
  grid::grid.newpage()
  
  # push the base grid layout matrix structure onto the canvas
  grid::pushViewport(
    grid::viewport(layout = grid::grid.layout(nrow, ncol))
  )
  
  # loop through every plot object in the list and assign coordinates
  for (i in seq_along(plot_list)) {
    row_idx <- ceiling(i / ncol)
    col_idx <- (i - 1) %% ncol + 1
    
    # render the plot directly into its designated grid cell coordinates
    print(
      plot_list[[i]], 
      vp = grid::viewport(layout.pos.row = row_idx, layout.pos.col = col_idx)
    )
  }
  
  # shut down the active device to commit and write the file to disk
  grDevices::dev.off()
}
```


## plot household counts

```{r function-plot_household_counts}
#' plot household aggregation counts using unicode location pins with bold plotmath ggrepel label text callouts
#'
#' @param data dataframe containing household level records
#' @param lat_col character name of the latitude column
#' @param lon_col character name of the longitude column
#' @param group_by character administrative grouping scale choice, either 'ward' or 'county'
#' @param compass_pos character layout position for the north arrow, defaults to 'bl'
#' @param font_size numeric. font size for labels
#'
#' @returns a ggplot object
#' @importFrom dplyr filter group_by summarise n count left_join mutate
#' @importFrom sf st_as_sf st_join st_centroid st_coordinates st_drop_geometry
#' @importFrom geodata gadm
#' @importFrom ggspatial annotation_north_arrow north_arrow_minimal
#' @importFrom ggrepel geom_text_repel
#' @importFrom grid arrow unit
#' @importFrom ggplot2 ggplot geom_sf geom_text aes labs theme_minimal theme element_text coord_sf element_rect
#' @export
plot_household_counts <- function(data, lat_col, lon_col, group_by = "ward", compass_pos = "bl", font_size = 2.8) {
  
  # isolate coordinates and filter out missing records
  plot_df <- data %>%
    dplyr::filter(!is.na(.data[[lat_col]]), !is.na(.data[[lon_col]]))
  
  # transform the raw dataframe into a coordinate spatial sf object (wgs84 projection)
  sf_df <- sf::st_as_sf(
    plot_df, 
    coords = c(lon_col, lat_col), 
    crs = 4326
  )
  
  # dynamically download the official boundaries from gadm database
  kenya_spatvector_l1 <- geodata::gadm(country = "KEN", level = 1, path = base::tempdir())
  kenya_counties_sf <- sf::st_as_sf(kenya_spatvector_l1)
  
  kenya_spatvector_l3 <- geodata::gadm(country = "KEN", level = 3, path = base::tempdir())
  kenya_wards_sf <- sf::st_as_sf(kenya_spatvector_l3)
  
  # aggregate counts and build mathematical bold display labels based on chosen administrative group scale
  if (group_by == "ward") {
    # spatially join household coordinates to identify active ward parents
    sf_joined <- suppressMessages(sf::st_join(sf_df, kenya_wards_sf, join = sf::st_intersects))
    
    # generate frequencies matching unique gadm layer identifiers and drop point geometry structures safely
    counts_df <- sf_joined %>%
      sf::st_drop_geometry() %>%
      dplyr::filter(!is.na(GID_3)) %>%
      dplyr::count(GID_3, name = "hh_count")
    
    # bind counts back onto geometry frame and format text explicitly inside plotmath expressions
    annotated_sf <- kenya_wards_sf %>%
      dplyr::left_join(counts_df, by = "GID_3") %>%
      dplyr::filter(!is.na(hh_count)) %>%
      dplyr::mutate(display_label = base::paste0("atop(bold('", NAME_3, "'), '(N = ", hh_count, ")')"))
    
  } else if (group_by == "county") {
    # spatially join household coordinates to identify active county parents
    sf_joined <- suppressMessages(sf::st_join(sf_df, kenya_counties_sf, join = sf::st_intersects))
    
    # generate frequencies matching unique gadm layer identifiers and drop point geometry structures safely
    counts_df <- sf_joined %>%
      sf::st_drop_geometry() %>%
      dplyr::filter(!is.na(GID_1)) %>%
      dplyr::count(GID_1, name = "hh_count")
    
    # bind counts back onto geometry frame and format text explicitly inside plotmath expressions
    annotated_sf <- kenya_counties_sf %>%
      dplyr::left_join(counts_df, by = "GID_1") %>%
      dplyr::filter(!is.na(hh_count)) %>%
      dplyr::mutate(display_label = base::paste0("atop(bold('", NAME_1, "'), '(N = ", hh_count, ")')"))
  } else {
    base::stop("invalid group_by selection, please configure as either 'ward' or 'county'")
  }
  
  # extract geographic interior centroids calculated from the largest continuous landmass polygon space
  label_points <- base::suppressWarnings(sf::st_centroid(annotated_sf, of_largest_polygon = TRUE))
  label_coords <- sf::st_coordinates(label_points)
  
  annotated_sf$label_lon <- label_coords[, 1]
  annotated_sf$label_lat <- label_coords[, 2]
  
  # map the underlying geographic map framework and lay labels over top
  p <- ggplot2::ggplot() +
    # render detailed ward lines using thin soft pink profiles
    ggplot2::geom_sf(
      data = kenya_wards_sf,
      fill = "#fff5f6",
      color = "#FFE4E8",
      linewidth = 0.3,
      inherit.aes = FALSE
    ) +
    # overlay the main county line borders using a thicker soft pink profile
    ggplot2::geom_sf(
      data = kenya_counties_sf,
      fill = NA,
      color = "#FFB7C5",
      linewidth = 1.0,
      inherit.aes = FALSE
    ) +
    # drop authentic unicode location pin symbols directly locked onto the precise internal centroids
    ggplot2::geom_text(
      data = annotated_sf,
      ggplot2::aes(x = label_lon, y = label_lat),
      label = "\U0001F4CD",
      size = 5,
      vjust = 0.5
    ) +
    # repel text descriptions outside regional boundaries using mathematical parsing to force bold styling
    ggrepel::geom_text_repel(
      data = annotated_sf,
      ggplot2::aes(x = label_lon, y = label_lat, label = display_label),
      color = "#722F37",
      size = font_size,
      parse = TRUE,
      box.padding = 1.2,
      point.padding = 0.3,
      max.overlaps = Inf,
      arrow = grid::arrow(length = grid::unit(0.015, "npc"), type = "closed"),
      segment.color = "#722F37",
      segment.size = 0.4
    )
  
  # format layout grids and text descriptions
  p <- p +
    # enforce true spatial map coordinates and automatically scale map frame bounds to whole country shape
    ggplot2::coord_sf(crs = 4326) +
    # inject a minimal clean compass style north arrow matching the soft pink layout color at the custom location
    ggspatial::annotation_north_arrow(
      location = compass_pos, 
      which_north = "true",
      style = ggspatial::north_arrow_minimal(line_col = "#FFB7C5", fill = "#FFB7C5")
    ) +
    ggplot2::labs(
      x = "Longitude",
      y = "Latitude"
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", hjust = 0.5, size = 13),
      legend.position = "bottom",
      legend.box = "vertical",
      # clean background panels
      panel.background = ggplot2::element_rect(fill = "#fff9fa", color = NA)
    )
  
  return(p)
}
```



## household count plot interactive

```{r function-plot_household_counts_interactive}
#' plot interactive household count markers using real location pin icons positioned safely inside boundaries
#'
#' @param data dataframe containing household level records
#' @param lat_col character name of the latitude column
#' @param lon_col character name of the longitude column
#' @param group_by character administrative grouping scale choice, either 'ward' or 'county'
#' @param compass_pos character layout position for the north arrow, defaults to 'topleft'
#'
#' @returns a leaflet map widget object
#' @importFrom dplyr filter group_by summarise n count left_join mutate
#' @importFrom sf st_as_sf st_join st_point_on_surface st_coordinates st_drop_geometry
#' @importFrom geodata gadm
#' @importFrom leaflet leaflet addPolygons addMarkers labelOptions leafletOptions addControl
#' @importFrom htmltools HTML
#' @export
plot_household_counts_interactive <- function(data, lat_col, lon_col, group_by = "ward", compass_pos = "topleft") {
  
  # isolate coordinates and filter out missing records
  plot_df <- data %>%
    dplyr::filter(!is.na(.data[[lat_col]]), !is.na(.data[[lon_col]]))
  
  # transform the raw dataframe into a coordinate spatial sf object (wgs84 projection)
  sf_df <- sf::st_as_sf(
    plot_df, 
    coords = c(lon_col, lat_col), 
    crs = 4326
  )
  
  # dynamically download the official boundaries from gadm database
  kenya_spatvector_l1 <- geodata::gadm(country = "KEN", level = 1, path = base::tempdir())
  kenya_counties_sf <- sf::st_as_sf(kenya_spatvector_l1)
  
  kenya_spatvector_l3 <- geodata::gadm(country = "KEN", level = 3, path = base::tempdir())
  kenya_wards_sf <- sf::st_as_sf(kenya_spatvector_l3)
  
  # aggregate counts and map tracking labels based on chosen administrative group scale
  if (group_by == "ward") {
    # spatially join household coordinates to identify active ward parents
    sf_joined <- base::suppressMessages(sf::st_join(sf_df, kenya_wards_sf, join = sf::st_intersects))
    
    # generate frequencies matching unique gadm layer identifiers and drop point geometry structures safely
    counts_df <- sf_joined %>%
      sf::st_drop_geometry() %>%
      dplyr::filter(!is.na(GID_3)) %>%
      dplyr::count(GID_3, name = "hh_count")
    
    # bind counts back onto geometry frame and format clean text combinations for interactive hover flags
    annotated_sf <- kenya_wards_sf %>%
      dplyr::left_join(counts_df, by = "GID_3") %>%
      dplyr::filter(!is.na(hh_count)) %>%
      dplyr::mutate(
        hover_label = base::paste0("<strong>Ward:</strong> ", NAME_3, "<br><strong>Households:</strong> ", hh_count)
      )
    
  } else if (group_by == "county") {
    # spatially join household coordinates to identify active county parents
    sf_joined <- base::suppressMessages(sf::st_join(sf_df, kenya_counties_sf, join = sf::st_intersects))
    
    # generate frequencies matching unique gadm layer identifiers and drop point geometry structures safely
    counts_df <- sf_joined %>%
      sf::st_drop_geometry() %>%
      dplyr::filter(!is.na(GID_1)) %>%
      dplyr::count(GID_1, name = "hh_count")
    
    # bind counts back onto geometry frame and format clean text combinations for interactive hover flags
    annotated_sf <- kenya_counties_sf %>%
      dplyr::left_join(counts_df, by = "GID_1") %>%
      dplyr::filter(!is.na(hh_count)) %>%
      dplyr::mutate(
        hover_label = base::paste0("<strong>County:</strong> ", NAME_1, "<br><strong>Households:</strong> ", hh_count)
      )
  } else {
    base::stop("invalid group_by selection, please configure as either 'ward' or 'county'")
  }
  
  # extract spatial surface coordinates guaranteed to be mathematically inside the polygon boundaries
  label_points <- base::suppressWarnings(sf::st_point_on_surface(annotated_sf))
  label_coords <- sf::st_coordinates(label_points)
  
  annotated_sf$label_lon <- label_coords[, 1]
  annotated_sf$label_lat <- label_coords[, 2]
  
  # design a custom html soft pink compass ornament using a safe portable unicode escape sequence for the arrow (\u25b2)
  compass_html <- htmltools::HTML('
    <div style="
      text-align: center; 
      font-family: sans-serif; 
      color: #FFB7C5; 
      background: rgba(255, 245, 246, 0.8); 
      padding: 6px 10px; 
      border-radius: 4px; 
      border: 2px solid #FFB7C5;
      font-weight: bold;
    ">
      <div style="font-size: 20px; line-height: 1;">\u25B2</div>
      <div style="font-size: 13px; margin-top: 2px;">N</div>
    </div>
  ')
  
  # initialize a clean leaflet map structure with zooming rules
  lf <- leaflet::leaflet(options = leaflet::leafletOptions(minZoom = 5, maxZoom = 14)) %>%
    # render detailed background ward lines using a very clean subtle light pink fill
    leaflet::addPolygons(
      data = kenya_wards_sf,
      fillColor = "#fffafb",
      fillOpacity = 0.6,
      color = "#FFE4E8",
      weight = 1,
      opacity = 0.8,
      group = "Wards"
    ) %>%
    # overlay the main county line borders using a thicker soft pink profile
    leaflet::addPolygons(
      data = kenya_counties_sf,
      fillColor = "transparent",
      color = "#FFB7C5",
      weight = 1.8,
      opacity = 1.0,
      group = "Counties"
    ) %>%
    # anchor the custom styled compass ornament securely to the canvas UI
    leaflet::addControl(html = compass_html, position = compass_pos) %>%
    # render native location pin icon markers that surface rich html label strings upon mouse hover
    leaflet::addMarkers(
      data = annotated_sf,
      lng = ~label_lon,
      lat = ~label_lat,
      label = ~base::lapply(hover_label, htmltools::HTML),
      labelOptions = leaflet::labelOptions(
        style = base::list(
          "font-family" = "sans-serif",
          "font-size" = "12px",
          "color" = "#722F37",
          "border-color" = "#FFB7C5"
        )
      )
    )
  
  return(lf)
}
```

## get partner beneficiaries

```{r function-get_partner_beneficiaries}
#' filter and deduplicate beneficiary logs by agency partner
#'
#' @param df dataframe containing master registry logs
#' @param partner character keyword used to identify the target agency, defaults to 'pgi'
#' @param columns character vector of column names to distinct by, defaults to c('ben_name', 'ben_id_number')
#'
#' @returns a deduplicated dataframe containing partner specific records
#' @importFrom dplyr filter distinct across all_of
#' @export
get_partner_beneficiaries <- function(df = df_out, partner = "pgi", columns = c("ben_name", "ben_id_number")) {
  
  # match agency keyword and remove duplicates based on dynamic columns argument configuration
  processed_df <- df %>%
    dplyr::filter(base::grepl(partner, enum_egency, ignore.case = TRUE)) %>%
    dplyr::distinct(dplyr::across(dplyr::all_of(columns)), .keep_all = TRUE)
  
  return(processed_df)
}
```



```{r function-check_disability}
#' validate and recode household disability flags based on repeat group ages
#'
#' @param df_out dataframe containing the main household data
#' @param df_disab dataframe containing the individual repeat group data
#' @param uuid_col character string of the ID column name in df_out (defaults to "uuid")
#' @param age numeric lower age limit
#'
#' @returns an updated version of df_out with corrected disability flags
#'
#' @export
#'
#' @examples
#' # lower case comments without dots or dashes
#' # create dummy main dataset
#' main_df <- data.frame(
#'   uuid = c("hh_1", "hh_2", "hh_3"),
#'   HHM_disability = c("yes", "yes", "no"),
#'   stringsAsFactors = FALSE
#' )
#' 
#' # create dummy repeat group dataset
#' repeat_df <- data.frame(
#'   `_uuid` = c("hh_1", "hh_1", "hh_2", "hh_2", "hh_3"),
#'   HHM_disability = c("yes", "no", "yes", "yes", "no"),
#'   prot_disability_ind_age = c(1, 35, 2, 1, 40),
#'   stringsAsFactors = FALSE,
#'   check.names = FALSE
#' )
#' 
#' # run the function
#' output_df <- check_disability(main_df, repeat_df, uuid_col = "uuid")
check_disability <- function(df_out, df_disab, uuid_col = "uuid", age = 6) {
  
  # step 1: isolate active disability rows and find the maximum age per household
  hh_max_disab_age <- df_disab %>%
    dplyr::filter(stringr::str_detect(base::tolower(HHM_disability), "yes")) %>%
    dplyr::group_by(`_uuid`) %>%
    dplyr::summarise(
      max_age = base::max(prot_disability_ind_age, na.rm = TRUE),
      .groups = "drop"
    )
  
  # step 2: extract household IDs where the maximum disabled age is 2 or under
  infant_only_uuids <- hh_max_disab_age %>%
    dplyr::filter(max_age < age) %>%
    dplyr::pull(`_uuid`)
  
  # step 3: calculate diagnostics for console visibility
  total_infant_only_hh <- base::length(infant_only_uuids)
  
  # step 4: recode the main dataset if any invalid target households exist
  df_out_updated <- df_out %>%
    dplyr::mutate(
      # lower case comments without dots or dashes
      HHM_disability = dplyr::if_else(
        .data[[uuid_col]] %in% infant_only_uuids,
        "no",
        HHM_disability
      )
    )
  
  # step 5: print diagnostic feedback to the console safely
  base::message("--- disability check summary ---")
  base::message(base::sprintf("households with active disability loops: %d", base::nrow(hh_max_disab_age)))
  base::message(base::sprintf("households where all disabled members are < %d (total: %d)", age, total_infant_only_hh))
  
  if (total_infant_only_hh > 0) {
    base::message("action: invalid household flags successfully recoded to 'no'")
  } else {
    base::message("action: no adjustments needed, all household flags are structurally valid")
  }
  base::message("--------------------------------")
  
  return(df_out_updated)
}
```

## calculate disability counts

```{r function-calculate_disability_counts}
#' calculate a row level individual disability indicator
#'
#' @param data dataframe containing the individual disability columns
#' @param disability_cols character vector of column names representing the domains
#'
#' @returns an integer vector of 1s and 0s indicating if the individual has any disability
#' @export
calculate_disability_counts <- function(data, 
                                        disability_cols = c(
                                          "level_seeing_disability",
                                          "level_hearing_disability",
                                          "level_selfcare_disability",
                                          "level_remembering_disability",
                                          "level_walking_disability",
                                          "level_communicating_disability"
                                        )) {
  
  # extract target columns to safely check row conditions
  subset_matrix <- data %>% 
    dplyr::select(dplyr::any_of(disability_cols))
  
  # match case-insensitively against the exact literal strings provided
  # lower case comments without dots
  match_matrix <- base::apply(subset_matrix, 2, function(col) {
    stringr::str_to_lower(base::as.character(col)) %in% c("some difficulty", "a lot of difficulty", "cannot do at all")
  })
  
  # collapse individual domains down to a single binary indicator per row
  dplyr::if_else(base::rowSums(match_matrix, na.rm = TRUE) > 0, 1L, 0L)
}

```

## hh disability burden  counts

counts multiple forms of disability among household members.

```{r function-calculate_hh_disability_counts}
#' calculate the total number of disabled individuals at the household level
#'
#' @param data dataframe containing the individual disability columns and a household identifier
#' @param hh_id_col character name of the column storing the unique household id
#' @param disability_cols character vector of column names representing the domains
#'
#' @returns an integer vector representing the total disabled household members for each row
#' @export
calculate_hh_disability_counts <- function(data, 
                                           hh_id_col = "uuid",
                                           disability_cols = c(
                                             "level_seeing_disability",
                                             "level_hearing_disability",
                                             "level_selfcare_disability",
                                             "level_remembering_disability",
                                             "level_walking_disability",
                                             "level_communicating_disability"
                                           )) {
  
  # extract target columns to safely check row conditions
  subset_matrix <- data %>% 
    dplyr::select(dplyr::any_of(disability_cols))
  
  # match case-insensitively against the exact literal choices provided
  # lower case comments without dots
  match_matrix <- base::apply(subset_matrix, 2, function(col) {
    stringr::str_to_lower(base::as.character(col)) %in% c("some difficulty", "a lot of difficulty", "cannot do at all")
  })
  
  # collapse individual domains down to a single binary indicator per row
  ind_disabled <- dplyr::if_else(base::rowSums(match_matrix, na.rm = TRUE) > 0, 1L, 0L)
  
  # create a temporary tracking block to calculate the broadcasted household sums safely
  result_vector <- data %>%
    dplyr::mutate(temp_ind_disabled = ind_disabled) %>%
    dplyr::group_by(.data[[hh_id_col]]) %>%
    dplyr::mutate(total_disabled_in_hh = base::sum(temp_ind_disabled, na.rm = TRUE)) %>%
    dplyr::ungroup() %>%
    dplyr::pull(total_disabled_in_hh)
  
  return(result_vector)
}
```


## recode early childhood disability

```{r function-recode_early_childhood_disability}
#' reset disability columns based on age thresholds
#'
#' @param data dataframe containing the disability and age columns
#' @param disability_cols character vector of column names representing the domains
#' @param age_col character name of the column storing age values
#'
#' @returns a modified dataframe with recoded disability entries for children under 6
#' @export
recode_early_childhood_disability <- function(data,
                                              disability_cols = c(
                                                "level_seeing_disability",
                                                "level_hearing_disability",
                                                "level_selfcare_disability",
                                                "level_remembering_disability",
                                                "level_walking_disability",
                                                "level_communicating_disability"
                                              ),
                                              age_col = "prot_disability_ind_age") {
  data %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(disability_cols),
        ~ dplyr::if_else(
          # check if age exists and falls below the 6 year target threshold
          !base::is.na(.data[[age_col]]) & .data[[age_col]] < 6,
          "No difficulty",
          # preserve the original cell contents if criteria are not met
          base::as.character(.x)
        )
      )
    )
}
```


## recode factor to character

```{r function-recode_factor_to_character}
#' recode factors to character
#'
#' @param x dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{recode_factor_to_character()}
recode_factor_to_character <- function(x) {
  x %>%
  dplyr::mutate(
    dplyr::across(
      dplyr::where(labelled::is.labelled),
      # convert labels to factors, drop unused levels, and cast to character strings
      ~ .x %>%
        as_factor() %>%
        forcats::fct_drop() %>%
        base::as.character()
    )
  )
  
}
```



## get dupes with name

```{r function-get_dupes_with_name}
#' get duplicate rows, track the checked field, and subset down to designated columns via an argument
#'
#' @param data a dataframe or tibble
#' @param col the unquoted column name to check for duplicates, defaults to ben_name
#' @param cols_select a list of unquoted column expressions to select, defaults to a predefined set
#'
#' @returns a filtered and pruned dataframe of duplicate rows
#' @importFrom rlang ensyms exprs
#' @importFrom janitor get_dupes
#' @importFrom dplyr mutate select
#' @export
get_dupes_with_name <- function(
    data, 
    col = ben_name, 
    cols_select = rlang::exprs(county, enum_egency, enum_name,ben_name, ben_id_number,reg_num, dup_col, uuid)
) {
  
  # capture the checking columns safely as a list of symbols
  col_syms <- rlang::ensyms(col)
  
  # convert the captured symbols into character strings for label grouping
  col_strings <- base::as.character(col_syms)
  combined_label <- base::paste(col_strings, collapse = ", ")
  
  # run the janitor deduplication filter across all target symbols and append the tracking string
  processed_df <- data %>%
    janitor::get_dupes(!!!col_syms) %>%
    dplyr::mutate(dup_col = combined_label)
  
  # inject and combine the checked column symbols directly into the selection expression list
  final_select_exprs <- base::unique(c(col_syms, cols_select))
  
  # evaluate the complete unified selection expression list dynamically
  result <- processed_df %>%
    dplyr::select(!!!final_select_exprs)
  
  return(result)
}
```

## label data 

```{r function-label_kobo_columns}
#' replace data column names with question labels from a kobo asset form metadata structure
#'
#' @param data dataframe containing the downloaded survey submission records
#' @param kobo_asset a labelled list/dataframe compiled from the xlsform or robotoolbox
#' @param lang character optional specific language label column to use if multilingual, e.g., 'English (en)'
#'
#' @returns dataframe with renamed columns
#' @importFrom purrr map_chr
#' @importFrom dplyr rename_with everything
#' @export
label_kobo_columns <- function(data, kobo_asset, lang = NULL) {
  
  # extract the label attributes from the kobo asset columns
  label_mapping <- purrr::map_chr(kobo_asset, function(col) {
    lbls <- attr(col, "label")
    if (is.null(lbls)) return(NA_character_)
    
    # handle multilingual assets where labels might be a named vector
    if (!is.null(lang) && lang %in% names(lbls)) {
      return(lbls[lang])
    }
    
    # default to the first available label if no language is specified
    return(lbls[1])
  })
  
  # drop missing or empty labels to keep original column names where mapping fails
  label_mapping <- label_mapping[!is.na(label_mapping) & label_mapping != ""]
  
  # rename columns using the named vector mapping lookup
  result <- data %>%
    dplyr::rename_with(
      ~ ifelse(.x %in% names(label_mapping), label_mapping[.x], .x),
      dplyr::everything()
    )
  
  return(result)
}
```

## read xls form as kobo labelled assset

```{r function-read_xlsform_asset}
#' read an xlsform file and compile it into a kobo asset metadata list
#'
#' @param xlsform_path character. path to your local xlsform .xlsx or .xls file
#'
#' @returns a named list structured exactly like a robotoolbox kobo_asset object
#' @importFrom readxl read_excel
#' @importFrom dplyr filter select
#' @export
read_xlsform_asset <- function(xlsform_path) {
  
  # 1. read the survey design sheet from the workbook
  survey_sheet <- readxl::read_excel(xlsform_path, sheet = "survey")
  
  # 2. filter down to rows that represent real questions with names
  survey_questions <- survey_sheet %>%
    dplyr::filter(!is.na(name))
  
  # 3. isolate all available label columns (handles multilingual headers like label::English (en))
  label_cols <- names(survey_questions)[stringr::str_starts(names(survey_questions), "label")]
  
  # if no complex language markers exist, default back to standard label
  if (length(label_cols) == 0 && "label" %in% names(survey_questions)) {
    label_cols <- "label"
  }
  
  # 4. dynamically build the mocked asset list structure
  kobo_asset <- list()
  
  for (i in seq_len(nrow(survey_questions))) {
    col_name <- survey_questions$name[i]
    
    # gather all label strings for the current question row
    labels_vec <- c()
    for (lbl_col in label_cols) {
      lbl_val <- survey_questions[[lbl_col]][i]
      
      if (!is.na(lbl_val) && lbl_val != "") {
        # normalize name format: strip 'label::' prefix if present to isolate language key
        lang_key <- stringr::str_remove(lbl_col, "^label::")
        if (lang_key == "label") lang_key <- "default"
        
        labels_vec[lang_key] <- lbl_val
      }
    }
    
    # structure each element as an empty mock column container holding label metadata
    col_trigger <- list()
    attr(col_trigger, "label") <- labels_vec
    
    kobo_asset[[col_name]] <- col_trigger
  }
  
  return(kobo_asset)
}
```


## convert kobo header labels to column values

```{r function-convert_excel_headers}
#' convert labeled excel headers to kobo code/value headers using form schema
#'
#' @param target_df data.frame. the raw excel data frame with text headers (df)
#' @param project_id character. kobo project uid
#' @param url character. Kobotoolbox url
#' @param token character. project token
#'
#' @returns a data frame with clean value headers matching df_lab
#' @export
convert_excel_headers <- function(
    target_df,
    url = Sys.getenv("kobo_url2"),
    project_id = Sys.getenv("mpca_id"),
    token = Sys.getenv("kobo_token")
) {
  
  # 1. pull the asset form schema from the api
  robotoolbox::kobo_setup(url = url, token = token)
  asset <- robotoolbox::kobo_asset(project_id)
  form_schema <- robotoolbox::kobo_form(asset)
  
  # 2. isolate question row rows with non-empty names and labels
  # clean the labels exactly how R cleans column names upon import
  mapping_dict <- form_schema %>%
    dplyr::filter(!is.na(name), !is.na(label)) %>%
    dplyr::mutate(
      clean_label = janitor::make_clean_names(label)
    ) %>%
    dplyr::select(clean_label, name) %>%
    dplyr::distinct(clean_label, .keep_all = TRUE) %>%
    tibble::deframe() # turns it into a named vector: c(clean_label = "name")
  
  # 3. safely handle complex select_multiple columns
  # kobo appends choice suffixes to the question stem for binary choice columns
  # e.g., 'what_is_the_current_source_of_livelihood..._formal_employment' -> 'livelihood_source_f_employment'
  current_names <- names(target_df)
  new_names <- sapply(current_names, function(col_name) {
    
    # direct hit check against standard questions
    if (col_name %in% names(mapping_dict)) {
      return(mapping_dict[[col_name]])
    }
    
    # check if this column is a split choice from a select_multiple question block
    # matches the longest matching stem and finds the underlying code value
    matches <- sapply(names(mapping_dict), function(lbl) stringr::str_starts(col_name, fixed(lbl)))
    if (any(matches)) {
      best_fit_lbl <- names(which(matches))[which.max(nchar(names(which(matches))))]
      base_code_name <- mapping_dict[[best_fit_lbl]]
      
      # extract just the choice suffix appended to the end of the text label
      suffix <- stringr::str_remove(col_name, paste0("^", regexp_quote(best_fit_lbl), "_?"))
      
      # reconstruct the backend column name string structure
      return(paste0(base_code_name, "_", suffix))
    }
    
    # return original name if no match found (e.g. metadata markers like uuid, start, end)
    return(col_name)
  })
  
  # 4. assign the translated backend names onto your excel dataset
  names(target_df) <- unname(new_names)
  
  # 5. a quick universal adjustment pass for project-agnostic group paths
  # fixes remnants of section wrappers if your schema lists them cleanly
  target_df <- target_df %>%
    dplyr::rename_with(~ stringr::str_replace_all(.x, "\\.", "_"))
  
  return(target_df)
}

# helper function to escape characters for safe regex evaluation
regexp_quote <- function(kv) {
  gsub("([.[{}()^*+?|\\^$])", "\\\\\\1", kv)
}
```

## label with xlsform

```{r function-label_with_xlsform}
#' convert and label an excel export using xlsform workbook sheets
#'
#' @param data_df data.frame. the raw exported kobo data with label headers (df)
#' @param xlsform_path character. path to your local xlsform .xlsx workbook
#'
#' @returns a clean data frame with code headers and labeled categorical values
#' @export
label_with_xlsform <- function(data_df, xlsform_path) {
  
  # 1. read the design sheets from the xlsform template
  survey_sheet  <- readxl::read_excel(xlsform_path, sheet = "survey")
  choices_sheet <- readxl::read_excel(xlsform_path, sheet = "choices")
  
  # 2. handle column header conversion (label string -> code name)
  # ensure we only target actual question rows with values
  survey_clean <- survey_sheet %>%
    dplyr::filter(!is.na(name), !is.na(label)) %>%
    dplyr::mutate(clean_label = janitor::make_clean_names(label))
  
  # generate a named mapping dictionary vector: c(clean_label = "name")
  header_dict <- survey_clean %>%
    dplyr::select(clean_label, name) %>%
    dplyr::distinct(clean_label, .keep_all = TRUE) %>%
    tibble::deframe()
  
  current_names <- names(data_df)
  new_names <- sapply(current_names, function(col_name) {
    # check for a direct match
    if (col_name %in% names(header_dict)) {
      return(header_dict[[col_name]])
    }
    
    # check for select_multiple split columns
    # loops through labels to find if col_name starts with a known stem label
    matches <- sapply(names(header_dict), function(lbl) stringr::str_starts(col_name, fixed(lbl)))
    if (any(matches)) {
      best_fit_lbl <- names(which(matches))[which.max(nchar(names(which(matches))))]
      base_code_name <- header_dict[[best_fit_lbl]]
      
      # isolate the text choice suffix and append it back to the clean code stem
      suffix <- stringr::str_remove(col_name, paste0("^", gsub("([.[{}()^*+?|\\^$])", "\\\\\\1", best_fit_lbl), "_?"))
      return(paste0(base_code_name, "_", suffix))
    }
    return(col_name)
  })
  
  names(data_df) <- unname(new_names)
  
  # 3. handle categorical row value labeling
  # extract the choice list links from the question types (e.g., 'select_one list_name')
  questions_with_choices <- survey_clean %>%
    dplyr::filter(stringr::str_detect(type, "select_")) %>%
    dplyr::mutate(
      list_name = stringr::str_extract(type, "(?<=select_[a-zA-Z0-9_]{1,100}[:space:]).*")
    ) %>%
    dplyr::filter(!is.na(list_name)) %>%
    dplyr::select(name, list_name)
  
  # build dictionaries for each choice group inside the choices sheet
  choices_clean <- choices_sheet %>%
    dplyr::filter(!is.na(list_name), !is.na(name), !is.na(label)) %>%
    dplyr::mutate(
      name = as.character(name),
      lookup_key = paste0(list_name, "__", name)
    )
  
  choices_lookup <- choices_clean %>%
    dplyr::select(lookup_key, label) %>%
    dplyr::distinct(lookup_key, .keep_all = TRUE) %>%
    tibble::deframe()
  
  # create a question-to-list mapping vector
  q_to_list <- questions_with_choices %>% tibble::deframe()
  
  # 4. map row values across matching select columns
  data_df <- data_df %>%
    dplyr::mutate(dplyr::across(dplyr::everything(), function(col) {
      col_name <- dplyr::cur_column()
      
      if (col_name %in% names(q_to_list)) {
        list_name <- q_to_list[[col_name]]
        col_str <- as.character(col)
        
        # construct lookup codes: listname__value
        lookup_keys <- paste0(list_name, "__", col_str)
        
        col_mapped <- dplyr::if_else(
          lookup_keys %in% names(choices_lookup),
          choices_lookup[lookup_keys],
          col_str
        )
        return(col_mapped)
      }
      return(col)
    }))
  
  # universal pass to remove group dot leftovers if present
  data_df <- data_df %>%
    dplyr::rename_with(~ stringr::str_replace_all(.x, "\\.", "_"))
  
  return(data_df)
}
```


## parse hh size

```{r function-parse_hhsize}
#' Parse and extract household size as an integer vector
#'
#' This function sanitizes a character vector of household sizes containing
#' mixed entries (e.g., "5(3F:2M)", "Size 9", "s6"), discarding data entry 
#' noise like phone numbers, registration IDs, or full names.
#'
#' @param x character. A vector containing raw household size values.
#'
#' @returns An integer vector of clean household sizes.
#' @export
#'
#' @examples
#' \dontrun{
#' clean_sizes <- parse_hhsize(df$ben_hhsize)
#' }
parse_hhsize <- function(x) {
  
  # convert vector to character and make uniform for regex matching
  working_vec <- stringr::str_trim(as.character(x))
  working_vec <- stringr::str_to_lower(working_vec)
  
  # perform structural parsing based on prioritized text patterns
  clean_vec <- dplyr::case_when(
    # extract leading total digits before bracketed breakdowns: '5(3f:2m)' -> 5
    stringr::str_detect(working_vec, "^\\d+\\s*\\(") ~ 
      stringr::str_extract(working_vec, "^\\d+"),
    
    # isolate strings with descriptive prefix layout: 'size 9' or 'fs- 5' -> 9 or 5
    stringr::str_detect(working_vec, "^(size|fs|fs-|s)\\s*\\d+$") ~ 
      stringr::str_extract(working_vec, "\\d+"),
    
    # isolate strings with standalone gender suffix layout: '3m' or '6f' -> 3 or 6
    stringr::str_detect(working_vec, "^\\d+[mf]$") ~ 
      stringr::str_extract(working_vec, "^\\d+"),
    
    # safely convert known text noise, registration IDs, and phone numbers to NA
    stringr::str_detect(working_vec, "-|\\s[a-z]|^07") ~ NA_character_,
    
    # keep remaining clean structures intact
    TRUE ~ working_vec
  )
  
  # return final casted integer vector
  return(as.integer(clean_vec))
}
```


## run sampling schemes

```{r function-run_sample}
#' run sampling schemes
#'
#' @param df_frame dataframe of sampling frame
#' @param df_sample_n dataframe of sample size
#' @param dir charcter. of output path
#' @param type charcter. monitoring stage 
#' @param name charcter. file name for otput
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{run_sample()}
run_sample <- function(df_frame, df_sample_n, dir = Sys.getenv("dir_unocha"), type = "PDM", name = "FDG"){
  large_villages <- df_frame %>%
    summarise(
      n = n(),
      .by = c(county, village)
    ) %>%
    filter(n >= 50)
  
  ds <- df_frame %>%
    dplyr::filter(grepl("^vuln", eligibility)) %>%
    filter(
      village %in% large_villages$village
    )
  
  dd <- draw_pps_sample(
    data = ds,
    allocation_df = df_sample_n,
    
  )
  
  cols <- c("county",
            "subloc",
            "ward",
            "village$",
            "ben_name",
            "ben_id_number",
            "reg_num$",
            "proxy_first",
            "proxy_second",
            "proxy_num"
            
  )
  sample_out <- dd %>%
    select(grep(paste(cols, collapse = "|"), names(.)))
  out_lst <- split(sample_out, sample_out$county) %>%
    purrr::set_names(sort(unique(sample_out$county)))
  
  acheck::write_list_to_excel(
    out_lst,
    file_path = file.path(dir, type, "sample", paste0(name, "_sample_", Sys.Date(), ".xlsx"))
  )
}

```

<!--
There can be other functions, examples and tests in your flat template.
Each of them will be inflated in a different file, provided that there is a level-1 or level-2 section title to separate from previous functions.
-->



<!--
# There can be development actions

Create a chunk with 'development' actions

- The chunk needs to be named `development` or `dev`
- It contains functions that are used for package development only
- Note that you may want to store most of these functions in the 0-dev_history.Rmd file

These are only included in the present flat template file, their content will not be part of the package anywhere else.
-->

```{r development-inflate, eval=FALSE}
# Keep eval=FALSE to avoid infinite loop in case you hit the knit button
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_full.Rmd", vignette_name = "Get started")
```


# Inflate your package

You're one inflate from paper to box.
Build your package from this very Rmd using `fusen::inflate()`

- Verify your `"DESCRIPTION"` file has been updated
- Verify your function is in `"R/"` directory
- Verify your test is in `"tests/testthat/"` directory
- Verify this Rmd appears in `"vignettes/"` directory
