---
title: "Data checks"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- Run this 'development' chunk -->
<!-- Store every call to library() that you need to explore your functions -->

```{r development, include=FALSE}
library(testthat)
library(roxygen2)
```

<!--
 You need to run the 'description' chunk in the '0-dev_history.Rmd' file before continuing your code there.

If it is the first time you use {fusen}, after 'description', you can directly run the last chunk of the present file with inflate() inside.
--> 

```{r development-load}
# Load already included functions if relevant
pkgload::load_all(export_all = FALSE)
```

# Include some data examples in your package

<!-- 
 Store your dataset in a directory named "inst/" at the root of your project.
 Use it for your tests in this Rmd thanks to `pkgload::load_all()` to make it available
and `system.file()` to read it in your examples.

- There already is a dataset in the "inst/" directory to be used in the examples below
-->

```{r development-dataset}
# Run all this chunk in the console directly
# There already is a dataset in the "inst/" directory
# Make the dataset file available to the current Rmd during development
pkgload::load_all(path = here::here(), export_all = FALSE)

# You will be able to read your example data file in each of your function examples and tests as follows - see chunks below
# datafile <- system.file("nyc_squirrels_sample.csv", package = "acheck")
# nyc_squirrels <- read.csv(datafile, encoding = "UTF-8")
# 
# nyc_squirrels
```


## Add gtsummary object to word

<!--
Create a chunk for the core of the function

- The chunk needs to be named `function` at least
- It contains the code of a documented function
- The chunk can also be named `function-my_median` to make it easily
findable in your Rmd
- Let the `@examples` part empty, and use the next `examples` chunk instead to present reproducible examples

After inflating the template

-  This function code will automatically be added in a new file in the "R/" directory
-->

```{r development-function}
# Prepare the code of your function here
```

```{r function-add_gt_to_word}
format_num <- gtsummary::label_style_number()

#' Add a gtsummary objet to word doc
#'
#' @param doc word doc created as `officer::read_docx()`
#' @param gt_obj gtsummary object
#' @param title character header
#' @param landscape logical. page orientation
#' @importFrom flextable  body_add_flextable
#' @importFrom officer body_add_par body_add_break body_end_section_landscape
#'
#' @returns a word doc
#' @export
#'
#' @examples
#' \dontrun{
#' add_gt_to_word(doc, gt_obj, title, landscape = FALSE)
#' }
add_gt_to_word <- function(doc, gt_obj, title, landscape = FALSE) {
  # Convert gtsummary to flextable
  ft <- gtsummary::as_flex_table(gt_obj) %>%
    flextable::autofit()
  
  # Add title and table
  doc <- doc %>%
    body_add_par(title, style = "heading 2") %>%
    body_add_flextable(ft) %>%
    body_add_par("") # spacing
  
  # Handle Orientation and Page Breaks
  if (landscape) {
    # Close the current landscape section
    doc <- doc %>% body_end_section_landscape() 
  } else {
    # If not landscape, just add a standard page break
    doc <- doc %>% body_add_break()
  }
  
  return(doc)
}

```



<!--
Create a chunk with an example of use for your function

- The chunk needs to be named `examples` at least
- It contains working examples of your function
- The chunk is better be named `examples-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This example will automatically be added in the '@examples' part of our function above in the "R/" directory
- This example will automatically be added in the vignette created from this Rmd template
-->


<!--
Create a chunk with a test of use for your function

- The chunk needs to be named `tests` at least
- It contains working tests of your function
- The chunk is better be named `tests-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This test code will automatically be added in the "tests/testthat/" directory
-->

```{r tests-add_gt_to_word}
library(testthat)
library(officer)
library(gtsummary)
library(dplyr)
test_that("add_gt_to_word works correctly under different conditions", {
  
  # setup: create dummy data and a basic gtsummary object
  mock_df <- data.frame(
    status = c("active", "inactive", "active"),
    age = c(25, 34, 45)
  )
  mock_gt <- gtsummary::tbl_summary(mock_df, by = status)
  
  # test 1: verification of the standard portrait flow
  doc_portrait <- officer::read_docx()
  res_portrait <- add_gt_to_word(
    doc = doc_portrait, 
    gt_obj = mock_gt, 
    title = "Test Portrait Title", 
    landscape = FALSE
  )
  
  # check that an officer document object is returned
  expect_s3_class(res_portrait, "rdocx")
  
  # convert doc to content dataframe to check text insertion
  content_p <- officer::docx_summary(res_portrait)
  expect_true(any(content_p$text == "Test Portrait Title"))
  
  # test 2: verification of the landscape flow
  doc_landscape <- read_docx()
  res_landscape <- add_gt_to_word(
    doc = doc_landscape, 
    gt_obj = mock_gt, 
    title = "Test Landscape Title", 
    landscape = TRUE
  )
  
  expect_s3_class(res_landscape, "rdocx")
  
  content_l <- docx_summary(res_landscape)
  expect_true(any(content_l$text == "Test Landscape Title"))
  
  # test 3: safe verification of landscape structural layout differences
  # landscape structural blocks alter row structures or total summary elements
  expect_false(identical(nrow(content_p), nrow(content_l)))
})
```

## clean names

```{r function-clean_username}
#' Clean names by deleting extra spaces and convert to title case
#'
#' @param x character vector of names to clean
#'
#' @returns a vector
#' @export
#'
#' @examples
#' \dontrun{clean_username()}
clean_username <- function(x){
  stringr::str_to_title(stringr::str_squish(x))
}
```



## export list of gt objects/dataframes to word

```{r  function-export_list_to_word}
#' Export a list of dataframes to word
#'
#' @param table_list list of dataframes
#' @param doc_path character vector of output directory
#' @importFrom flextable  body_add_flextable
#' @importFrom officer body_add_par body_add_break 
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_list_to_word()}
export_list_to_word <- function(table_list, doc_path) {
  
  # initialize a new word document
  doc <- officer::read_docx()
  
  # iterate through the named list
  # .x is the table/df, .y is the name of the list element
  purrr::iwalk(table_list, function(.x, .y) {
    
    # logic to handle different table types
    if (inherits(.x, "gtsummary")) {
      # convert gtsummary to flextable
      ft <- gtsummary::as_flex_table(.x)
    } else {
      # treat as standard dataframe and format
      ft <- flextable::flextable(.x) %>%
        flextable::theme_booktabs() %>%
        flextable::autofit() %>%
        flextable::bold(part = "header")
    }
    
    # add the header and the table to the document
    doc <<- doc %>%
      # use the list name as a header 2
      body_add_par(.y, style = "heading 2") %>%
      body_add_flextable(ft) %>%
      # add a small spacer after each table
      body_add_par("") 
  })
  
  # save the completed document
  print(doc, target = doc_path)
}
```


## Create gtsummary table

```{r function-create_summary_table}
#' create a `gtsummary` table
#'
#' @param data dataframe
#' @param cols columns to include
#' @param labels character. labels
#' @param group_by character. grouping variable
#' @param strata character. stratifying variable
#' @param percent_type character. percents by row or column
#' @param dichotomous_as_continuous logical. treat binary 0/1 columns as continuous
#' @param report_median logical. whether to include the median alongside the mean for continuous variables
#' @param label_characteristic character. custom header text for the characteristic column
#'
#' @returns a gtsummary object that renders across all formats natively
#' @export
create_summary_table <- function(data, 
                                 cols = dplyr::everything(), 
                                 labels = list(), 
                                 group_by = NULL,
                                 strata = NULL,
                                 percent_type = "column",
                                 dichotomous_as_continuous = FALSE,
                                 report_median = FALSE,
                                 label_characteristic = "Characteristic") {
  
  # internal helper to build the base table structure
  build_summary <- function(df) {
    summary_types <- list(all_categorical() ~ "categorical")
    if (dichotomous_as_continuous) {
      summary_types <- c(summary_types, list(gtsummary::all_dichotomous() ~ "continuous"))
    }
    
    continuous_stat <- if (report_median) "{mean} ({median})" else "{mean}"
    
    df %>%
      dplyr::select({{cols}}, dplyr::any_of(group_by)) %>%
      gtsummary::tbl_summary(
        by = dplyr::any_of(group_by),
        label = labels,
        percent = percent_type,
        type = summary_types,
        statistic = list(
          all_continuous() ~ continuous_stat,
          all_categorical() ~ "{n} ({p}%)"
        ),
        digits = list(
          # lower case comments without dots or dashes
          # dynamically switch style format if handling raw numbers vs percentage indexes
          all_continuous() ~ function(x) {
            val <- na.omit(x)
            if (length(val) > 0 && max(val) > 1) {
              return(gtsummary::style_number(x, digits = 0))
            } else {
              return(paste0(gtsummary::style_number(x * 100, digits = 0), "%"))
            }
          },
          all_categorical() ~ c(0, 1)
        ),
        missing = "no"
      ) %>%
      gtsummary::add_n() %>%
      gtsummary::bold_labels() %>%
      gtsummary::modify_header(label ~ paste0("**", label_characteristic, "**")) %>%
      gtsummary::modify_footnote(gtsummary::all_stat_cols() ~ NA) %>%
      gtsummary::modify_table_body(function(x) {
        attr(x, "table_styling")$columns_width <- NULL
        return(x)
      })
  }
  
  # logic to handle stratification cleanly
  raw_gtsummary <- if (!is.null(strata)) {
    data %>%
      gtsummary::tbl_strata(
        strata = all_of(strata),
        .tbl_fun = ~ build_summary(.x)
      )
  } else {
    build_summary(data)
  }
  
  return(raw_gtsummary)
}
```

## create a listing table using `gtreg{}`

```{r function-create_listing_table}

#' create `gtreg` table listing of data
#'
#' @param data dataframe
#' @param cols columns to include
#' @param labels character. labels
#' @param group_by character. grouping variable 
#'
#' @returns a list
#' @export
#'
#' @examples
#'\dontrun{create_listing_table()}
create_listing_table <- function(data,
                                 cols = dplyr::everything(),
                                 labels = list(),
                                 group_by = NULL) {
  data %>%
    {
      if (!is.null(group_by))
        dplyr::arrange(., .data[[group_by]])
      else
        .
    } %>%
    dplyr::select({{ cols }}, dplyr::any_of(group_by)) %>%
    gtreg::tbl_listing() %>%
    gtsummary::bold_labels()
}

```

## clean a vector of strings

```{r function-clean_string}
# clean string

#' clean string of spaces and non alphanumeric characters and make lower case
#'
#' @param x a vector of characters
#'
#' @returns a vector
#' @export
#'
#' @examples
#' \dontrun{clean_String('Nachola124')}


clean_string <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_to_lower() %>%
    stringr::str_replace_all("[^a-z0-9]", "") %>%
    stringr::str_trim()
}
```

## recode beneficiary camps to known refugee camps

```{r function-clean_camp}
# clean ben_camp

#' clean recode beneficiary camp 
#'
#' @param x dataframe
#' @param col_name  character. column with camp names
#'
#' @returns dataframe
#' @importFrom dplyr case_when all_of across
#' @export
#'
#' @examples
#' \dontrun{clean_camp(x, col_name = "ben_camp")}
clean_camp <- function(x, col_name = "ben_camp"){
  x  %>% 
    dplyr::mutate(
      across(all_of(col_name), ~ case_when(
        # use stringr or grepl with the .x pronoun
        grepl("\\bna\\b", .x, ignore.case = TRUE) & grepl("HIAS", reforgp, ignore.case = T) ~ "Nairobi",
        grepl("^k", .x, ignore.case = TRUE) ~ "Kakuma",
        grepl("^nai|east|urban|umoja", .x, ignore.case = TRUE) ~ "Nairobi",
        grepl("^daga|ley$|^ifo|f0|fo|^haga|^v|dad|daad", .x, ignore.case = TRUE) ~ "Daadab",
        TRUE ~ .x
      ))
    )
}
```


## get group columns

```{r function-get_group_cols}
#' Extract column names from a data dictionary based on structural boundaries
#'
#' @param dict dataframe representing the data dictionary
#' @param name_col string. The column containing the variable names to extract
#' @param vulnerability_df a dataframe with vulnerability categories
#' @param vulnerability_class character vector of vulnerability class
#'
#' @returns a character vector of column names
#' @export
get_group_cols <- function(dict, vulnerability_df, vulnerability_class = "physio", name_col = "name") {
  
  df <- vulnerability_df %>%
    dplyr::filter(
      grepl(vulnerability_class, label, ignore.case = TRUE)
    )
  dict %>%
    dplyr::filter(n >= df$n & n < df$n_end) %>%
    # remove true NAs and any literal "NA" character strings
    dplyr::filter(!is.na(!!rlang::sym(name_col)) & !!rlang::sym(name_col) != "NA") %>%
    dplyr::pull(!!rlang::sym(name_col))
}
```


## find households in ESR data

```{r function-check_esr_registrations}
# registrations appearing in ESR
#' Title
#'
#' @param data1 dataframe. registration data
#' @param data2 dataframe. ESR data
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{check_esr_registrations(data1 = df_clean, data2 = df_esr)}
check_esr_registrations <- function(data1 = df_clean, data2 = df_esr) {

  
  # 1. prepare lookup keys from data2
  # combine names and clean id and phone for matching
  data2_prepared <- data2 %>%
    dplyr::mutate(
      full_name_clean = clean_string(paste(first_name, middle_name, last_name)),
      first_surname = clean_string(paste(first_name, last_name)),
      first_secname = clean_string(paste(first_name, middle_name)),
      id_clean        = clean_string(id_number1),
      phone_clean     = clean_string(phone_number)
    )
  
  lookup_ids    <- unique(data2_prepared$id_clean)
  lookup_names  <- unique(data2_prepared$full_name_clean)
  lookup_names1  <- unique(data2_prepared$first_surname)
  lookup_names2  <- unique(data2_prepared$first_secname)
  lookup_phones <- unique(data2_prepared$phone_clean)
  
  # 2. apply matching logic to data1
  data1 %>%
    dplyr::mutate(
      is_in_second_dataset = case_when(
        # check by id number
        clean_string(ben_id_number) %in% lookup_ids ~ "yes",
        
        # check by phone number using reg_num_phone
        clean_string(reg_num_phone) %in% lookup_phones ~ "yes",
        
        # check by combined name
        clean_string(ben_name) %in% lookup_names ~ "yes",
        clean_string(ben_first_surname) %in% lookup_names1 ~ "yes",
        clean_string(ben_first_secname) %in% lookup_names2 ~ "yes",
        
        # default to no if no matches are found
        TRUE ~ "no"
      )
    )
}
```

## get standard villages

```{r function-get_standard_villages}
#' separate villages collapsed by forward slash ("|")
#'
#' @param data1 dataframe. ESR/partner data
#' @param data2 dataframe. registration data
#'
#' @returns a dataframe
#' @importFrom stringr str_extract str_replace_all
#' @export
#'
#' @examples
#' \dontrun{get_standard_villages(data1 = df_partner, data2 = df_clean)}
get_standard_villages <- function(data1 = df_partner, data2 = df_clean) {
  
  # get unique reference villages and sort by length descending
  # this ensures long names are matched before substrings
  villages <- unique(data2$village2)
  villages_sorted <- villages[order(nchar(villages), decreasing = TRUE)]
  
  # escape special characters and collapse for regex
  village_pattern <- paste(stringr::str_escape(villages_sorted), collapse = "|")
  
  data1 %>%
    # separate rows based on the forward slash
    # this turns "village1/village2" into two distinct rows
    tidyr::separate_rows(village, sep = "/") %>%
    dplyr::mutate(
      # extract the standard name using the pattern
      village2 = str_extract(village, regex(village_pattern, ignore_case = TRUE)),
      # capitalize first letters of village names as per your requirement
      # this handles slashes if they appear in the result
      village2 = str_replace_all(village2, "(^|/)([a-z])", function(m) toupper(m))
    )
}
```

## pull kobo data

```{r function-pull_kobo_data}
#' pull data from kobo API
#'
#' @param url character. Kobotoolbox url
#' @param project_id character. project id
#' @param token character. project token
#'
#' @returns a list
#' @export
#'
#' @examples
#' \dontrun{pull_kobo_data()}
pull_kobo_data <- function(
    url = Sys.getenv("kobo_url"),
    project_id = Sys.getenv("unocha_id"),
    token = Sys.getenv("unocha_token")
    
){
  
  robotoolbox::kobo_setup(url = url, token = token)
  
  df <- robotoolbox::kobo_data(project_id)
  return(df)
}
```

## categorize numerics

```{r function-cut_numeri}
#' categorize numeric cols
#'
#' @param x a vector of numeric column
#' @param breaks numeric vector of breaks
#' @param labels character vector of labels
#'
#' @returns vector
#' @export
#'
#' @examples
#' cut_numeric(c(1:10), c(-Inf,5, Inf), c("<=5", ">=6"))
cut_numeric <- function(
    x, 
    breaks = c(16, 35, 60, Inf), 
    labels = c("16 - 35", "35 - 59", ">= 60")
){
  out <- cut(
    x,
    # 16 is the start, 35 and 60 are the breaks, inf covers everything above
    breaks = breaks,
    # labels for the resulting factor
    labels = labels,
    # include the lowest value and make intervals right-open [x, y)
    include.lowest = TRUE,
    right = FALSE
  )
  return(out)
}
```

## export gtsummary object to excel workbook

```{r function-export_to_excel_formatted}
#' Export/write gtsummary output to excel workbook
#'
#' @param list_of_objects list of `gtsummary` object
#' @param file_name character. output filename and associated path
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_to_excel_formatted()} 
export_to_excel_formatted <- function(
    list_of_objects,
    file_name = paste0("UNOCHA_Summary_Report_", Sys.Date(), ".xlsx")
) {
  
  # create the workbook object
  wb <- openxlsx::createWorkbook()
  
  # define styles
  neg_style <- createStyle(fontColour = "#FF0000", textDecoration = "bold")
  # style for bolding the row headers (the first column)
  bold_header_style <- createStyle(textDecoration = "bold")
  
  # iterate through the named list
  purrr::iwalk(list_of_objects, function(obj, sheet_name) {
    
    openxlsx::addWorksheet(wb, sheet_name)
    
    # conversion logic
    df_to_write <- if (inherits(obj, "gtsummary")) {
      as_tibble(obj)
    } else {
      as.data.frame(obj)
    }
    
    openxlsx::writeData(wb, sheet_name, df_to_write)
    
    # 1. bold the row headers (first column)
    # rows starts at 2 to skip the excel header row
    openxlsx::addStyle(
      wb, 
      sheet_name, 
      style = bold_header_style, 
      cols = 1, 
      rows = 2:(nrow(df_to_write) + 1), 
      gridExpand = TRUE
    )
    
    # 2. apply red color coding for negative values
    openxlsx::conditionalFormatting(
      wb, 
      sheet_name, 
      cols = 1:ncol(df_to_write), 
      rows = 2:(nrow(df_to_write) + 1), 
      rule = "< 0", 
      style = neg_style
    )
    
    openxlsx::setColWidths(wb, sheet_name, cols = 1:ncol(df_to_write), widths = "auto")
  })
  
  openxlsx::saveWorkbook(wb, file = file.path("output", file_name), overwrite = TRUE)
}
```

## add a new sheet to existing workbook

```{r function-export_to_excel_appended}
#' Export a sheet to existing workbook
#'
#' @param list_of_objects list of `gtsummary` object
#' @param sheetname character. worksheet name
#' @param file_name character. output file name
#' @param output_dir character. output directory
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_to_excel_appended()}
export_to_excel_appended <- function(
    list_of_objects = df_target_p,
    sheetname = "Target and Registered ESR",
    file_name = paste0("UNOCHA_Summary_Report_", Sys.Date(), ".xlsx"),
    output_dir = "output"
) {
  
  # create output dir
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # build full path safely
  file_path <- file.path(output_dir, basename(file_name))
  
  message("Using file path: ", normalizePath(file_path, mustWork = FALSE))
  
  # load existing workbook OR create new
  if (file.exists(file_path)) {
    
    message("Loading existing workbook...")
    
    wb <- tryCatch(
      openxlsx::loadWorkbook(file_path),
      error = function(e) {
        message("Failed to load workbook: ", e$message)
        openxlsx::createWorkbook()
      }
    )
    
  } else {
    
    message("Workbook does not exist. Creating new workbook...")
    wb <- openxlsx::createWorkbook()
    
  }
  
  # excel sheet name limit
  sheetname <- substr(sheetname, 1, 31)
  
  # remove existing sheet if present
  if (sheetname %in% names(wb)) {
    openxlsx::removeWorksheet(wb, sheetname)
  }
  
  addWorksheet(wb, sheetname)
  
  # styles
  header_style <- createStyle(
    fontSize = 12,
    textDecoration = "bold",
    fontColour = "#0066A1"
  )
  
  bold_row_style <- openxlsx::createStyle(textDecoration = "bold")
  
  neg_style <- openxlsx::createStyle(
    fontColour = "#FF0000",
    textDecoration = "bold"
  )
  
  current_row <- 1
  
  purrr::iwalk(list_of_objects, function(obj, list_name) {
    
    df_to_write <- if (inherits(obj, "gtsummary")) {
      as_tibble(obj)
    } else {
      as.data.frame(obj)
    }
    
    # section title
    openxlsx::writeData(
      wb,
      sheet = sheetname,
      x = list_name,
      startRow = current_row
    )
    
    openxlsx::addStyle(
      wb,
      sheet = sheetname,
      style = header_style,
      cols = 1,
      rows = current_row
    )
    
    current_row <<- current_row + 1
    
    # write table
    openxlsx::writeData(
      wb,
      sheet = sheetname,
      x = df_to_write,
      startRow = current_row,
      headerStyle = bold_row_style
    )
    
    # bold first column
    if (nrow(df_to_write) > 0) {
      
      openxlsx::addStyle(
        wb,
        sheet = sheetname,
        style = bold_row_style,
        cols = 1,
        rows = (current_row + 1):(current_row + nrow(df_to_write)),
        gridExpand = TRUE,
        stack = TRUE
      )
      
      openxlsx::conditionalFormatting(
        wb,
        sheet = sheetname,
        cols = 1:ncol(df_to_write),
        rows = (current_row + 1):(current_row + nrow(df_to_write)),
        rule = "<0",
        style = neg_style
      )
    }
    
    current_row <<- current_row + nrow(df_to_write) + 3
  })
  
  # autofit
  openxlsx::setColWidths(
    wb,
    sheet = sheetname,
    cols = 1:50,
    widths = "auto"
  )
  
  # save
  openxlsx::saveWorkbook(
    wb,
    file = file_path,
    overwrite = TRUE
  )
  
  message("Workbook saved successfully.")
}
```

## get duplicates

```{r function-get_duplicates}
#' get duplicate beneficiaries on select variables
#'
#' @param x dataframe
#' @param vars_to_check character vector of variables to check
#' @param phone_var character vector. phone number variable for hard check
#' @param name_var character vector. name of respondent
#' @importFrom dplyr group_by across n n_distinct filter any_of ungroup distinct
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_duplicates()}
get_duplicates <- function(
    x,
    vars_to_check = c("benef_name", "ben_number"),
    phone_var = "ben_number",
    name_var = "benef_name"
){
  
  # 1. find phone sharing: same phone, different names
  phone_sharing <- x %>%
    group_by(across(any_of(phone_var))) %>%
    filter(n() > 1 & n_distinct(across(any_of(name_var))) > 1)
  
  # 2. find name sharing: same name, different phones
  name_sharing <- x %>%
    group_by(across(any_of(name_var))) %>%
    filter(n() > 1 & n_distinct(across(any_of(phone_var))) > 1)
  
  # 3. find exact duplicates across the vars_to_check
  exact_dups <- x %>%
    group_by(across(any_of(vars_to_check))) %>%
    filter(n() > 1)
  
  # combine all instances and remove overlapping rows
  res <- bind_rows(phone_sharing, name_sharing, exact_dups) %>%
    distinct() %>%
    ungroup()
  
  # sort by the primary variables for easy review
  res %>%
    arrange(across(any_of(vars_to_check)))
}
```

## get distinct

```{r function-get_distinct}
#' get distinct beneficiaries on select variables
#'
#' @param x dataframe
#' @param vars_to_check character vector of columns to check
#' @param phone_var character vector of phone number column
#' @param name_var character vector of respondent name
#' @importFrom dplyr group_by across n n_distinct filter any_of ungroup distinct select pull row_number bind_rows slice
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_distinct()}
get_distinct <- function(
    x,
    vars_to_check = c("benef_name", "ben_number"),
    phone_var = "ben_number",
    name_var = "benef_name"
){
  # capture exact duplicates and instances of shared name with different phones
  anomalies <- x %>%
    mutate(temp_id = row_number()) %>%
    {
      bind_rows(
        # scenario 2 shared name different phones
        group_by(., across(any_of(name_var))) %>% 
          filter(n() > 1 & n_distinct(across(any_of(phone_var))) > 1),
        
        # scenario 3 exact duplicates
        group_by(., across(any_of(vars_to_check))) %>% 
          filter(n() > 1)
      )
    } %>%
    distinct(temp_id, .keep_all = TRUE)
  
  # isolate clean rows that dont participate in those loops
  clean_data <- x %>%
    mutate(temp_id = row_number()) %>%
    filter(!(temp_id %in% anomalies$temp_id))
  
  # resolve the anomalies by keeping just the first instance per unique variable combo
  resolved_anomalies <- anomalies %>%
    group_by(across(any_of(vars_to_check))) %>%
    slice(1) %>%
    ungroup()
  
  # bind clean entries and resolved entries back together
  bind_rows(clean_data, resolved_anomalies) %>%
    arrange(temp_id) %>%
    select(-temp_id)
}
```

## plot bar

```{r function-plot_bar}
#' plot bar graphs
#'
#' @param data dataframe
#' @param x_var character vector x-axis variable
#' @param y_var character vector y-axis variable
#' @param fill_col character vector of hex color code for fill
#' @param title character vector of title
#' @param subtitle character vector of subtitle
#' @param x_lab character vector of label for x-axis
#' @param y_lab character vector of label for y-axis
#' @importFrom rlang ensym
#'
#' @returns list of plot object
#' @export
#'
#' @examples
#' \dontrun{plot_bar()}
plot_bar <- function(data, 
                     x_var, 
                     y_var = NULL, 
                     fill_col = "#0066A1",
                     title = "variable distribution", 
                     subtitle = NULL, 
                     x_lab = NULL, 
                     y_lab = "count") {
  
  x_sym <- rlang::ensym(x_var)
  
  if (!is.null(y_var)) {
    y_sym <- ensym(y_var)
    
    # case 1: plotting pre-calculated values with percent
    # we calculate total sum to derive percentages for the labels
    total_sum <- sum(data[[as_label(y_sym)]], na.rm = TRUE)
    
    # p <- ggplot(data, aes(x = fct_reorder(as.factor(!!x_sym), !!y_sym), y = !!y_sym)) +
    p <- ggplot2::ggplot(data, aes(x = !!x_sym, y = !!y_sym)) +
      ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
      ggplot2::geom_text(
        ggplot2::aes(label = paste0(!!y_sym, " (", scales::percent(!!y_sym / total_sum, accuracy = 0.1), ")")),
        vjust = -0.5, size = 3, fontface = "bold"
      )
    
  } else {
    # case 2: plotting frequencies with dynamic percent
    # p <- ggplot(data, aes(x = fct_infreq(as.factor(!!x_sym)))) +
    p <- ggplot2::ggplot(data, aes(x = !!x_sym)) +
      ggplot2::geom_bar(fill = fill_col, alpha = 0.9, width = 0.75) +
      ggplot2::geom_text(
        stat = "count",
        ggplot2::aes(label = after_stat(paste0(count, " (", scales::percent(count / sum(count), accuracy = 0.1), ")"))),
        vjust = -0.5, size = 3, fontface = "bold"
      )
  }
  
  p <- p +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = if(is.null(x_lab)) as_label(x_sym) else x_lab,
      y = y_lab
    ) +
    # expand y axis further to accommodate longer labels
    ggplot2::scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14, color = "#231F20"),
      plot.subtitle = element_text(size = 11, color = "#58595B"),
      axis.title = element_text(face = "bold", color = "#231F20"),
      axis.text.x = element_text(angle = 45, hjust = 1, color = "#231F20"),
      axis.text.y = element_text(color = "#231F20")
    )
  
  return(p)
}
```

## add custom total

```{r function-add_custom_total}
#' calculate column-wise totals
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{add_custom_total()}
add_custom_total <- function(df) {
  # calculate totals
  totals <- df %>%
    dplyr::summarise(
      Village = "Total",
      # only sum unique targeted values per village to avoid double counting
      `Targeted BNF` = base::sum(unique(`Targeted BNF`), na.rm = TRUE),
      `Registered BNF` = sum(`Registered BNF`, na.rm = TRUE),
      Variance = sum(Variance, na.rm = TRUE),
      `ESR Households` = sum(`ESR Households`, na.rm = TRUE),
      partner = NA
    )
  
  # combine with original data
  out <- bind_rows(df, totals)
  
  return(out)
}
```

## collapse villages

```{r function-collapse_villages}
#' collapse villages that share targets, separating by /
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{collapse_villages()}
collapse_villages <- function(df) {
  df_collapsed <- df %>%
    # group by the columns where you want to identify duplicates
    dplyr::group_by(`Targeted BNF`, `ESR Households`, partner) %>%
    dplyr::summarise(
      # collapse village names with a slash
      village = paste(unique(village), collapse = " / "),
      # sum the registered beneficiaries
      `Registered BNF` = sum(`Registered BNF`, na.rm = TRUE),
      # keep the group order clean
      .groups = "drop"
    ) %>%
    # relocate columns to match your original structure
    dplyr::select(village, `Targeted BNF`, `ESR Households`, partner, `Registered BNF`)
  
  return(df_collapsed)
}
```

## find partial matches

```{r function-get_partial_matches}
#' find partial matches between focal and beneficiary
#'
#' @param df dataframe
#' @param focal character vector of focal variable
#' @param beneficiary character vector of beneficiary variable
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{get_partial_matches()}
get_partial_matches <- function(df, focal = "focal_name", beneficiary = "benef_name") {
  df %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      is_partial = {
        f_raw <- .data[[focal]]
        b_raw <- .data[[beneficiary]]
        
        # 1 early exit check for missing values or empty strings
        if (is.na(f_raw) || is.na(b_raw) || f_raw == "" || b_raw == "" || f_raw == "NA" || b_raw == "NA") {
          FALSE
        } else {
          f_str <- as.character(f_raw)
          b_str <- as.character(b_raw)
          
          # 2 basic substring check first using base r grepl to avoid vector length bugs
          matched <- grepl(b_str, f_str, fixed = TRUE) | grepl(f_str, b_str, fixed = TRUE)
          
          # 3 if basic check fails look for out of order word inclusion
          if (isTRUE(matched)) {
            matched
          } else {
            # clean and split both strings into words
            f_words <- unlist(stringr::str_split(f_str, "\\s+"))
            b_words <- unlist(stringr::str_split(b_str, "\\s+"))
            
            # identify which name is shorter by token count
            if (length(f_words) <= length(b_words)) {
              short_words <- f_words
              long_str <- b_str
            } else {
              short_words <- b_words
              long_str <- f_str
            }
            
            # verify if every word of the shorter name exists in the longer string
            all(purrr::map_lgl(short_words, ~ grepl(.x, long_str, fixed = TRUE)))
          }
        }
      }
    ) %>% 
    dplyr::ungroup()
}
```

## get personally identifying columns

```{r function-get_pii}
#' Get personally identifying columns
#'
#' @param df dataframe
#' @param vars string of identifying columns
#'
#' @returns a list of 3 dataframes: identifying, de-identified, and original df
#' @export
#'
#' @examples
#' \dontrun{get_pii()}
get_pii <- function(
    df = df_clean,
    vars = c("lat", "long", "name", "mobile", "gps", "phone", "enum_num$", "^(enum|reg|ben|proxy|sec)_.*(name|num|id)")
){
  cols <- paste(vars, collapse = "|")
  df_pii <- df %>% select(
    grep(cols, names(.))
  )
  df_depii <- df %>%
    dplyr::select(
      -grep(cols, names(.))
    )
  out <- list(df_pii, df_depii, df) %>%
    purrr::set_names(c("personaly-identifying", "de-identified", "original-data"))
  return(out)
}
```

## export a list to excel

```{r function-write_list_to_excel}
#' export a named list of data frames to an excel workbook
#' @param df_list a named list of data frames
#' @param file_path character string for the output path
#' @returns NULL
#' @export
#' @examples
#' \dontrun{write_list_to_excel()}
write_list_to_excel <- function(df_list, file_path) {
  
  # create a new workbook object
  wb <- openxlsx::createWorkbook()
  
  # define a professional header style
  header_style <- createStyle(
    fgFill = "#4F81BD", 
    halign = "center", 
    textDecoration = "bold",
    fontColour = "white",
    border = "bottom"
  )
  
  # iterate through the named list
  # .x is the dataframe, .y is the name of the list element
  purrr::iwalk(df_list, function(.x, .y) {
    
    # add a new worksheet with the name of the list element
    # excel sheet names are limited to 31 characters
    sheet_name <- substr(.y, 1, 31)
    openxlsx::addWorksheet(wb, sheet_name)
    
    # write the data to the sheet
    openxlsx::writeData(wb, sheet_name, .x, headerStyle = header_style)
    
    # adjust column widths to fit content automatically
    openxlsx::setColWidths(wb, sheet_name, cols = 1:ncol(.x), widths = "auto")
    
    # freeze the top row for easier navigation
    openxlsx::freezePane(wb, sheet_name, firstRow = TRUE)
  })
  
  # save the workbook to the specified path
  openxlsx::saveWorkbook(wb, file_path, overwrite = TRUE)
}
```

## pre-process unocha registration


```{r function-preprocess_reg}
#' preprocess unocha registrations
#'
#' @param df dataframe
#' @importFrom dm dm_flatten_to_tbl
#' @importFrom dplyr mutate group_by case_when if_else filter bind_rows distinct
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{preprocess_reg()}
#' 
preprocess_reg <- function(df){
  df_out <- df %>% 
  dm_flatten_to_tbl(main)

df_out <- df_out %>%
  mutate(across(where(labelled::is.labelled), as.character))
df_out <- df_out %>%
  dplyr::mutate(
    village = case_when(
      grepl("chini", village, ignore.case = T) & grepl(
        "kalakacha", sublocation, ignore.case = T
      ) ~ "Kalakacha Chini",
      TRUE ~ village
    )
  )
df_out <- df_out %>%
  mutate(
    village = case_when(
      grepl("Emmaus|Kirengender", village, ignore.case = T) ~ "Kirengender",
      grepl("salama|Kajificheni", village, ignore.case = T) ~ "Salama",
      grepl("kalkacha_juu", village, ignore.case = T) ~ "Kalakacha Juu",
      TRUE ~ village
    )
#     troublesome
    ,
    enum_egency = case_when(
      grepl("mitobini", village, ignore.case = T) &
      grepl("pgi", enum_egency, ignore.case = T) ~ "ALDEF",
      # grepl("Milimani", village, ignore.case = T) & 
      #   date(start) >= date("2026-05-13") ~ "ALDEF",
      # grepl("kurole", village, ignore.case = T) ~ "ALDEF",
      grepl("elwachu", village, ignore.case = T) ~ "ALDEF",
      TRUE ~ enum_egency
    )
  )

df_out <- df_out %>% 
  dplyr::mutate(
    ben_first_surname = paste(first_name, third_name),
    ben_first_secname = paste(first_name, second_name)
  )

df_out <- df_out %>% 
  dplyr::mutate(
    HHH_age_calc = if_else(
      !is.na(resp_age), resp_age, HHH_age
    )
  )

df_out <- df_out %>%
  dplyr::mutate(dplyr::across(dplyr::any_of(
    c("county", "subcounty", "ward", "sublocation", "village")
  ), ~ gsub("_", " ", .x)))

df_out <- df_out %>%
  mutate(
    proxy_name = stringr::str_squish(
      paste(
        dplyr::coalesce(name_proxy_first, ""),
        dplyr::coalesce(name_proxy_second, ""),
        dplyr::coalesce(name_proxy_third, "")
      )
    ) 
  )


df_out <- df_out %>%
  dplyr::mutate(dplyr::across(
    dplyr::any_of(
      c(
        "county",
        "subcounty",
        "ward",
        "sublocation",
        "village",
        "ben_name",
        "proxy_name"
      )
    ),
    ~ stringr::str_to_title(stringr::str_squish(.x))
  ))

df_out <- df_out %>%
  dplyr::mutate(
    age_category = cut_numeric(HHH_age_calc),
    hh_size_category = cut_numeric(
      hh_size,
      breaks = c(1, 3, 6, 10, Inf),
      labels = c("1 - 2", "3 - 5", "6 - 9", ">= 10")
    ),
    hh_disabl_category = cut_numeric(
      hh_disability_count,
      breaks = c(0,1, 3, 5, Inf),
      labels = c("0", "1 - 2", "3 - 4", ">= 5")
    )
  )

  
df_out <- df_out %>%
  labelled::set_variable_labels(
    age_category = "Household\n Head Age",
    HHH_chronic = "HH Head\n Chronic illness",
    HHM_chronic = "HH Member\n Chronic illness",
    average_income = "Monthly Income",
    hh_size = "Household Size",
    hh_size_category = "Household Size",
    FCSGName = "Food Consumption\n Score",
    eligibility = "Eligibility",
    enum_egency = "Agency",
    enum_name = "Enumerator\n first name",
    first_name = "respondent\n first name",
    reg_num = "Respondent Phone",
    trust_proxy = "Trust proxy",
    enum_egency = "Partner",
    proxy_num_consent = "proxy consent",
    name_proxy_first = "proxy\n first name",
    proxy_num = "proxy mobile",
    proxy_relation = "proxy relation",
    HHM_disability = "Household\n disability",
    HHH_chronic = "Household\nchronic illness"
  )

# registration started May 3 ----------------------------------------------

 
  
  # distinct(
  #   ben_name, reg_num, ben_id_number, .keep_all = T
  # ) %>%
  # distinct(
  #   reg_num, ben_id_number, .keep_all = T
  # ) %>% 
  
# reg_num can be NA but distinct() would only pick one row


}
```

## slice reg data

```{r function-slice_reg}
#' slice registration to match starting dates
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{slice_reg()}
slice_reg <- function(df){
  df_out <- df %>% 
  dplyr::filter(
#     start of reg
    start >= lubridate::date("2026-05-03")
  ) %>%
  dplyr::filter(
    !(
    lubridate::as_date(start) <= lubridate::date("2026-05-06") & 
    grepl("SND", enum_egency, ignore.case = TRUE)
  )
  )
  df_out
}
```


## get partner edit files

```{r function-get_edit_data}
#' Read partner registration edit files
#'
#' @param partner character string of partners e.g., "PGI" 
#' @param cols_exclude character vector of columns to exclude from edits
#' @param sheet_name character vector of workbook sheet name
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_edit_data()}
get_edit_data <- function(
    partner = "PGI",
    cols_exclude = c("village", "ben_name"),
    sheet_name = NULL
){
  p_file <- list.files(
    file.path(here::here(),"edits"), pattern = partner, full.names = TRUE,
    ignore.case = TRUE
  )
  if(!is.null(sheet_name)){
    sheet_f <- sheet_name
  }else{
    sheet_f <- readxl::excel_sheets(p_file)
  }
  
  df <- readxl::read_excel(p_file, sheet = sheet_f) %>%
    clean_names()   %>% mutate(
      ben_name = stringr::str_squish(ben_name) %>% stringr::str_to_title()
    ) 
  df_1 <- df %>%
    
    # left_join(
    #   df_out %>% 
    #   #   filter(
    #   #   grepl("Tana", county, ignore.case = T)
    #   # ) %>%
    #     select(
    #     ben_name, ben_id_number
    #   )
    # ) %>% 
    filter(
      !grepl(paste(cols_exclude, collapse = "|"), var_to_change)
    )
  return(df_1)
}
```



## Proportional to size random sample

```{r function-draw_pps_sample}
#' Draw PPS Household Sample
#'
#' @param data Household-level data frame.
#' @param county_var County variable name.
#' @param village_id_var Village identifier variable name.
#' @param allocation_df County allocation table.
#' @param alloc_size_var County sample size variable.
#' @param seed Random seed.
#'
#' @return Sampled household data frame.
#' @export
draw_pps_sample <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){
  
  set.seed(seed)
  
  # ----------------------------
  # STEP 1: PPS village allocation
  # ----------------------------
  
  village_alloc <- data %>%
    dplyr::count(
      dplyr::across(
        dplyr::all_of(
          c(county_var, village_id_var)
        )
      ),
      name = "village_size"
    ) %>%
    dplyr::left_join(
      allocation_df,
      by = county_var
    ) %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::all_of(county_var)
      )
    ) %>%
    dplyr::mutate(
      
      county_total = sum(village_size),
      
      raw_alloc =
        (.data[[alloc_size_var]] *
           village_size) /
        county_total,
      
      base_alloc = floor(raw_alloc),
      
      frac = raw_alloc - base_alloc,
      
      remaining =
        first(.data[[alloc_size_var]]) -
        sum(base_alloc),
      
      sample_alloc =
        base_alloc +
        as.integer(
          dplyr::row_number(
            dplyr::desc(frac)
          ) <= remaining
        )
      
    ) %>%
    dplyr::ungroup()
  
  # ----------------------------
  # STEP 2: Sample village-by-village
  # ----------------------------
  
  sampled <- village_alloc %>%
    
    dplyr::group_split(
      dplyr::across(
        dplyr::all_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    
    purrr::map_dfr(function(v){
      
      n_take <- v$sample_alloc[[1]]
      
      hh <- data %>%
        dplyr::semi_join(
          v,
          by = c(
            county_var,
            village_id_var
          )
        )
      
      dplyr::slice_sample(
        hh,
        n = min(
          n_take,
          nrow(hh)
        )
      )
      
    })
  
  sampled
}

```

## proportional to size sample Hamilton redistribution

```{r function-draw_pps_hamilton}
#' Draw a Sample of Households Proportional to Village Size
#'
#' This function selects a stratified random sample of individual household rows
#' from within counties. The probability of a household being selected is
#' proportional to the total size (household count) of the village it belongs to,
#' using the Hamilton (largest remainder) method for allocation. After initial
#' allocation, any village falling below \code{min_village_sample} has its
#' allocation lifted to the minimum by borrowing units from the most oversampled
#' donor villages within the same county.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var A bare or quoted column name identifying the county stratum.
#' @param village_id_var A bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing two columns: the county identifier
#'   and the allocated household sample size for that county.
#' @param alloc_size_var A bare or quoted column name in \code{allocation_df}
#'   indicating the target sample size for each county.
#' @param seed An integer used to initialise the pseudo-random number generator
#'   for reproducible sampling. Defaults to the numeric representation of the
#'   current system date.
#' @param min_village_sample A single integer specifying the minimum number of
#'   households that must be sampled from any village. Villages initially allocated
#'   fewer than this value borrow units from the most oversampled villages in the
#'   same county. The effective minimum is capped at the village's actual household
#'   count, so small villages are never over-sampled. Defaults to \code{8L}.
#'
#' @return A data frame with the same columns as \code{data}, containing only the
#'   selected household rows. Villages with a final allocation of zero are excluded
#'   from the output entirely.
#'
#' @details
#' \strong{Allocation (step 1 — Hamilton / largest remainder):}
#' For each county, each village receives a proportional raw allocation
#' \eqn{n_v = N_{county} \times (s_v / S_{county})}, where \eqn{s_v} is the
#' village household count and \eqn{S_{county}} is the county total. Allocations
#' are floored and fractional remainders are distributed one unit at a time to the
#' villages with the largest fractional parts until the county target is met.
#'
#' \strong{Redistribution (step 2 — minimum floor):}
#' Redistribution is performed independently within each county. Donor villages —
#' those whose allocation exceeds their own effective minimum — are sorted in
#' descending order of surplus and drained greedily until the total deficit of
#' sub-minimum villages is recovered or the donor pool is exhausted. A
#' \code{\link[base]{warning}} is issued when a county cannot fully satisfy the
#' minimum for all its villages.
#'
#' \strong{Sampling (step 3):}
#' Within each village, households are drawn via simple random sampling without
#' replacement using random uniform ordering. Villages with a final
#' allocation of zero are silently dropped.
#'
#' @seealso \code{\link[dplyr]{slice_sample}}, \code{\link[dplyr]{group_modify}}
#'
#' @importFrom dplyr count across any_of left_join group_by mutate ungroup group_modify filter arrange row_number select
#' @importFrom rlang .data sym !!
#' @importFrom stats runif
#'
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw household data
#' raw_households <- data.frame(
#'   county_id  = rep(c("County_A", "County_B"), each = 10),
#'   village_id = c(rep("V1", 3), rep("V2", 7), rep("V3", 4), rep("V4", 6)),
#'   household_id = paste0("HH_", 1:20)
#' )
#'
#' # household sample size targets per county
#' target_allocations <- data.frame(
#'   county_id   = c("County_A", "County_B"),
#'   sample_size = c(20, 25)
#' )
#'
#' # run reproducible PPS household sampling with an 8-household village minimum
#' pps_household_sample <- draw_pps_sample(
#'   data               = raw_households,
#'   county_var         = "county_id",
#'   village_id_var     = "village_id",
#'   allocation_df      = target_allocations,
#'   alloc_size_var     = "sample_size",
#'   seed               = 123,
#'   min_village_sample = 8L
#' )
#' }
draw_pps_hamilton <- function(
    data,
    county_var         = "county",
    village_id_var     = "village",
    allocation_df,
    alloc_size_var     = "sample",
    seed               = as.numeric(Sys.Date()),
    min_village_sample = 8L
) {
  
  # preserve random seed state if it exists to maintain reproducibility safely
  if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
    old_seed <- get(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
    on.exit(assign(".Random.seed", old_seed, envir = .GlobalEnv))
  }
  
  set.seed(seed)
  
  # convert input column names safely to standard strings
  county_str <- as.character(substitute(county_var))
  if (!county_str %in% names(data)) county_str <- county_var
  
  village_str <- as.character(substitute(village_id_var))
  if (!village_str %in% names(data)) village_str <- village_id_var
  
  # hamilton pps allocation
  village_alloc <- data %>%
    dplyr::count(
      dplyr::across(dplyr::any_of(c(county_str, village_str))),
      name = "village_size"
    ) %>%
    dplyr::left_join(allocation_df, by = county_str) %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(county_str))) %>%
    dplyr::mutate(
      county_total = sum(.data$village_size),
      raw_alloc    = .data[[alloc_size_var]] * .data$village_size / .data$county_total,
      sample_alloc = floor(.data$raw_alloc),
      remainder    = .data[[alloc_size_var]][1] - sum(floor(.data$raw_alloc)),
      frac         = .data$raw_alloc - floor(.data$raw_alloc),
      sample_alloc = .data$sample_alloc +
        ifelse(rank(-.data$frac, ties.method = "first") <= .data$remainder, 1L, 0L)
    ) %>%
    dplyr::ungroup()
  
  # minimum size redistribution
  village_alloc <- village_alloc %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(county_str))) %>%
    dplyr::group_modify(function(county_df, key) {
      
      alloc   <- county_df$sample_alloc
      
      # enforce the floor to be exactly min_village_sample capped at available population
      eff_min <- pmin(min_village_sample, county_df$village_size)
      
      deficit_idx <- which(alloc < eff_min)
      if (length(deficit_idx) == 0L) return(county_df)
      
      deficit        <- eff_min[deficit_idx] - alloc[deficit_idx]
      total_deficit  <- sum(deficit)
      surplus        <- pmax(alloc - eff_min, 0L)
      total_surplus  <- sum(surplus)
      
      if (total_surplus == 0L) {
        warning(
          "County '", key[[1]], "': not enough surplus to satisfy minimum of ",
          min_village_sample, " for all villages. Lifting as much as possible."
        )
      }
      
      donor_order <- order(-surplus)
      remaining   <- min(total_deficit, total_surplus)
      
      for (i in donor_order) {
        if (remaining <= 0L) break
        give      <- min(surplus[i], remaining)
        alloc[i]  <- alloc[i] - give
        remaining <- remaining - give
      }
      
      recovered <- min(total_deficit, total_surplus)
      for (j in seq_along(deficit_idx)) {
        if (recovered <= 0L) break
        give                  <- min(deficit[j], recovered)
        alloc[deficit_idx[j]] <- alloc[deficit_idx[j]] + give
        recovered             <- recovered - give
      }
      
      county_df$sample_alloc <- alloc
      county_df
    }) %>%
    dplyr::ungroup()
  
  # sample households within each village
  # join final target allocation boundaries back to full population dataset rows
  joined_data <- data %>%
    dplyr::left_join(
      dplyr::select(village_alloc, dplyr::any_of(c(county_str, village_str, "sample_alloc"))),
      by = c(county_str, village_str)
    )
  
  # shuffle and slice rows matching exact targets without using group modify
  sampled <- joined_data %>%
    dplyr::filter(!is.na(.data$sample_alloc) & .data$sample_alloc > 0L) %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(c(county_str, village_str)))) %>%
    dplyr::mutate(rand_draw_order = stats::runif(dplyr::n())) %>%
    dplyr::arrange(.data$rand_draw_order, .by_groups = TRUE) %>%
    dplyr::filter(dplyr::row_number() <= .data$sample_alloc) %>%
    dplyr::ungroup() %>%
    dplyr::select(-dplyr::any_of(c("sample_alloc", "rand_draw_order")))
  
  return(sampled)
}

```


## Stratified sampling

```{r function-draw_strs_sample}
#' Draw a Stratified Simple Random Sample of Households
#'
#' This function selects a stratified simple random sample of individual household 
#' rows from within counties. Every household within a given county has an equal 
#' probability of selection, independent of village size.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param allocation_df A data frame containing two columns: the county identifier 
#'   and the allocated household sample size for that county.
#' @param alloc_size_var The bare or quoted column name in `allocation_df` 
#'   indicating the target sample size.
#' @param seed An integer value used to initialize the pseudo-random number generator 
#'   for reproducible sampling. Defaults to `42`.
#'
#' @return A data frame containing the individually selected household rows.
#'
#' @importFrom dplyr group_by mutate ungroup inner_join arrange row_number filter across any_of desc n
#' @importFrom rlang sym
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw household data
#' raw_households <- data.frame(
#'   county_id = rep(c("County_A", "County_B"), each = 10),
#'   village_id = c(rep("V1", 3), rep("V2", 7), rep("V3", 4), rep("V4", 6)),
#'   household_id = paste0("HH_", 1:20)
#' )
#'
#' # household sample size targets per county
#' target_allocations <- data.frame(
#'   county_id = c("County_A", "County_B"),
#'   sample_size = c(2, 3)
#' )
#'
#' # run reproducible stratified simple random sampling
#' strs_sample <- draw_strs_sample(
#'   data = raw_households,
#'   county_var = "county_id",
#'   allocation_df = target_allocations,
#'   alloc_size_var = "sample_size",
#'   seed = 123
#' )
#' }
draw_strs_sample <- function(
    data,
    county_var = "county",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){

  set.seed(seed)

  joined_frame <- data %>%
    dplyr::inner_join(
      allocation_df,
      by = county_var
    )

  joined_frame %>%
    dplyr::group_by(
      dplyr::across(dplyr::any_of(county_var))
    ) %>%
    dplyr::slice_sample(
      n = min(dplyr::n(), first(.data[[alloc_size_var]]))
    ) %>%
    dplyr::ungroup()
}
```


## two stage cluster random sampling

```{r function-draw_cluster_sample}
#' Draw a Two-Stage Cluster Sample of Households
#'
#' This function performs two-stage cluster sampling within counties. In the 
#' first stage, a specified number of unique villages are selected randomly. In the 
#' second stage, a fixed number of households are randomly drawn from each 
#' of those selected villages. Both target sizes are pulled from the allocation frame.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param village_id_var The bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing columns for the county identifier, 
#'   the target village sample size, and the target household sample size per village.
#' @param alloc_villages_var The bare or quoted column name in `allocation_df` 
#'   indicating the number of villages to sample per county.
#' @param alloc_hh_var The bare or quoted column name in `allocation_df` 
#'   indicating the number of households to sample per chosen village.
#' @param seed An integer value used to initialize the pseudo-random number generator 
#'   for reproducible sampling. Defaults to `42`.
#'
#' @return A data frame containing the cluster-sampled household rows.
#'
#' @importFrom dplyr group_by mutate ungroup arrange row_number filter across any_of distinct n inner_join select
#' @importFrom rlang sym
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw household data
#' raw_households <- data.frame(
#'   county_id = rep(c("County_A", "County_B"), each = 15),
#'   village_id = c(rep("V1", 5), rep("V2", 5), rep("V3", 5), 
#'                  rep("V4", 5), rep("V5", 5), rep("V6", 5)),
#'   household_id = paste0("HH_", 1:30)
#' )
#'
#' # multi-stage target allocations dataframe
#' cluster_allocations <- data.frame(
#'   county_id = c("County_A", "County_B"),
#'   target_villages = c(2, 2),
#'   target_hh_per_village = c(3, 2)
#' )
#'
#' # run dynamic cluster sampling
#' cluster_sample <- draw_cluster_sample(
#'   data = raw_households,
#'   county_var = "county_id",
#'   village_id_var = "village_id",
#'   allocation_df = cluster_allocations,
#'   alloc_villages_var = "target_villages",
#'   alloc_hh_var = "target_hh_per_village",
#'   seed = 42
#' )
#' }
draw_cluster_sample <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_villages_var = "target_villages",
    alloc_hh_var = "target_hh_per_village",
    seed = 42
){

  set.seed(seed)

  selected_villages <- data %>%
    dplyr::distinct(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    dplyr::inner_join(
      allocation_df,
      by = county_var
    ) %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(county_var)
      )
    ) %>%
    dplyr::slice_sample(
      n = min(
        dplyr::n(),
        first(.data[[alloc_villages_var]])
      )
    ) %>%
    dplyr::ungroup()

  data %>%
    dplyr::inner_join(
      selected_villages,
      by = c(county_var, village_id_var)
    ) %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    dplyr::slice_sample(
      n = min(
        dplyr::n(),
        first(.data[[alloc_hh_var]])
      )
    ) %>%
    dplyr::ungroup()
}
```

## systematic sampling county level

```{r function-draw_systematic_across_villages}
#' Draw a Systematic Sample of Households Spreading Across Villages
#'
#' This function selects a systematic random sample of individual household rows 
#' from within counties. By sorting by village first, the systematic interval 
#' automatically spreads the selected sample evenly across the different villages.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param village_id_var The bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing two columns: the county identifier 
#'   and the allocated household sample size for that county.
#' @param alloc_size_var The bare or quoted column name in `allocation_df` 
#'   indicating the target sample size.
#' @param seed An integer value used to initialize the pseudo-random number generator. 
#'   Defaults to `42`.
#'
#' @return A data frame containing the systematically selected household rows.
#'
#' @importFrom dplyr group_by mutate ungroup inner_join arrange row_number filter across any_of n select
#' @importFrom rlang sym
#' @export
draw_systematic_across_villages <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){

  set.seed(seed)

  joined_frame <- data %>%
    dplyr::inner_join(
      allocation_df,
      by = county_var
    ) %>%
    dplyr::arrange(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    )

  joined_frame %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(county_var)
      )
    ) %>%
    dplyr::group_modify(~{

      df <- .x

      N <- nrow(df)
      n_target <- min(
        first(df[[alloc_size_var]]),
        N
      )

      k <- N / n_target

      start <- sample(
        seq_len(ceiling(k)),
        1
      )

      idx <- floor(
        start + (0:(n_target-1))*k
      )

      idx <- pmin(idx, N)

      df[idx, , drop = FALSE]

    }) %>%
    dplyr::ungroup()
}
```


## systematic sample village level allocation

```{r function-draw_systematic_within_villages}
#' Draw a Systematic Sample of Households Within Each Village
#'
#' This function applies systematic random sampling to households independently 
#' within each village cluster based on specific target allocations for each village.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param village_id_var The bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing columns for the village identifier 
#'   and its assigned household sample target.
#' @param alloc_size_var The bare or quoted column name in `allocation_df` 
#'   indicating the village-level target sample size.
#' @param seed An integer value used to initialize the pseudo-random number generator. 
#'   Defaults to `42`.
#'
#' @return A data frame containing the systematically selected household rows.
#'
#' @importFrom dplyr group_by mutate ungroup inner_join arrange row_number filter across any_of n select
#' @importFrom rlang sym
#' @export
draw_systematic_within_villages <- function(
    data,
    county_var = "county",
    village_id_var = "village",
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
){

  set.seed(seed)

  joined_frame <- data %>%
    dplyr::inner_join(
      allocation_df,
      by = village_id_var
    ) %>%
    dplyr::arrange(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    )

  joined_frame %>%
    dplyr::group_by(
      dplyr::across(
        dplyr::any_of(
          c(county_var, village_id_var)
        )
      )
    ) %>%
    dplyr::group_modify(~{

      df <- .x

      N <- nrow(df)

      n_target <- min(
        first(df[[alloc_size_var]]),
        N
      )

      k <- N / n_target

      start <- sample(
        seq_len(ceiling(k)),
        1
      )

      idx <- floor(
        start + (0:(n_target-1))*k
      )

      idx <- pmin(idx, N)

      df[idx, , drop = FALSE]

    }) %>%
    dplyr::ungroup()
}
```

## apply corrections

```{r function-apply_corrections}
#' Apply Data Cleaning Corrections Using an Embedded Cleaning Log
#'
#' This function updates a main dataset using a correction data frame. It applies
#' standard cell corrections and handles row deletions last. Deletions are processed 
#' using strict unique UUID matching if a true uuid column is present, falling back to 
#' exact name matching combined with multiple context validation columns if a general 
#' non-unique identifier or name-only tracking is used.
#'
#' @param data A data frame containing the main dataset to be cleaned.
#' @param id_var The bare or quoted column name representing the unique identifier.
#' @param corrections A data frame containing the corrections log along with the matching keys.
#' @param var_to_change_col Column name in `corrections` holding target names to change.
#' @param correct_value_col Column name in `corrections` holding updated values.
#' @param ben_name_col Column name in both datasets holding beneficiary names.
#' @param context_cols Optional character vector. Secondary column names present in both 
#'   datasets (e.g., `c("ben_id_number", "ward", "village")`) used to build a composite key 
#'   and avoid accidental collisions of identical or similar names.
#'
#' @return A data frame matching the structure of `data`.
#' @export
apply_corrections <- function(
    data, 
    id_var = "uuid", 
    corrections, 
    var_to_change_col = "var_to_change", 
    correct_value_col = "correct_value",
    ben_name_col = "ben_name",
    context_cols = c("ben_id_number", "ward", "village")
) {
  
  id_str <- as.character(substitute(id_var))
  if (!id_str %in% names(data)) id_str <- id_var
  
  var_change_str <- as.character(substitute(var_to_change_col))
  if (!var_change_str %in% names(corrections)) var_change_str <- var_to_change_col
  
  correct_val_str <- as.character(substitute(correct_value_col))
  if (!correct_val_str %in% names(corrections)) correct_val_str <- correct_value_col
  
  ben_name_str <- as.character(substitute(ben_name_col))
  if (!ben_name_str %in% names(corrections)) ben_name_str <- ben_name_col
  
  cleaned_data <- data
  
  # extract only the unique variables that actually exist in the main data
  distinct_vars <- corrections %>%
    dplyr::filter(!!rlang::sym(var_change_str) %in% names(cleaned_data)) %>%
    dplyr::pull(!!rlang::sym(var_change_str)) %>%
    unique()
  
  # 1 execute standard cell corrections first
  for (current_var in distinct_vars) {
    sub_corrections <- corrections %>%
      dplyr::filter(!!rlang::sym(var_change_str) == current_var) %>%
      dplyr::select(dplyr::any_of(c(ben_name_str, correct_val_str)))
    
    target_class <- class(cleaned_data[[current_var]])[1]
    
    if (target_class %in% c("numeric", "integer")) {
      sub_corrections[[correct_val_str]] <- as.numeric(sub_corrections[[correct_val_str]])
    } else if (target_class == "logical") {
      sub_corrections[[correct_val_str]] <- as.logical(sub_corrections[[correct_val_str]])
    } else {
      sub_corrections[[correct_val_str]] <- as.character(sub_corrections[[correct_val_str]])
    }
    
    join_target_name <- paste0("new_val_", current_var)
    sub_corrections <- dplyr::rename(
      sub_corrections, 
      !!rlang::sym(join_target_name) := !!rlang::sym(correct_val_str)
    )
    
    cleaned_data <- cleaned_data %>%
      dplyr::left_join(sub_corrections, by = ben_name_str) %>%
      dplyr::mutate(
        !!rlang::sym(current_var) := ifelse(
          !is.na(!!rlang::sym(join_target_name)), 
          !!rlang::sym(join_target_name), 
          !!rlang::sym(current_var)
        )
      ) %>%
      dplyr::select(-dplyr::any_of(join_target_name))
  }
  
  # 2 handle deletions as the final operation with compound keys
  if (var_change_str %in% names(corrections) && ben_name_str %in% names(corrections) && ben_name_str %in% names(data)) {
    
    # filter log down strictly to the rows designated for deletion using case insensitive grepl
    delete_log <- corrections %>%
      dplyr::filter(grepl("delete", !!rlang::sym(var_change_str), ignore.case = TRUE))
    
    if (nrow(delete_log) > 0) {
      
      # premium bypass: only use direct ID deletion if it is explicitly a unique uuid column
      if (tolower(id_str) == "uuid" && id_str %in% names(corrections) && id_str %in% names(data)) {
        
        delete_ids <- delete_log %>%
          dplyr::pull(!!rlang::sym(id_str)) %>%
          unique()
        
        cleaned_data <- cleaned_data %>%
          dplyr::filter(!(!!rlang::sym(id_str) %in% delete_ids))
        
      } else {
        # traditional fallback: filter context columns down to only those that actually exist in both datasets
        # this ensures that if id_str is "ben_id_number", it gets lumped safely into the compound validation checks
        all_potential_contexts <- unique(c(id_str, context_cols))
        valid_contexts <- all_potential_contexts[all_potential_contexts %in% names(data) & all_potential_contexts %in% names(corrections)]
        
        if (length(valid_contexts) > 0) {
          # combine beneficiary name and all valid context values into a multi-variable string key
          # using individual rowwise evaluations to avoid vector collapse traps
          build_composite_key <- function(df, name_col, ctx_cols) {
            df %>%
              dplyr::rowwise() %>%
              dplyr::mutate(
                comp_key = paste(
                  c(as.character(.data[[name_col]]), mget(ctx_cols)), 
                  collapse = "|||"
                )
              ) %>%
              dplyr::ungroup() %>%
              dplyr::pull(comp_key)
          }
          
          # generate keys for both datasets
          delete_keys <- build_composite_key(delete_log, ben_name_str, valid_contexts)
          
          # apply filter by checking the main data against the delete keys
          cleaned_data <- cleaned_data %>%
            dplyr::rowwise() %>%
            dplyr::filter(
              !paste(c(as.character(.data[[ben_name_str]]), mget(valid_contexts)), collapse = "|||") %in% delete_keys
            ) %>%
            dplyr::ungroup()
          
        } else {
          # fallback to strict exact name checking if no valid context variables were matched
          delete_names <- unique(delete_log[[ben_name_str]])
          cleaned_data <- cleaned_data %>%
            dplyr::filter(!(!!rlang::sym(ben_name_str) %in% delete_names))
        }
      }
    }
  }
  
  return(cleaned_data)
}

```

## read excel workbook

```{r function-read_workbook}
#' Read Excel sheets or an entire workbook into a list or data frame with clean names
#'
#' @description
#' Iterates through available worksheets within an Excel file or targets a specific sheet, 
#' extracts content while skipping a designated number of rows, and standardizes column names 
#' using snake_case. If the target evaluation resolves to exactly one worksheet, it automatically 
#' drops the list format and returns a single data frame directly.
#'
#' @param path Character string representing the path to the Excel file.
#' @param bind_rows Logical indicator. If TRUE, binds multiple extracted sheets into 
#'   a single unified data frame using a compiler identifier column. Defaults to FALSE.
#' @param id Character string naming the optional identifier column that tracks 
#'   the source sheet name when `bind_rows` is TRUE. Defaults to "sheet_name".
#' @param skip Integer specifying the number of rows to skip at the start of reading. Defaults to 0.
#' @param sheet_name Character string specifying a single sheet to read. If NULL, all sheets are processed. Defaults to NULL.
#'
#' @return A single data frame if only one sheet is processed or `bind_rows = TRUE`. 
#'   Otherwise, a named list of data frames representing individual sheets.
#'
#' @importFrom readxl excel_sheets read_excel
#' @importFrom purrr set_names map list_rbind
#' @importFrom janitor clean_names
#'
#' @export
#'
#' @examples
#' \dontrun{
#' # read a specific sheet and skip the first 2 rows
#' df_raw <- read_workbook("survey_data.xlsx", sheet_name = "Main_Data", skip = 2)
#' }
read_workbook <- function(path, bind_rows = FALSE, id = "sheet_name", skip = 0, sheet_name = NULL) {
  
  # determine target sheet names based on user input selection
  if (!is.null(sheet_name)) {
    sheet_names <- sheet_name
  } else {
    sheet_names <- readxl::excel_sheets(path)
  }
  
  # short circuit and return a single data frame instantly if only one sheet is target evaluated
  if (length(sheet_names) == 1) {
    single_df <- readxl::read_excel(path = path, sheet = sheet_names, skip = skip) %>%
      janitor::clean_names()
    return(single_df)
  }
  
  # extract sheet names, read contents, and standardize variables to snake case
  sheets_list <- sheet_names %>%
    purrr::set_names() %>%
    purrr::map(~ {
      readxl::read_excel(path = path, sheet = .x, skip = skip) %>%
        janitor::clean_names()
    })
  
  # conditionally collapse the list rows into a single data frame if specified
  if (bind_rows) {
    combined_data <- purrr::list_rbind(sheets_list, names_to = id)
    return(combined_data)
  }
  
  return(sheets_list)
}
```


## standardize names

```{r function-standardize_names}
#' Map and Standardize Administrative Names to Reference Values
#'
#' This function takes a dataset containing entered administrative names (such as
#' sublocations or villages) and cross-references them against an official lookup
#' table. It matches names using an optimal string alignment distance algorithm
#' to automatically correct typos, casing differences, and minor variations.
#'
#' @param data A data frame containing the main survey or household records.
#' @param match_col A bare or quoted column name in \code{data} containing the 
#'   entered names that need correction.
#' @param ref_df A data frame serving as the official reference lookup table.
#' @param ref_col A bare or quoted column name in \code{ref_df} containing the
#'   true, standardized reference names.
#' @param max_dist A numeric value specifying the maximum allowed string distance
#'   for a match to be considered valid. Greater values allow more generous 
#'   typo corrections. Defaults to \code{2}.
#'
#' @return A data frame with the same rows as \code{data}, where the original
#'   \code{match_col} values are replaced by the true values from \code{ref_col}.
#'   Unmatched records will retain their original values or return \code{NA} if 
#'   completely outside the distance threshold.
#'
#' @importFrom dplyr mutate row_number across any_of
#' @importFrom rlang .data
#' @importFrom stringdist amatch
#'
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw survey data with typos
#' survey_records <- data.frame(
#'   household_id = 1:4,
#'   entered_sublocation = c("Muguga ", "Kikuyu", "sigona", "UnknownPlace")
#' )
#'
#' # true official reference registry
#' official_sublocations <- data.frame(
#'   true_name = c("Muguga", "Kikuyu", "Sigona", "Karai")
#' )
#'
#' # clean and standardize the sublocation entries
#' cleaned_data <- standardize_names(
#'   data      = survey_records,
#'   match_col = "entered_sublocation",
#'   ref_df    = official_sublocations,
#'   ref_col   = "true_name",
#'   max_dist  = 3
#' )
#' }
standardize_names <- function(
    data,
    match_col,
    ref_df,
    ref_col,
    max_dist = 2
) {
  
  # convert input column names safely to standard strings
  match_str <- as.character(substitute(match_col))
  if (!match_str %in% names(data)) match_str <- match_col
  
  ref_str <- as.character(substitute(ref_col))
  if (!ref_str %in% names(ref_df)) ref_str <- ref_col
  
  # extract the target reference vector clean of missing entries
  ref_vector <- as.character(na.omit(ref_df[[ref_str]]))
  
  # extract and clean input entries for robust string matching
  raw_entries <- as.character(data[[match_str]])
  clean_raw   <- trimws(tolower(raw_entries))
  clean_ref   <- trimws(tolower(ref_vector))
  
  # find matching vector indices based on optimal string alignment distance
  match_indices <- stringdist::amatch(
    x          = clean_raw,
    table      = clean_ref,
    method     = "osa",
    maxDist    = max_dist
  )
  
  # map matched indices back to original true casing from reference list
  standardized_values <- ifelse(
    !is.na(match_indices),
    ref_vector[match_indices],
    raw_entries
  )
  
  # assign corrected values back to the original target column mask
  standardized_df <- data %>%
    dplyr::mutate(
      dplyr::across(
        dplyr::any_of(match_str),
        ~ standardized_values
      )
    )
  
  return(standardized_df)
}
```


## find name matches

```{r function-find_name_matches}
#' Identify Potential Beneficiary Name Matches From a Master Dataset
#'
#' @description
#' Compares a column of names in a flagged dataset against a column of names in a 
#' master reference dataset using optimized vectorized string distance matrices. 
#' If a potential match is identified within a specified text edit threshold, 
#' the candidate name from the master dataset is mapped into a new column named `name_possible`.
#'
#' @param flagged_data A data frame containing the flagged records needing verification.
#' @param master_data A data frame containing the verified master benchmark records.
#' @param flagged_col The bare or quoted column name of the name column in `flagged_data`. Defaults to "ben_name".
#' @param master_col The bare or quoted column name of the name column in `master_data`. Defaults to "ben_name".
#' @param max_dist Integer specifying the maximum allowed string distance/edits to qualify as a match. Defaults to 3.
#'
#' @return A data frame matching the structure of `flagged_data`, with an additional 
#'   `name_possible` character column containing the matched master dataset name (or NA if no match is found).
#'
#' @importFrom dplyr mutate left_join select any_of tibble
#' @importFrom utils adist
#' @importFrom rlang sym !! :=
#' @importFrom stringdist stringdistmatrix
#'
#' @export
#'
#' @examples
#' \dontrun{
#' verified_df <- find_name_matches(bad_data, good_data, max_dist = 2)
#' }


find_name_matches <- function(
    flagged_data,
    master_data,
    flagged_col = "ben_name",
    master_col  = "ben_name",
    max_dist    = 3
) {
  flag_str <- as.character(substitute(flagged_col))
  if (!flag_str %in% names(flagged_data)) flag_str <- flagged_col
  
  master_str <- as.character(substitute(master_col))
  if (!master_str %in% names(master_data)) master_str <- master_col
  
  master_pool <- unique(na.omit(as.character(master_data[[master_str]])))
  flagged_vec <- as.character(flagged_data[[flag_str]])
  
  sort_tokens <- function(x) {
    vapply(
      strsplit(tolower(trimws(x)), "[[:space:]\\-]+"),
      function(toks) paste(sort(toks), collapse = " "),
      character(1)
    )
  }
  
  flagged_lower <- tolower(flagged_vec)
  master_lower  <- tolower(master_pool)
  flagged_tok   <- sort_tokens(flagged_vec)
  master_tok    <- sort_tokens(master_pool)
  
  # blocking key: first character of the first token
  flagged_block <- substr(flagged_tok, 1, 1)
  master_block  <- substr(master_tok,  1, 1)
  
  name_possible <- rep(NA_character_, length(flagged_vec))
  
  for (block in unique(flagged_block)) {
    fi <- which(flagged_block == block)   # row indices in flagged
    mi <- which(master_block  == block)   # candidate indices in master
    
    if (length(fi) == 0 || length(mi) == 0) next
    
    dist_whole <- stringdist::stringdistmatrix(
      flagged_lower[fi], master_lower[mi],
      method = "osa", nthread = parallel::detectCores()
    )
    dist_token <- stringdist::stringdistmatrix(
      flagged_tok[fi], master_tok[mi],
      method = "osa", nthread = parallel::detectCores()
    )
    dist_best <- pmin(dist_whole, dist_token)
    
    best_col  <- max.col(-dist_best, ties.method = "first")
    best_dist <- dist_best[cbind(seq_along(fi), best_col)]
    
    matched <- best_dist <= max_dist
    name_possible[fi[matched]] <- master_pool[mi[best_col[matched]]]
  }
  
  flagged_data[["name_possible"]] <- name_possible
  flagged_data
}

```


## match county records

```{r function-match_county_records}
#' Filter and Match Payroll Records Against Registration Data by County
#'
#' @description
#' Subsets registration data for a specific county, removes duplicate IDs,
#' joins the results with payroll records, and flags rows where the payroll name
#' is a partial match inside the registration name string.
#'
#' @param registration A data frame containing the primary registration benchmark records.
#' @param payroll A data frame containing the target payroll records to verify.
#' @param county_name Character string specifying the county name to filter on. Defaults to "Tana".
#'
#' @return A data frame matching the layout of the payroll dataset, appended with 
#'   columns from the subsetted registration data and a logical `names_match` flag.
#'
#' @importFrom dplyr filter select distinct left_join mutate
#' @importFrom stringi stri_detect_fixed
#'
#' @export
#' @examples
#' \dontrun{match_county_records()}
match_county_records <- function(registration, payroll, county_name = "Tana") {
  # filter registration pool for the target county and drop duplicate id records
  df_county_ref <- registration %>%
    dplyr::filter(grepl(county_name, county, ignore.case = TRUE)) %>%
    dplyr::select(ben_name, reg_num, ben_id_number) %>%
    dplyr::distinct(ben_id_number, .keep_all = TRUE)
  
  # merge payroll records with the clean county reference subset
  df_joined <- payroll %>%
    dplyr::left_join(
      df_county_ref,
      by = c("id_number" = "ben_id_number")
    ) %>%
    # remove residual system export columns starting with x or X
    dplyr::select(-grep("^[xX]", names(.)))
  
  # fast vectorized string detection — no r loops or nested scans
  df_joined %>%
    dplyr::mutate(
      names_match = !is.na(name_payroll) & 
        !is.na(ben_name) & 
        name_payroll != "" & 
        stringi::stri_detect_fixed(ben_name, name_payroll, case_insensitive = TRUE)
    )
}
```


## normalize names

```{r function-normalize_names}
#' Normalize a Name Vector
#'
#' @description
#' Cleans a character vector of names by collapsing extra whitespace,
#' trimming leading/trailing spaces, and converting to title case.
#'
#' @param x A character vector of names.
#'
#' @return A cleaned character vector.
#'
#' @importFrom stringr str_squish str_to_title
#'
#' @export
#' @examples
#' \dontrun{normalize_names()}
normalize_names <- function(x) {
  x <- stringr::str_squish(x)       # trims + collapses internal spaces
  x <- stringr::str_to_title(x)     # title case
  x
}
```


## Veification criteria table

```{r function-gen_verification_criteria_table}
#' Create the target beneficiary verification criteria and indicators summary
#'
#' @description
#' Builds a two column matrix pairing each humanitarian vulnerability 
#' criterion alongside its designated operational indicator visually 
#' merging repeating criteria rows and introducing complete outer frame grids 
#' that fully enclose column headers with explicit bold top and bottom lines
#'
#' @param df dataframe of verification criteria
#' @param title character vector of title
#' @return A styled flextable object containing the merged two column layout with 
#' fully enclosed bold grid line formatting
#'
#' @importFrom tibble tibble
#' @importFrom dplyr mutate lag
#' @importFrom flextable flextable as_paragraph as_chunk align valign width border_outer border_inner_h border_inner_v bg color bold font set_caption delete_part add_footer_lines
#' @importFrom officer fp_border fp_text
#'
#' @export
#' @examples
#' \dontrun{gen_verification_criteria_table()}
gen_verification_criteria_table <- function(
    df = verification_criteria,
    title = "Vulnerability Criteria"
    ) {
  
  # track total row count
  total_rows <- nrow(df)
  
  # track the indices where a new criteria block begins for main horizontal lines
  main_grid_rows <- c(7, 12, 15, 16)
  
  # define the explicit border properties using officer
  heavy_border <- officer::fp_border(color = "#0B1054", width = 2.5, style = "solid")
  mid_border   <- officer::fp_border(color = "#0B1054", width = 2, style = "solid")
  thin_border  <- officer::fp_border(color = "#d3d3d3", width = 1, style = "solid")
  
  # build the flextable layout structure
  styled_table <- flextable::flextable(df) %>%
    # visually merge repeating cell items sequentially along the vertical axis
    flextable::merge_v(j = ~ Criteria) %>%
    # map alignments and set relative layout sizing to fit standard pages
    flextable::align(align = "left", part = "all") %>%
    flextable::valign(valign = "top", part = "body") %>%
    flextable::width(j = "Criteria", width = 2.2) %>%
    flextable::width(j = "Indicator", width = 4.3) %>%
    
    # apply deep blue background fill with white bold text for column labels
    flextable::bg(bg = "#0B1054", part = "header") %>%
    flextable::color(color = "white", part = "header") %>%
    flextable::bold(part = "header") %>%
    
    # apply a standard baseline horizontal grid line separating every single indicator row
    flextable::border_inner_h(border = thin_border, part = "body") %>%
    
    # overlay bold main horizontal block dividers within the body row structures
    flextable::hline(i = (main_grid_rows - 1), border = mid_border, part = "body") %>%
    
    # construct complete outer frame borders and center divider line enclosing headers and rows
    flextable::border_outer(border = heavy_border, part = "all") %>%
    flextable::border_inner_v(border = mid_border, part = "all") %>%
    
    # add font settings and map the source footnote notes cleanly to the bottom layout
    flextable::font(fontname = "Arial", part = "all") %>%
    flextable::add_footer_lines(
      values = "Acronym Definitions: OTP - Outpatient Therapeutic Program; SFP - Supplementary Feeding Program; FCS - Food Consumption Score; HHs - Households"
    ) %>%
    flextable::font(fontname = "Arial", part = "footer")
  
  return(styled_table)
}
```

## get vulnerability criteria variables

```{r function-get_vulnerability_data}

#' get vulnerability criteria (physiological, nutritional etc)
#'
#' @returns a dataframe
#' @importFrom dplyr filter mutate select lead
#' @export
#'
#' @examples
#' \dontrun{get_vulnerability_data()}
get_vulnerability_data <- function(){
  df_dict <- acheck::survey %>%
    filter(!grepl("note", type)) %>%
    mutate(n = dplyr::row_number())
  
  vul_rows <- df_dict %>% 
    # filter(grepl("^physio|^social|^nutri|^food|gps", label, ignore.case = T)) %>%
    filter(
      grepl(paste(paste0("^e_", 2:6), collapse = "|"), name)
    ) %>%
    bind_rows(
      
      vul_rows <- df_dict %>% 
        # filter(grepl("^physio|^social|^nutri|^food|gps", label, ignore.case = T)) %>%
        filter(
          grepl("^GPS", label)
        )
    ) %>%
    select(type, name, label, n) %>%
    mutate(n_end = lead(n),
           n = n + 1) %>%
    filter(!grepl("geo", type)) %>%
    # mutate(
    #   n_end = case_when(
    #     label == "Food consumption score" ~ 257,
    #     label == "Nutritional vulnerability" ~ 265,
    #     TRUE ~ n_end
    #   )
    # ) %>%
    filter(! n == n_end) %>%
    filter(
      grepl("^e", name)
    )
  vul_rows
}

```

## automate indicators

```{r function-automate_indicators}
#' Automate binary indicator variable creation
#'
#' This function dynamically scans input columns, handles missing data structures,
#' cleans raw count variables, and generates binary 0/1 indicators based on 
#' underlying numeric and character type properties, preserving the original name.
#'
#' @param df A dataframe or tibble object.
#' @param cols A character vector specifying the target column names to process.
#'
#' @return A modified dataframe with updated raw counts and appended indicator columns.
#' @importFrom dplyr mutate across all_of if_else
#' @importFrom stringr str_detect
#' @export
automate_indicators <- function(df, cols) {
  df %>%
    # first pass clean all raw numeric count inputs to handle nas and zeros universally
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(cols),
        function(x) {
          # lower case comments without dots or dashes
          if (is.numeric(x)) {
            return(dplyr::if_else(is.na(x) | x == 0, 0, x))
          }
          return(x)
        }
      )
    ) %>%
    # second pass generate indicators and append ind cleanly to the original column names
    dplyr::mutate(
      dplyr::across(
        dplyr::all_of(cols), 
        function(x) {
          if (is.numeric(x)) {
            indicator <- dplyr::if_else(x == 0, 0, 1)
            return(indicator)
          } 
          
          indicator <- dplyr::if_else(
            stringr::str_detect(tolower(x), "no") | is.na(x), 
            0, 
            1
          )
          return(indicator)
        }, 
        .names = "{.col}_ind"
      )
    )
}
```

## donut plot

```{r function-plot_parts_donut}
#' visualize parts to whole using a donut layout framework
#'
#' @param data dataframe
#' @param fill_col character. the categorical column to slice
#' @param palette character vector. hex color values
#'
#' @returns a ggplot object
#' @importFrom dplyr count mutate across all_of
#' @importFrom ggplot2 ggplot aes geom_col geom_text scale_fill_manual theme_void theme coord_polar xlim position_stack
#' @importFrom scales percent
#' @export
#' @examples
#' \dontrun{plot_parts_donut()}
plot_parts_donut <- function(data, fill_col, palette = c("#81C784", "#FFB74D", "#E65100")) {
  
  # lower case comments without dots or dashes
  # map labels and compute dynamic cumulative coordinates for clean radial centering
  plot_df <- data %>%
    dplyr::count(dplyr::across(dplyr::all_of(fill_col))) %>%
    dplyr::mutate(
      pct = n / sum(n),
      pct_label = paste0(
        # .data[[fill_col]], 
        # "\n(", 
        scales::percent(pct, accuracy = 1)
        # ,")"
      )
    )
  
  ggplot2::ggplot(plot_df, ggplot2::aes(x = 3, y = pct, fill = .data[[fill_col]])) +
    # maintain stable donut ring layout thickness
    ggplot2::geom_col(width = 1, color = "white", size = 0.5) +
    # shift text position horizontally outward beyond the main donut frame
    ggplot2::geom_text(
      ggplot2::aes(x = 4.2, label = pct_label),
      position = ggplot2::position_stack(vjust = 0.5),
      color = "black",
      fontface = "bold",
      size = 3.5
    ) +
    ggplot2::scale_fill_manual(values = palette) +
    ggplot2::coord_polar(theta = "y") +
    # expand the coordinate map boundary to give labels complete breathing room
    ggplot2::xlim(c(1, 5.2)) +
    ggplot2::theme_void() +
    ggplot2::theme(
      legend.position = "bottom"
    )
}
```

## faceted donuts

```{r function-plot_faceted_donuts}
#' visualize parts to whole using an automated faceted donut grid via patchwork
#'
#' @param data dataframe
#' @param fill_col character the categorical variable to slice 
#' @param facet_col character the variable to split panels by 
#' @param palette character vector hex color values mapping to categories
#'
#' @returns a patchwork grid assembly of donut plots
#' @importFrom dplyr filter count mutate pull
#' @importFrom ggplot2 ggplot aes geom_col geom_text scale_fill_manual theme_void theme coord_polar xlim labs element_text position_stack
#' @importFrom patchwork wrap_plots plot_layout
#' @importFrom stringr str_to_title
#' @export
#' @examples
#' \dontrun{plot_faceted_donuts()}
plot_faceted_donuts <- function(data, 
                                fill_col, 
                                facet_col, 
                                palette = c("#388E3C", "#F57C00", "#D32F2F")) {
  
  # extract unique facet groups while ignoring missing rows
  facet_groups <- data %>%
    dplyr::filter(!is.na(.data[[facet_col]])) %>%
    dplyr::pull(.data[[facet_col]]) %>%
    unique() %>%
    sort()
  
  # dynamically determine optimal column layouts based on category lengths
  n_facets <- length(facet_groups)
  calculated_ncol <- if (n_facets <= 3) {
    n_facets
  } else if (n_facets == 4) {
    2
  } else {
    3
  }
  
  # generate individual donut objects iteratively
  plot_list <- lapply(facet_groups, function(group) {
    
    # filter down to single panel subset
    sub_df <- data %>% 
      dplyr::filter(.data[[facet_col]] == group, !is.na(.data[[fill_col]]))
    
    # safeguard against empty group metrics
    if (nrow(sub_df) == 0) return(NULL)
    
    plot_df <- sub_df %>%
      dplyr::count(dplyr::across(dplyr::all_of(fill_col))) %>%
      dplyr::mutate(
        pct = n / sum(n),
        pct_label = paste0(
          scales::percent(pct, accuracy = 1)
        )
      )
    
    ggplot2::ggplot(plot_df, ggplot2::aes(x = 3, y = pct, fill = .data[[fill_col]])) +
      # maintain donut thickness between inner and outer radius limits
      ggplot2::geom_col(width = 1, color = "white", size = 0.5) +
      # push text radius position out beyond the chart area boundary
      ggplot2::geom_text(
        ggplot2::aes(x = 4.2, label = pct_label),
        position = ggplot2::position_stack(vjust = 0.5),
        color = "black",
        fontface = "bold",
        size = 2.8
      ) +
      ggplot2::scale_fill_manual(values = palette) +
      ggplot2::coord_polar(theta = "y") +
      # expand structural coordinate canvas limits to prevent text clipping
      ggplot2::xlim(c(1, 5.2)) +
      ggplot2::labs(title = stringr::str_to_title(group)) +
      ggplot2::theme_void() +
      ggplot2::theme(
        # removed legend position here to let patchwork handle it
        plot.title = ggplot2::element_text(hjust = 0.5, face = "bold", size = 11, vjust = -1)
      )
  })
  
  # drop null items from tracking list
  # using base r filter option to keep namespaces minimal
  plot_list <- plot_list[!sapply(plot_list, is.null)]
  
  # stitch panels together and collect legends into a single unified guide
  combined_grid <- patchwork::wrap_plots(plot_list, ncol = calculated_ncol) +
    patchwork::plot_layout(guides = "collect") &
    ggplot2::theme(legend.position = "bottom")
  
  return(combined_grid)
}
```

## theme plots

```{r function-theme_plots}
#' publication ready theme wrapper using cowplot with gridlines
#'
#' @param base_size numeric base font size
#' @param base_family character font family
#'
#' @returns a ggplot2 theme object
#' @importFrom cowplot theme_minimal_grid
#' @importFrom ggplot2 theme element_text rel margin
#' @export
theme_plots <- function(base_size = 11, base_family = "sans") {
  
  # leverage cowplot minimal grid as the structural foundation
  cowplot::theme_minimal_grid(
    base_size = base_size,
    base_family = base_family,
    rel_title = 1.25
  ) +
    # fine tune alignment and margins
    ggplot2::theme(
      plot.title = ggplot2::element_text(
        face = "bold",
        hjust = 0.5,
        margin = ggplot2::margin(b = 10)
      ),
      plot.subtitle = ggplot2::element_text(
        hjust = 0.5,
        color = "grey30",
        face = "plain",
        margin = ggplot2::margin(b = 8)
      ),
      legend.position = "bottom"
    )
}
```

## save plots as .png

```{r function-save_plots_to_png}
#' save a list of active plots into a grid using the png device
#'
#' @param plot_list list of ggplot or grid objects
#' @param output_path character path where the combined grid will be saved
#' @param ncol integer number of columns in the grid
#' @param width integer width of output image in pixels
#' @param height integer height of output image in pixels
#' @param res integer nominal resolution in ppi
#'
#' @return invisible null saves file directly to disk
#' @importFrom grDevices png dev.off
#' @importFrom grid grid.newpage pushViewport viewport grid.layout
#' @export
save_plots_to_png <- function(plot_list, 
                                   output_path, 
                                   ncol = 2, 
                                   width = 1600, 
                                   height = 1200, 
                                   res = 144) {
  
  # calculate layout grid dimensions based on total plot count
  n_plots <- length(plot_list)
  nrow <- ceiling(n_plots / ncol)
  
  # open the native png graphics device to catch the upcoming draw calls
  grDevices::png(
    filename = output_path, 
    width = width, 
    height = height, 
    res = res
  )
  
  # initialize a fresh blank grid canvas page
  grid::grid.newpage()
  
  # push the base grid layout matrix structure onto the canvas
  grid::pushViewport(
    grid::viewport(layout = grid::grid.layout(nrow, ncol))
  )
  
  # loop through every plot object in the list and assign coordinates
  for (i in seq_along(plot_list)) {
    row_idx <- ceiling(i / ncol)
    col_idx <- (i - 1) %% ncol + 1
    
    # render the plot directly into its designated grid cell coordinates
    print(
      plot_list[[i]], 
      vp = grid::viewport(layout.pos.row = row_idx, layout.pos.col = col_idx)
    )
  }
  
  # shut down the active device to commit and write the file to disk
  grDevices::dev.off()
}
```

## household points static

```{r function-plot_household}
#' plot household density by clustering household level coordinates into villages
#'
#' @param data dataframe containing household level records
#' @param lat_col character name of the latitude column
#' @param lon_col character name of the longitude column
#' @param village_col character name of the village column to group by
#' @param county_col character optional name of the county column for coloring
#'
#' @returns a ggplot object
#' @importFrom dplyr filter group_by across all_of summarise n ungroup
#' @importFrom ggplot2 ggplot aes geom_point coord_equal scale_size_continuous scale_color_viridis_d labs theme_minimal theme element_text
#' @importFrom stringr str_to_title
#' @export
plot_household <- function(data, lat_col, lon_col, village_col, county_col = NULL) {
  
  # build grouping vector based on inputs
  group_vars <- village_col
  if (!is.null(county_col)) {
    group_vars <- c(county_col, group_vars)
  }
  
  # aggregate household coordinates to village level centers and counts
  plot_df <- data %>%
    dplyr::filter(!is.na(.data[[lat_col]]), !is.na(.data[[lon_col]])) %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(group_vars))) %>%
    dplyr::summarise(
      mean_lon = mean(.data[[lon_col]], na.rm = TRUE),
      mean_lat = mean(.data[[lat_col]], na.rm = TRUE),
      household_count = dplyr::n(),
      .groups = "drop"
    )
  
  # map the village center coordinates and bind size to household counts
  p <- ggplot2::ggplot(
    plot_df, 
    ggplot2::aes(
      x = mean_lon, 
      y = mean_lat, 
      size = household_count
    )
  )
  
  # apply color grouping if a county column is provided
  if (!is.null(county_col)) {
    p <- p + ggplot2::geom_point(ggplot2::aes(color = .data[[county_col]]), alpha = 0.7) +
      ggplot2::scale_color_viridis_d(option = "plasma", end = 0.85)
  } else {
    p <- p + ggplot2::geom_point(color = "#0073C2", alpha = 0.6)
  }
  
  # format layout grids and text descriptions
  p <- p +
    # enforce equal aspect ratio so kenya geography looks correct
    ggplot2::coord_equal() +
    # scale bubble sizes dynamically so larger villages pop out
    ggplot2::scale_size_continuous(range = c(3, 15)) +
    ggplot2::labs(
      x = "Longitude",
      y = "Latitude",
      size = "Households per Village",
      color = stringr::str_to_title(county_col)
    ) +
    ggplot2::theme_minimal() +
    ggplot2::theme(
      plot.title = ggplot2::element_text(face = "bold", hjust = 0.5, size = 13),
      legend.position = "bottom",
      legend.box = "vertical"
    )
  
  return(p)
}
```


## household plot interactive

```{r function-plot_household_leaflet}
#' generate an interactive leaflet map of household clusters
#'
#' @param data dataframe containing household level records
#' @param lat_col character name of the latitude column
#' @param lon_col character name of the longitude column
#' @param village_col character name of the village column
#' @param county_col character optional name of the county column
#'
#' @returns a leaflet map object
#' @importFrom dplyr filter group_by across all_of summarise n ungroup
#' @importFrom leaflet leaflet addProviderTiles addCircleMarkers colorFactor addLegend providers labelOptions
#' @importFrom htmltools HTML
#' @export
plot_household_leaflet <- function(data, lat_col, lon_col, village_col, county_col = NULL) {
  
  # build grouping vector based on inputs
  group_vars <- village_col
  if (!is.null(county_col)) {
    group_vars <- c(county_col, group_vars)
  }
  
  # aggregate household coordinates to village level centers and counts
  plot_df <- data %>%
    dplyr::filter(!is.na(.data[[lat_col]]), !is.na(.data[[lon_col]])) %>%
    dplyr::group_by(dplyr::across(dplyr::all_of(group_vars))) %>%
    dplyr::summarise(
      mean_lon = mean(.data[[lon_col]], na.rm = TRUE),
      mean_lat = mean(.data[[lat_col]], na.rm = TRUE),
      household_count = dplyr::n(),
      .groups = "drop"
    )
  
  # map values to a color column explicitly before entering leaflet
  if (!is.null(county_col)) {
    pal <- leaflet::colorFactor(palette = "viridis", domain = plot_df[[county_col]])
    plot_df$calculated_color <- pal(plot_df[[county_col]])
  } else {
    plot_df$calculated_color <- "#0073C2"
  }
  
  # configure mathematical radius translation in pixel limits
  max_count <- max(plot_df$household_count, na.rm = TRUE)
  min_count <- min(plot_df$household_count, na.rm = TRUE)
  
  get_radius <- function(count) {
    if (max_count == min_count) return(8)
    5 + 25 * (count - min_count) / (max_count - min_count)
  }
  
  # compile clean html strings for hover map popups
  labels <- sprintf(
    "<strong>Village:</strong> %s<br/><strong>Households:</strong> %d",
    plot_df[[village_col]],
    plot_df$household_count
  )
  if (!is.null(county_col)) {
    labels <- sprintf("<strong>County:</strong> %s<br/>%s", plot_df[[county_col]], labels)
  }
  html_labels <- lapply(labels, htmltools::HTML)
  
  # construct interactive spatial visualization canvas
  map <- leaflet::leaflet(plot_df) %>%
    leaflet::addProviderTiles(leaflet::providers$CartoDB.Positron) %>%
    leaflet::addCircleMarkers(
      lng = ~mean_lon,
      lat = ~mean_lat,
      radius = ~get_radius(household_count),
      color = ~calculated_color,
      stroke = TRUE,
      weight = 1,
      fillOpacity = 0.6,
      label = html_labels,
      labelOptions = leaflet::labelOptions(
        style = list("font-weight" = "normal", padding = "4px 8px"),
        textsize = "13px",
        direction = "auto"
      )
    )
  
  # attach legend module if grouped by county layers
  if (!is.null(county_col)) {
    map <- map %>%
      leaflet::addLegend(
        pal = pal,
        values = plot_df[[county_col]],
        opacity = 0.7,
        title = "County",
        position = "bottomright"
      )
  }
  
  return(map)
}
```

<!--
There can be other functions, examples and tests in your flat template.
Each of them will be inflated in a different file, provided that there is a level-1 or level-2 section title to separate from previous functions.
-->



<!--
# There can be development actions

Create a chunk with 'development' actions

- The chunk needs to be named `development` or `dev`
- It contains functions that are used for package development only
- Note that you may want to store most of these functions in the 0-dev_history.Rmd file

These are only included in the present flat template file, their content will not be part of the package anywhere else.
-->

```{r development-inflate, eval=FALSE}
# Keep eval=FALSE to avoid infinite loop in case you hit the knit button
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_full.Rmd", vignette_name = "Get started")
```


# Inflate your package

You're one inflate from paper to box.
Build your package from this very Rmd using `fusen::inflate()`

- Verify your `"DESCRIPTION"` file has been updated
- Verify your function is in `"R/"` directory
- Verify your test is in `"tests/testthat/"` directory
- Verify this Rmd appears in `"vignettes/"` directory
