---
title: "Data checks"
output: html_document
editor_options: 
  chunk_output_type: console
---

<!-- Run this 'development' chunk -->
<!-- Store every call to library() that you need to explore your functions -->

```{r development, include=FALSE}
library(testthat)
library(roxygen2)
```

<!--
 You need to run the 'description' chunk in the '0-dev_history.Rmd' file before continuing your code there.

If it is the first time you use {fusen}, after 'description', you can directly run the last chunk of the present file with inflate() inside.
--> 

```{r development-load}
# Load already included functions if relevant
pkgload::load_all(export_all = FALSE)
```

# Include some data examples in your package

<!-- 
 Store your dataset in a directory named "inst/" at the root of your project.
 Use it for your tests in this Rmd thanks to `pkgload::load_all()` to make it available
and `system.file()` to read it in your examples.

- There already is a dataset in the "inst/" directory to be used in the examples below
-->

```{r development-dataset}
# Run all this chunk in the console directly
# There already is a dataset in the "inst/" directory
# Make the dataset file available to the current Rmd during development
pkgload::load_all(path = here::here(), export_all = FALSE)

# You will be able to read your example data file in each of your function examples and tests as follows - see chunks below
# datafile <- system.file("nyc_squirrels_sample.csv", package = "acheck")
# nyc_squirrels <- read.csv(datafile, encoding = "UTF-8")
# 
# nyc_squirrels
```


## Add gtsummary object to word

<!--
Create a chunk for the core of the function

- The chunk needs to be named `function` at least
- It contains the code of a documented function
- The chunk can also be named `function-my_median` to make it easily
findable in your Rmd
- Let the `@examples` part empty, and use the next `examples` chunk instead to present reproducible examples

After inflating the template

-  This function code will automatically be added in a new file in the "R/" directory
-->

```{r development-function}
# Prepare the code of your function here
```

```{r function-add_gt_to_word}
format_num <- gtsummary::label_style_number()

#' Add a gtsummary objet to word doc
#'
#' @param doc word doc created as `officer::read_docx()`
#' @param gt_obj gtsummary object
#' @param title character header
#' @param landscape logical. page orientation
#' @importFrom flextable  body_add_flextable
#' @importFrom officer body_add_par body_add_break body_end_section_landscape
#'
#' @returns a word doc
#' @export
#'
#' @examples
#' \dontrun{
#' add_gt_to_word(doc, gt_obj, title, landscape = FALSE)
#' }
add_gt_to_word <- function(doc, gt_obj, title, landscape = FALSE) {
  # Convert gtsummary to flextable
  ft <- gtsummary::as_flex_table(gt_obj) %>%
    flextable::autofit()
  
  # Add title and table
  doc <- doc %>%
    body_add_par(title, style = "heading 2") %>%
    body_add_flextable(ft) %>%
    body_add_par("") # spacing
  
  # Handle Orientation and Page Breaks
  if (landscape) {
    # Close the current landscape section
    doc <- doc %>% body_end_section_landscape() 
  } else {
    # If not landscape, just add a standard page break
    doc <- doc %>% body_add_break()
  }
  
  return(doc)
}

```



<!--
Create a chunk with an example of use for your function

- The chunk needs to be named `examples` at least
- It contains working examples of your function
- The chunk is better be named `examples-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This example will automatically be added in the '@examples' part of our function above in the "R/" directory
- This example will automatically be added in the vignette created from this Rmd template
-->


<!--
Create a chunk with a test of use for your function

- The chunk needs to be named `tests` at least
- It contains working tests of your function
- The chunk is better be named `tests-my_median` to be handled
correctly when inflated as a vignette

After inflating the template

-  This test code will automatically be added in the "tests/testthat/" directory
-->

```{r tests-add_gt_to_word}
library(testthat)
library(officer)
library(gtsummary)
library(dplyr)
test_that("add_gt_to_word works correctly under different conditions", {
  
  # setup: create dummy data and a basic gtsummary object
  mock_df <- data.frame(
    status = c("active", "inactive", "active"),
    age = c(25, 34, 45)
  )
  mock_gt <- gtsummary::tbl_summary(mock_df, by = status)
  
  # test 1: verification of the standard portrait flow
  doc_portrait <- officer::read_docx()
  res_portrait <- add_gt_to_word(
    doc = doc_portrait, 
    gt_obj = mock_gt, 
    title = "Test Portrait Title", 
    landscape = FALSE
  )
  
  # check that an officer document object is returned
  expect_s3_class(res_portrait, "rdocx")
  
  # convert doc to content dataframe to check text insertion
  content_p <- officer::docx_summary(res_portrait)
  expect_true(any(content_p$text == "Test Portrait Title"))
  
  # test 2: verification of the landscape flow
  doc_landscape <- read_docx()
  res_landscape <- add_gt_to_word(
    doc = doc_landscape, 
    gt_obj = mock_gt, 
    title = "Test Landscape Title", 
    landscape = TRUE
  )
  
  expect_s3_class(res_landscape, "rdocx")
  
  content_l <- docx_summary(res_landscape)
  expect_true(any(content_l$text == "Test Landscape Title"))
  
  # test 3: safe verification of landscape structural layout differences
  # landscape structural blocks alter row structures or total summary elements
  expect_false(identical(nrow(content_p), nrow(content_l)))
})
```

## export list of gt objects/dataframes to word

```{r  function-export_list_to_word}
#' Export a list of dataframes to word
#'
#' @param table_list list of dataframes
#' @param doc_path character vector of output directory
#' @importFrom flextable  body_add_flextable
#' @importFrom officer body_add_par body_add_break 
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_list_to_word()}
export_list_to_word <- function(table_list, doc_path) {
  
  # initialize a new word document
  doc <- officer::read_docx()
  
  # iterate through the named list
  # .x is the table/df, .y is the name of the list element
  purrr::iwalk(table_list, function(.x, .y) {
    
    # logic to handle different table types
    if (inherits(.x, "gtsummary")) {
      # convert gtsummary to flextable
      ft <- gtsummary::as_flex_table(.x)
    } else {
      # treat as standard dataframe and format
      ft <- flextable::flextable(.x) %>%
        flextable::theme_booktabs() %>%
        flextable::autofit() %>%
        flextable::bold(part = "header")
    }
    
    # add the header and the table to the document
    doc <<- doc %>%
      # use the list name as a header 2
      body_add_par(.y, style = "heading 2") %>%
      body_add_flextable(ft) %>%
      # add a small spacer after each table
      body_add_par("") 
  })
  
  # save the completed document
  print(doc, target = doc_path)
}
```


## Create gtsummary table

```{r function-create_summary_table}
#' create a `gtsummary` table
#'
#' @param data dataframe
#' @param cols columns to include
#' @param labels character. labels
#' @param group_by character. grouping variable
#' @param strata character. stratifying variable
#' @param percent_type character. percents by row or column
#'
#' @returns a list of gtsummary object 
#' @export
#'
#' @examples
#'\dontrun{create_summary_table()}
create_summary_table <- function(data, 
                                 cols = dplyr::everything(), 
                                 labels = list(), 
                                 group_by = NULL,
                                 strata = NULL,
                                 percent_type = "column") {
  
  
  # internal helper to build the base table
  build_summary <- function(df) {
    df %>%
      # any_of(group_by) handles the case where group_by might be null
      dplyr::select({{cols}}, dplyr::any_of(group_by)) %>%
      gtsummary::tbl_summary(
        by = dplyr::any_of(group_by),
        label = labels,
        percent = percent_type,
        type = list(all_categorical() ~ "categorical"),
        statistic = list(
          all_continuous() ~ "{mean} ({median})",
          all_categorical() ~ "{n} ({p}%)"
        ),
        digits = list(
          all_continuous() ~ purrr::partial(style_number, big.mark = ","),
          all_categorical() ~ c(0, 1)
        ),
        missing = "no"
      ) %>%
      gtsummary::add_n() %>%
      gtsummary::bold_labels()
  }
  
  # logic to handle stratification
  if (!is.null(strata)) {
    data %>%
      # ensure the strata column exists before passing to tbl_strata
      gtsummary::tbl_strata(
        strata = all_of(strata),
        .tbl_fun = ~ build_summary(.x)
      )
  } else {
    build_summary(data)
  }
}
```

## create a listing table using `gtreg{}`

```{r function-create_listing_table}

#' create `gtreg` table listing of data
#'
#' @param data dataframe
#' @param cols columns to include
#' @param labels character. labels
#' @param group_by character. grouping variable 
#'
#' @returns a list
#' @export
#'
#' @examples
#'\dontrun{create_listing_table()}
create_listing_table <- function(data,
                                 cols = dplyr::everything(),
                                 labels = list(),
                                 group_by = NULL) {
  data %>%
    {
      if (!is.null(group_by))
        dplyr::arrange(., .data[[group_by]])
      else
        .
    } %>%
    dplyr::select({{ cols }}, dplyr::any_of(group_by)) %>%
    gtreg::tbl_listing() %>%
    gtsummary::bold_labels()
}

```

## clean a vector of strings

```{r function-clean_string}
# clean string

#' clean string of spaces and non alphanumeric characters and make lower case
#'
#' @param x a vector of characters
#'
#' @returns a vector
#' @export
#'
#' @examples
#' \dontrun{clean_String('Nachola124')}


clean_string <- function(x) {
  x %>%
    as.character() %>%
    stringr::str_to_lower() %>%
    stringr::str_replace_all("[^a-z0-9]", "") %>%
    stringr::str_trim()
}
```

## recode beneficiary camps to known refugee camps

```{r function-clean_camp}
# clean ben_camp

#' clean recode beneficiary camp 
#'
#' @param x dataframe
#' @param col_name  character. column with camp names
#'
#' @returns dataframe
#' @importFrom dplyr case_when all_of across
#' @export
#'
#' @examples
#' \dontrun{clean_camp(x, col_name = "ben_camp")}
clean_camp <- function(x, col_name = "ben_camp"){
  x  %>% 
    dplyr::mutate(
      across(all_of(col_name), ~ case_when(
        # use stringr or grepl with the .x pronoun
        grepl("\\bna\\b", .x, ignore.case = TRUE) & grepl("HIAS", reforgp, ignore.case = T) ~ "Nairobi",
        grepl("^k", .x, ignore.case = TRUE) ~ "Kakuma",
        grepl("^nai|east|urban|umoja", .x, ignore.case = TRUE) ~ "Nairobi",
        grepl("^daga|^ifo|f0|fo|^haga|^v|dad|daad", .x, ignore.case = TRUE) ~ "Daadab",
        TRUE ~ .x
      ))
    )
}
```


## find households in ESR data

```{r function-check_esr_registrations}
# registrations appearing in ESR
#' Title
#'
#' @param data1 dataframe. registration data
#' @param data2 dataframe. ESR data
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{check_esr_registrations(data1 = df_clean, data2 = df_esr)}
check_esr_registrations <- function(data1 = df_clean, data2 = df_esr) {

  
  # 1. prepare lookup keys from data2
  # combine names and clean id and phone for matching
  data2_prepared <- data2 %>%
    dplyr::mutate(
      full_name_clean = clean_string(paste(first_name, middle_name, last_name)),
      first_surname = clean_string(paste(first_name, last_name)),
      first_secname = clean_string(paste(first_name, middle_name)),
      id_clean        = clean_string(id_number1),
      phone_clean     = clean_string(phone_number)
    )
  
  lookup_ids    <- unique(data2_prepared$id_clean)
  lookup_names  <- unique(data2_prepared$full_name_clean)
  lookup_names1  <- unique(data2_prepared$first_surname)
  lookup_names2  <- unique(data2_prepared$first_secname)
  lookup_phones <- unique(data2_prepared$phone_clean)
  
  # 2. apply matching logic to data1
  data1 %>%
    dplyr::mutate(
      is_in_second_dataset = case_when(
        # check by id number
        clean_string(ben_id_number) %in% lookup_ids ~ "yes",
        
        # check by phone number using reg_num_phone
        clean_string(reg_num_phone) %in% lookup_phones ~ "yes",
        
        # check by combined name
        clean_string(ben_name) %in% lookup_names ~ "yes",
        clean_string(ben_first_surname) %in% lookup_names1 ~ "yes",
        clean_string(ben_first_secname) %in% lookup_names2 ~ "yes",
        
        # default to no if no matches are found
        TRUE ~ "no"
      )
    )
}
```

## get standard villages

```{r function-get_standard_villages}
#' separate villages collapsed by forward slash ("|")
#'
#' @param data1 dataframe. ESR/partner data
#' @param data2 dataframe. registration data
#'
#' @returns a dataframe
#' @importFrom stringr str_extract str_replace_all
#' @export
#'
#' @examples
#' \dontrun{get_standard_villages(data1 = df_partner, data2 = df_clean)}
get_standard_villages <- function(data1 = df_partner, data2 = df_clean) {
  
  # get unique reference villages and sort by length descending
  # this ensures long names are matched before substrings
  villages <- unique(data2$village2)
  villages_sorted <- villages[order(nchar(villages), decreasing = TRUE)]
  
  # escape special characters and collapse for regex
  village_pattern <- paste(stringr::str_escape(villages_sorted), collapse = "|")
  
  data1 %>%
    # separate rows based on the forward slash
    # this turns "village1/village2" into two distinct rows
    tidyr::separate_rows(village, sep = "/") %>%
    dplyr::mutate(
      # extract the standard name using the pattern
      village2 = str_extract(village, regex(village_pattern, ignore_case = TRUE)),
      # capitalize first letters of village names as per your requirement
      # this handles slashes if they appear in the result
      village2 = str_replace_all(village2, "(^|/)([a-z])", function(m) toupper(m))
    )
}
```

## pull kobo data

```{r function-pull_kobo_data}
#' pull data from kobo API
#'
#' @param url character. Kobotoolbox url
#' @param project_id character. project id
#' @param token character. project token
#'
#' @returns a list
#' @export
#'
#' @examples
#' \dontrun{pull_kobo_data()}
pull_kobo_data <- function(
    url = Sys.getenv("kobo_url"),
    project_id = Sys.getenv("unocha_id"),
    token = Sys.getenv("unocha_token")
    
){
  
  robotoolbox::kobo_setup(url = url, token = token)
  
  df <- robotoolbox::kobo_data(project_id)
  return(df)
}
```

## categorize numerics

```{r function-cut_numeri}
#' categorize numeric cols
#'
#' @param x a vector of numeric column
#' @param breaks numeric vector of breaks
#' @param labels character vector of labels
#'
#' @returns vector
#' @export
#'
#' @examples
#' cut_numeric(c(1:10), c(-Inf,5, Inf), c("<=5", ">=6"))
cut_numeric <- function(
    x, 
    breaks = c(16, 35, 60, Inf), 
    labels = c("16 - 35", "35 - 59", ">= 60")
){
  out <- cut(
    x,
    # 16 is the start, 35 and 60 are the breaks, inf covers everything above
    breaks = breaks,
    # labels for the resulting factor
    labels = labels,
    # include the lowest value and make intervals right-open [x, y)
    include.lowest = TRUE,
    right = FALSE
  )
  return(out)
}
```

## export gtsummary object to excel workbook

```{r function-export_to_excel_formatted}
#' Export/write gtsummary output to excel workbook
#'
#' @param list_of_objects list of `gtsummary` object
#' @param file_name character. output filename and associated path
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_to_excel_formatted()} 
export_to_excel_formatted <- function(
    list_of_objects,
    file_name = paste0("UNOCHA_Summary_Report_", Sys.Date(), ".xlsx")
) {
  
  # create the workbook object
  wb <- openxlsx::createWorkbook()
  
  # define styles
  neg_style <- createStyle(fontColour = "#FF0000", textDecoration = "bold")
  # style for bolding the row headers (the first column)
  bold_header_style <- createStyle(textDecoration = "bold")
  
  # iterate through the named list
  purrr::iwalk(list_of_objects, function(obj, sheet_name) {
    
    openxlsx::addWorksheet(wb, sheet_name)
    
    # conversion logic
    df_to_write <- if (inherits(obj, "gtsummary")) {
      as_tibble(obj)
    } else {
      as.data.frame(obj)
    }
    
    openxlsx::writeData(wb, sheet_name, df_to_write)
    
    # 1. bold the row headers (first column)
    # rows starts at 2 to skip the excel header row
    openxlsx::addStyle(
      wb, 
      sheet_name, 
      style = bold_header_style, 
      cols = 1, 
      rows = 2:(nrow(df_to_write) + 1), 
      gridExpand = TRUE
    )
    
    # 2. apply red color coding for negative values
    openxlsx::conditionalFormatting(
      wb, 
      sheet_name, 
      cols = 1:ncol(df_to_write), 
      rows = 2:(nrow(df_to_write) + 1), 
      rule = "< 0", 
      style = neg_style
    )
    
    openxlsx::setColWidths(wb, sheet_name, cols = 1:ncol(df_to_write), widths = "auto")
  })
  
  openxlsx::saveWorkbook(wb, file = file.path("output", file_name), overwrite = TRUE)
}
```

## add a new sheet to existing workbook

```{r function-export_to_excel_appended}
#' Export a sheet to existing workbook
#'
#' @param list_of_objects list of `gtsummary` object
#' @param sheetname character. worksheet name
#' @param file_name character. output file name
#' @param output_dir character. output directory
#'
#' @returns NULL
#' @export
#'
#' @examples
#' \dontrun{export_to_excel_appended()}
export_to_excel_appended <- function(
    list_of_objects = df_target_p,
    sheetname = "Target and Registered ESR",
    file_name = paste0("UNOCHA_Summary_Report_", Sys.Date(), ".xlsx"),
    output_dir = "output"
) {
  
  # create output dir
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  # build full path safely
  file_path <- file.path(output_dir, basename(file_name))
  
  message("Using file path: ", normalizePath(file_path, mustWork = FALSE))
  
  # load existing workbook OR create new
  if (file.exists(file_path)) {
    
    message("Loading existing workbook...")
    
    wb <- tryCatch(
      openxlsx::loadWorkbook(file_path),
      error = function(e) {
        message("Failed to load workbook: ", e$message)
        openxlsx::createWorkbook()
      }
    )
    
  } else {
    
    message("Workbook does not exist. Creating new workbook...")
    wb <- openxlsx::createWorkbook()
    
  }
  
  # excel sheet name limit
  sheetname <- substr(sheetname, 1, 31)
  
  # remove existing sheet if present
  if (sheetname %in% names(wb)) {
    openxlsx::removeWorksheet(wb, sheetname)
  }
  
  addWorksheet(wb, sheetname)
  
  # styles
  header_style <- createStyle(
    fontSize = 12,
    textDecoration = "bold",
    fontColour = "#0066A1"
  )
  
  bold_row_style <- openxlsx::createStyle(textDecoration = "bold")
  
  neg_style <- openxlsx::createStyle(
    fontColour = "#FF0000",
    textDecoration = "bold"
  )
  
  current_row <- 1
  
  purrr::iwalk(list_of_objects, function(obj, list_name) {
    
    df_to_write <- if (inherits(obj, "gtsummary")) {
      as_tibble(obj)
    } else {
      as.data.frame(obj)
    }
    
    # section title
    openxlsx::writeData(
      wb,
      sheet = sheetname,
      x = list_name,
      startRow = current_row
    )
    
    openxlsx::addStyle(
      wb,
      sheet = sheetname,
      style = header_style,
      cols = 1,
      rows = current_row
    )
    
    current_row <<- current_row + 1
    
    # write table
    openxlsx::writeData(
      wb,
      sheet = sheetname,
      x = df_to_write,
      startRow = current_row,
      headerStyle = bold_row_style
    )
    
    # bold first column
    if (nrow(df_to_write) > 0) {
      
      openxlsx::addStyle(
        wb,
        sheet = sheetname,
        style = bold_row_style,
        cols = 1,
        rows = (current_row + 1):(current_row + nrow(df_to_write)),
        gridExpand = TRUE,
        stack = TRUE
      )
      
      openxlsx::conditionalFormatting(
        wb,
        sheet = sheetname,
        cols = 1:ncol(df_to_write),
        rows = (current_row + 1):(current_row + nrow(df_to_write)),
        rule = "<0",
        style = neg_style
      )
    }
    
    current_row <<- current_row + nrow(df_to_write) + 3
  })
  
  # autofit
  openxlsx::setColWidths(
    wb,
    sheet = sheetname,
    cols = 1:50,
    widths = "auto"
  )
  
  # save
  openxlsx::saveWorkbook(
    wb,
    file = file_path,
    overwrite = TRUE
  )
  
  message("Workbook saved successfully.")
}
```

## get duplicates

```{r function-get_duplicates}
#' get duplicate beneficiaries on select variables
#'
#' @param x dataframe
#' @param vars_to_check character vector of variables to check
#' @param phone_var character vector. phone number variable for hard check
#' @param name_var character vector. name of respondent
#' @importFrom dplyr group_by across n n_distinct filter any_of ungroup distinct
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_duplicates()}
get_duplicates <- function(
    x,
    vars_to_check = c("benef_name", "ben_number"),
    phone_var = "ben_number",
    name_var = "benef_name"
){
  
  # 1. find phone sharing: same phone, different names
  phone_sharing <- x %>%
    group_by(across(any_of(phone_var))) %>%
    filter(n() > 1 & n_distinct(across(any_of(name_var))) > 1)
  
  # 2. find name sharing: same name, different phones
  name_sharing <- x %>%
    group_by(across(any_of(name_var))) %>%
    filter(n() > 1 & n_distinct(across(any_of(phone_var))) > 1)
  
  # 3. find exact duplicates across the vars_to_check
  exact_dups <- x %>%
    group_by(across(any_of(vars_to_check))) %>%
    filter(n() > 1)
  
  # combine all instances and remove overlapping rows
  res <- bind_rows(phone_sharing, name_sharing, exact_dups) %>%
    distinct() %>%
    ungroup()
  
  # sort by the primary variables for easy review
  res %>%
    arrange(across(any_of(vars_to_check)))
}
```

## get distinct

```{r function-get_distinct}
#' get distinct beneficiaries on select variables
#'
#' @param x dataframe
#' @param vars_to_check character vector of columns to check
#' @param phone_var character vector of phone number column
#' @param name_var character vector of respondent name
#' @importFrom dplyr group_by across n n_distinct filter any_of ungroup distinct select pull row_number bind_rows 
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{get_distinct()}
get_distinct <- function(
    x,
    vars_to_check = c("benef_name", "ben_number"),
    phone_var = "ben_number",
    name_var = "benef_name"
){
  # identify the ids or row indices of all duplicates first
  # we use your duplicate logic to find what to exclude
  duplicates <- x %>%
    # use row_number to keep track of exactly which rows are flagged
    mutate(temp_id = row_number()) %>%
    {
      # find rows that fail any of the duplicate checks
      bind_rows(
        # scenario 1: shared phone, different names
        group_by(., across(any_of(phone_var))) %>% 
          filter(n() > 1 & n_distinct(across(any_of(name_var))) > 1),
        
        # scenario 2: shared name, different phones
        group_by(., across(any_of(name_var))) %>% 
          filter(n() > 1 & n_distinct(across(any_of(phone_var))) > 1),
        
        # scenario 3: exact duplicates
        group_by(., across(any_of(vars_to_check))) %>% 
          filter(n() > 1)
      )
    } %>%
    pull(temp_id) %>%
    unique()
  
  # return the original data minus the duplicate rows
  x %>%
    mutate(temp_id = row_number()) %>%
    filter(!(temp_id %in% duplicates)) %>%
    select(-temp_id)
}
```

## plot bar

```{r function-plot_bar}
#' plot bar graphs
#'
#' @param data dataframe
#' @param x_var character vector x-axis variable
#' @param y_var character vector y-axis variable
#' @param fill_col character vector of hex color code for fill
#' @param title character vector of title
#' @param subtitle character vector of subtitle
#' @param x_lab character vector of label for x-axis
#' @param y_lab character vector of label for y-axis
#' @importFrom rlang ensym
#'
#' @returns list of plot object
#' @export
#'
#' @examples
#' \dontrun{plot_bar()}
plot_bar <- function(data, 
                     x_var, 
                     y_var = NULL, 
                     fill_col = "#0066A1",
                     title = "variable distribution", 
                     subtitle = NULL, 
                     x_lab = NULL, 
                     y_lab = "count") {
  
  x_sym <- rlang::ensym(x_var)
  
  if (!is.null(y_var)) {
    y_sym <- ensym(y_var)
    
    # case 1: plotting pre-calculated values with percent
    # we calculate total sum to derive percentages for the labels
    total_sum <- sum(data[[as_label(y_sym)]], na.rm = TRUE)
    
    # p <- ggplot(data, aes(x = fct_reorder(as.factor(!!x_sym), !!y_sym), y = !!y_sym)) +
    p <- ggplot2::ggplot(data, aes(x = !!x_sym, y = !!y_sym)) +
      ggplot2::geom_col(fill = fill_col, alpha = 0.9, width = 0.75) +
      ggplot2::geom_text(
        ggplot2::aes(label = paste0(!!y_sym, " (", scales::percent(!!y_sym / total_sum, accuracy = 0.1), ")")),
        vjust = -0.5, size = 3, fontface = "bold"
      )
    
  } else {
    # case 2: plotting frequencies with dynamic percent
    # p <- ggplot(data, aes(x = fct_infreq(as.factor(!!x_sym)))) +
    p <- ggplot2::ggplot(data, aes(x = !!x_sym)) +
      ggplot2::geom_bar(fill = fill_col, alpha = 0.9, width = 0.75) +
      ggplot2::geom_text(
        stat = "count",
        ggplot2::aes(label = after_stat(paste0(count, " (", scales::percent(count / sum(count), accuracy = 0.1), ")"))),
        vjust = -0.5, size = 3, fontface = "bold"
      )
  }
  
  p <- p +
    ggplot2::labs(
      title = title,
      subtitle = subtitle,
      x = if(is.null(x_lab)) as_label(x_sym) else x_lab,
      y = y_lab
    ) +
    # expand y axis further to accommodate longer labels
    ggplot2::scale_y_continuous(expand = expansion(mult = c(0, 0.2))) +
    ggplot2::theme_minimal(base_size = 12) +
    ggplot2::theme(
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      plot.title = element_text(face = "bold", size = 14, color = "#231F20"),
      plot.subtitle = element_text(size = 11, color = "#58595B"),
      axis.title = element_text(face = "bold", color = "#231F20"),
      axis.text.x = element_text(angle = 45, hjust = 1, color = "#231F20"),
      axis.text.y = element_text(color = "#231F20")
    )
  
  return(p)
}
```

## add custom total

```{r function-add_custom_total}
#' calculate column-wise totals
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{add_custom_total()}
add_custom_total <- function(df) {
  # calculate totals
  totals <- df %>%
    dplyr::summarise(
      Village = "Total",
      # only sum unique targeted values per village to avoid double counting
      `Targeted BNF` = base::sum(unique(`Targeted BNF`), na.rm = TRUE),
      `Registered BNF` = sum(`Registered BNF`, na.rm = TRUE),
      Variance = sum(Variance, na.rm = TRUE),
      `ESR Households` = sum(`ESR Households`, na.rm = TRUE),
      partner = NA
    )
  
  # combine with original data
  out <- bind_rows(df, totals)
  
  return(out)
}
```

## collapse villages

```{r function-collapse_villages}
#' collapse villages that share targets, separating by /
#'
#' @param df dataframe
#'
#' @returns dataframe
#' @export
#'
#' @examples
#' \dontrun{collapse_villages()}
collapse_villages <- function(df) {
  df_collapsed <- df %>%
    # group by the columns where you want to identify duplicates
    dplyr::group_by(`Targeted BNF`, `ESR Households`, partner) %>%
    dplyr::summarise(
      # collapse village names with a slash
      village = paste(unique(village), collapse = " / "),
      # sum the registered beneficiaries
      `Registered BNF` = sum(`Registered BNF`, na.rm = TRUE),
      # keep the group order clean
      .groups = "drop"
    ) %>%
    # relocate columns to match your original structure
    dplyr::select(village, `Targeted BNF`, `ESR Households`, partner, `Registered BNF`)
  
  return(df_collapsed)
}
```

## find partial matches

```{r function-get_partial_matches}
#' find partial matches between focal and beneficiary
#'
#' @param df dataframe
#' @param focal character vector of focal variable
#' @param beneficiary character vector of beneficiary variable
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{get_partial_matches()}
get_partial_matches <- function(df, focal = "focal_name", beneficiary = "benef_name") {
  df %>%
    dplyr::rowwise() %>%
    dplyr::mutate(
      is_partial = stringr::str_detect(.data[[focal]], fixed(.data[[beneficiary]])) | 
        stringr::str_detect(.data[[beneficiary]], fixed(.data[[focal]]))
    ) %>% 
    dplyr::ungroup()
}
```

## get personally identifying columns

```{r function-get_pii}
#' Get personally identifying columns
#'
#' @param df dataframe
#' @param vars string of identifying columns
#'
#' @returns a list of 3 dataframes: identifying, de-identified, and original df
#' @export
#'
#' @examples
#' \dontrun{get_pii()}
get_pii <- function(
    df = df_clean,
    vars = c("lat", "long", "name", "mobile", "gps", "phone", "enum_num$", "^(enum|reg|ben|proxy|sec)_.*(name|num|id)")
){
  cols <- paste(vars, collapse = "|")
  df_pii <- df %>% select(
    grep(cols, names(.))
  )
  df_depii <- df %>%
    dplyr::select(
      -grep(cols, names(.))
    )
  out <- list(df_pii, df_depii, df) %>%
    purrr::set_names(c("personaly-identifying", "de-identified", "original-data"))
  return(out)
}
```

## export a list to excel

```{r function-write_list_to_excel}
#' export a named list of data frames to an excel workbook
#' @param df_list a named list of data frames
#' @param file_path character string for the output path
#' @returns NULL
#' @export
#' @examples
#' \dontrun{write_list_to_excel()}
write_list_to_excel <- function(df_list, file_path) {
  
  # create a new workbook object
  wb <- openxlsx::createWorkbook()
  
  # define a professional header style
  header_style <- createStyle(
    fgFill = "#4F81BD", 
    halign = "center", 
    textDecoration = "bold",
    fontColour = "white",
    border = "bottom"
  )
  
  # iterate through the named list
  # .x is the dataframe, .y is the name of the list element
  purrr::iwalk(df_list, function(.x, .y) {
    
    # add a new worksheet with the name of the list element
    # excel sheet names are limited to 31 characters
    sheet_name <- substr(.y, 1, 31)
    openxlsx::addWorksheet(wb, sheet_name)
    
    # write the data to the sheet
    openxlsx::writeData(wb, sheet_name, .x, headerStyle = header_style)
    
    # adjust column widths to fit content automatically
    openxlsx::setColWidths(wb, sheet_name, cols = 1:ncol(.x), widths = "auto")
    
    # freeze the top row for easier navigation
    openxlsx::freezePane(wb, sheet_name, firstRow = TRUE)
  })
  
  # save the workbook to the specified path
  openxlsx::saveWorkbook(wb, file_path, overwrite = TRUE)
}
```

## pre-process unocha registration


```{r function-preprocess_reg}
#' preprocess unocha registrations
#'
#' @param df dataframe
#' @importFrom dm dm_flatten_to_tbl
#' @importFrom dplyr mutate group_by case_when if_else filter bind_rows distinct
#'
#' @returns a dataframe
#' @export
#'
#' @examples
#' \dontrun{preprocess_reg()}
#' 
preprocess_reg <- function(df){
  df_out <- df %>% 
  dm_flatten_to_tbl(main)

df_out <- df_out %>%
  mutate(across(where(labelled::is.labelled), as.character))
df_out <- df_out %>%
  dplyr::mutate(
    village = case_when(
      grepl("chini", village, ignore.case = T) & grepl(
        "kalakacha", sublocation, ignore.case = T
      ) ~ "Kalakacha Chini",
      TRUE ~ village
    )
  )
df_out <- df_out %>%
  mutate(
    village = case_when(
      grepl("Emmaus|Kirengender", village, ignore.case = T) ~ "Kirengender",
      grepl("salama|Kajificheni", village, ignore.case = T) ~ "Salama",
      grepl("kalkacha_juu", village, ignore.case = T) ~ "Kalakacha Juu",
      TRUE ~ village
    )
#     troublesome
    ,
    enum_egency = case_when(
      grepl("mitobini", village, ignore.case = T) &
      grepl("pgi", enum_egency, ignore.case = T) ~ "ALDEF",
      # grepl("Milimani", village, ignore.case = T) & 
      #   date(start) >= date("2026-05-13") ~ "ALDEF",
      # grepl("kurole", village, ignore.case = T) ~ "ALDEF",
      grepl("elwachu", village, ignore.case = T) ~ "ALDEF",
      TRUE ~ enum_egency
    )
  )

df_out <- df_out %>% 
  dplyr::mutate(
    ben_first_surname = paste(first_name, third_name),
    ben_first_secname = paste(first_name, second_name)
  )

df_out <- df_out %>% 
  dplyr::mutate(
    HHH_age_calc = if_else(
      !is.na(resp_age), resp_age, HHH_age
    )
  )

df_out <- df_out %>% 
  mutate(
    county = gsub("_", " ", county)
  )
df_out <- df_out %>%
  dplyr::mutate(
    age_category = cut_numeric(HHH_age_calc),
    hh_size_category = cut_numeric(
      hh_size,
      breaks = c(1, 3, 6, 10, Inf),
      labels = c("1 - 2", "3 - 5", "6 - 9", ">= 10")
    ),
    hh_disabl_category = cut_numeric(
      hh_disability_count,
      breaks = c(0,1, 3, 5, Inf),
      labels = c("0", "1 - 2", "3 - 4", ">= 5")
    )
  )

  
df_out <- df_out %>%
  labelled::set_variable_labels(
    age_category = "Household\n Head Age",
    HHH_chronic = "HH Head\n Chronic illness",
    HHM_chronic = "HH Member\n Chronic illness",
    average_income = "Monthly Income",
    hh_size = "Household Size",
    hh_size_category = "Household Size",
    FCSGName = "Food Consumption\n Score",
    eligibility = "Eligibility",
    enum_egency = "Agency",
    enum_name = "Enumerator\n first name",
    first_name = "respondent\n first name",
    reg_num = "Respondent Phone",
    trust_proxy = "Trust proxy",
    enum_egency = "Partner",
    proxy_num_consent = "proxy consent",
    name_proxy_first = "proxy\n first name",
    proxy_num = "proxy mobile",
    proxy_relation = "proxy relation",
    HHM_disability = "Household\n disability",
    HHH_chronic = "Household\nchronic illness"
  )

# registration started May 3 ----------------------------------------------

df_out <- df_out %>% 
  dplyr::filter(
#     start of reg
    start >= lubridate::date("2026-05-03")
  ) %>%
  dplyr::filter(!(start < lubridate::date("2026-05-07") & grepl("SND", enum_egency))) 
  
  # distinct(
  #   ben_name, reg_num, ben_id_number, .keep_all = T
  # ) %>%
  # distinct(
  #   reg_num, ben_id_number, .keep_all = T
  # ) %>% 
  
# reg_num can be NA but distinct() would only pick one row

df_clean <- df_out %>%
  filter(is.na(reg_num)) %>%
  bind_rows(
    df_out %>%
      filter(!is.na(reg_num)) %>%
      distinct(reg_num, .keep_all = TRUE)
  ) %>%
  mutate(
    village = gsub("_", " ", village),
    county = gsub("_", " ", county)
  )

}
```


## Proportional to size random sample

```{r function-draw_pps_sample}
#' Draw a Sample of Households Proportional to Village Size
#'
#' This function selects a stratified random sample of individual household rows 
#' from within counties. The probability of a household being selected is 
#' proportional to the total size (household count) of the village it belongs to.
#'
#' @param data A data frame containing raw household-level records.
#' @param county_var The bare or quoted column name identifying the county stratum.
#' @param village_id_var The bare or quoted column name identifying the unique village.
#' @param allocation_df A data frame containing two columns: the county identifier 
#'   and the allocated household sample size for that county.
#' @param alloc_size_var The bare or quoted column name in `allocation_df` 
#'   indicating the target sample size.
#' @param seed An integer value used to initialize the pseudo-random number generator 
#'   for reproducible sampling. Defaults to `42`.
#'
#' @return A data frame containing the individually selected household rows.
#'
#' @importFrom dplyr group_by mutate ungroup inner_join arrange row_number filter across any_of desc n
#' @importFrom rlang sym
#' @export
#'
#' @examples
#' \dontrun{
#' # dummy raw household data
#' raw_households <- data.frame(
#'   county_id = rep(c("County_A", "County_B"), each = 10),
#'   village_id = c(rep("V1", 3), rep("V2", 7), rep("V3", 4), rep("V4", 6)),
#'   household_id = paste0("HH_", 1:20)
#' )
#'
#' # household sample size targets per county
#' target_allocations <- data.frame(
#'   county_id = c("County_A", "County_B"),
#'   sample_size = c(2, 3)
#' )
#'
#' # run reproducible pps household sampling
#' pps_household_sample <- draw_pps_sample(
#'   data = raw_households,
#'   county_var = "county_id",
#'   village_id_var = "village_id",
#'   allocation_df = target_allocations,
#'   alloc_size_var = "sample_size",
#'   seed = 123
#' )
#' }
draw_pps_sample <- function(
    data, 
    county_var = "county", 
    village_id_var = "village", 
    allocation_df,
    alloc_size_var = "sample",
    seed = as.numeric(Sys.Date())
) {
  
  # preserve global environment random seed state if it exists
  if (exists(".Random.seed", envir = .GlobalEnv, inherits = FALSE)) {
    old_seed <- get(".Random.seed", envir = .GlobalEnv, inherits = FALSE)
    on.exit(assign(".Random.seed", old_seed, envir = .GlobalEnv))
  }
  
  # set the pseudo-random seed state for this execution context
  set.seed(seed)
  
  # calculate village size and append it back to every household row natively
  household_weights <- data %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(c(county_var, village_id_var)))) %>%
    dplyr::mutate(village_size = dplyr::n()) %>%
    dplyr::ungroup()
  
  # join the target household allocations per county
  joined_frame <- dplyr::inner_join(
    household_weights, 
    allocation_df, 
    by = rlang::set_names(county_var, county_var)
  )
  
  # generate a uniform random component combined with size weights for pps without replacement
  # then slice the top target records per county grouping
  sampled_households <- joined_frame %>%
    dplyr::group_by(dplyr::across(dplyr::any_of(county_var))) %>%
    dplyr::mutate(
      rand_weight = stats::runif(dplyr::n())^(1 / .data$village_size)
    ) %>%
    dplyr::arrange(dplyr::desc(.data$rand_weight), .by_groups = TRUE) %>%
    dplyr::filter(dplyr::row_number() <= unique(!!rlang::sym(alloc_size_var))) %>%
    dplyr::ungroup() %>%
    dplyr::select(-.data$rand_weight)
  
  return(sampled_households)
}
```

<!--
There can be other functions, examples and tests in your flat template.
Each of them will be inflated in a different file, provided that there is a level-1 or level-2 section title to separate from previous functions.
-->



<!--
# There can be development actions

Create a chunk with 'development' actions

- The chunk needs to be named `development` or `dev`
- It contains functions that are used for package development only
- Note that you may want to store most of these functions in the 0-dev_history.Rmd file

These are only included in the present flat template file, their content will not be part of the package anywhere else.
-->

```{r development-inflate, eval=FALSE}
# Keep eval=FALSE to avoid infinite loop in case you hit the knit button
# Execute in the console directly
fusen::inflate(flat_file = "dev/flat_full.Rmd", vignette_name = "Get started")
```


# Inflate your package

You're one inflate from paper to box.
Build your package from this very Rmd using `fusen::inflate()`

- Verify your `"DESCRIPTION"` file has been updated
- Verify your function is in `"R/"` directory
- Verify your test is in `"tests/testthat/"` directory
- Verify this Rmd appears in `"vignettes/"` directory
